---
title: "Lists (practice)"
---

```{r setup, include=FALSE,eval=TRUE}
pulse <- read_pulse() 
```

## Primary exercises

1. *Create and investigate a list*.  
Three students received different sets of grades (Amy: 1,6,7,9,10; Bob: 6,7,4,3,5,2,2,1,4; Dan: 9,9,10).  
In a variable `scores` create a `list` (the names of the list elements should be the names of the students and the values should be the corresponding grades).  
Print the list, its `class`, `length` and structure (`str`) of `scores`.
```{r}
scores <- list(
  Amy = c( 1,6,7,9,10 ),
  Bob = c( 6,7,4,3,5,2,2,1,4 ),
  Dan = c( 9,9,10 )
)
scores
class( scores )
length( scores )
str( scores )
```

2. *Add an element*, *change an element*.  
Reuse `scores` from the previous exercise.  
Add there grades for Eve (7,3,5,8,8,9) and print the list.  
Then, for Dan merge new grades (8,8,6,7) with the existing grades (hint: use the combine function `c` to combine existing Dan's grades with the new grades then put the result back to `scores`; do not type again `9,9,10`).
```{r}
scores[[ 'Eve' ]] <- c(7,3,5,8,8,9)
scores
scores[[ "Dan" ]] <- c( scores[[ "Dan" ]], c(8,8,6,7) )
scores
```

3. *Single and double bracket operators*.  
Reuse `scores` from the previous exercises.  
Investigate the difference between `scores[[ "Bob" ]]` and `scores[ "Bob" ]`.  
Look at what is printed and what is the class of each result.  
Then compare `scores[[ c( "Amy", "Bob" ) ]]` with `scores[ c( "Amy", "Bob" ) ]`.  
Understand, why the error is reported.
```{r error=TRUE}
scores[[ "Bob" ]]             # Returns the value of Bob element (vector)
scores[ "Bob" ]               # Creates a new list with only Bob there (list)
class( scores[[ "Bob" ]] )
class( scores[ "Bob" ] )
scores[[ c( "Amy", "Bob" ) ]] # A list is needed to return two elements
scores[ c( "Amy", "Bob" ) ]   # This creates a list, so many elements are ok
```

4. *Dollar operator*.  
Reuse `scores` from the previous exercises.  
Investigate the (lack of) difference between `scores$Bob` and `scores[[ "Bob" ]]`.  
Look at what is printed and what is the class of each result.  
Then compare `scores$Bo` with `scores[[ "Bo" ]]`.  
Understand, why the `NULL` is returned.

```{r}
scores$Bob        # another way to access Bob
scores[[ "Bob" ]] # get an element with exact name Bob
class( scores$Bob )
class( scores[[ "Bob" ]] )
scores$Bo         # strange matching of names, it still finds Bob
scores[[ "Bo" ]]  # there is no "Bo" so NULL is returned
```

## Extra exercises

1. *A list returned by a function; test for association/correlation*.  
For this exercise we need two random numerical vectors.  
Let's create `x` and `y`, each of 30 elements sampled from the normal distribution: `x <- rnorm( 30 )` and `y <- rnorm( 30 )`.  
Print these vectors. You may also produce a scatter plot: `plot( x, y )`.  
  \
The function `cor.test` tests for association between corresponding elements of two vectors.  
Use `h <- cor.test( x, y )` and print `h` to see a report of the association test.  
Internally `h` is stored as a list. Print names of the elements stored in `h`.  
Now, read `Help` for `cor.test`. In the section `Value` you will see the description of the `h` elements.  
Get directy the values of elements `estimate` and `p.value`.
```{r}
x <- rnorm( 30 )
y <- rnorm( 30 )
x
y
plot( x, y )
h <- cor.test( x, y )
h
names( h )
h[[ 'estimate' ]]
h[[ 'p.value' ]]
```

2. *A nested list*.  
Let's extend the concept of `scores` to describe various topics (see the code below).  
Check `class` and `str` of `scores`.  
Calculate how many students are in the `scores` list.  
Get Dan's scores in physics.

```{r echo=TRUE}
scores <- list(
  Amy = list(
    math = c( 1,6,7,9,10 ),
    biology = c( 7,6,8 )
  ),
  Bob = list(
    math = c( 6,7,4,3,5,2,2,1,4 ),
    physics = c( 8,7 )
  ),
  Dan = list(
    math = c( 9,9,10 ),
    physics = c( 10, 10, 10 ),
    biology = c( 3, 5, 7 )
  )
)
```

```{r}
class( scores )
str( scores )
length( scores )      # number of students
length( scores$Bob )  # number of topics for which Bob has scores
scores[[ "Dan" ]][[ "physics" ]]
scores$Dan$physics
scores$Dan[[ "physics" ]]
```

## Multitopic exercises

1. *(ADV) Split a table into list of tables by a column factor; merge back*.  
Some functions might require an input to be provided as a list of tables.  
Let's assume that the `pulse` table should be split into a list of table parts based on the `exercise` argument.  
Load the `pulse.csv` data to variable `pulse`.  
Try `l <- pulse %>% split( .$exercise )` and investigate the `class`, `length` and `names` of the result `l`.  
Use double square bracket to extract the part for `exercise` being `low`.  
Finally, check that with `bind_rows` applied to `l` you can recreate the `pulse` table (but with a different order of rows).
```{r}
l <- pulse %>% split( .$exercise )  # . represents the object on the left side of %>%
class( l )
length( l )
names( l )
l[[ "low" ]]
recreatedPulse <- bind_rows( l )
dim( pulse )
dim( recreatedPulse )
```

2. *(ADV) Split a table by a column and write each part to a different file*.  
Continue with the setup of the previous exercise.  
Study/type/exectute the following example.  
Find the newly created files in your filesystem.

```{r echo=TRUE,eval=FALSE}
l <- pulse %>% split( .$exercise )
exercises <- names( l )                   # name in l of each table chunk
for( exercise in exercises ) {            # exercise will be a name of a single chunk
  fileName <- paste0( "pulse_", exercise, ".csv" )  # name of the file for the chunk
  message( "Writing file '", fileName, "'..." )
  write_csv( l[[ exercise ]], file = fileName )
}
```

