---
title: "Mutate (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises


1.  In the survey dataset add a new column `feet` with heights reported in feet unit (1 foot = 30.48 cm).

```{r}
mutate(survey, feet=height/30.48)
```

2. In the survey dataset add a new column `diffWritingHandSpan` : the difference of span1 (writing hand) and span2 (non-writing hand). 

```{r}
mutate(survey, diffWritingHandSpan=span1-span2)
```

3. In the pulse dataset add new weight variables `pounds` and `stones` (1 kg  = 2.20462 pound = 0.157473 stone).

```{r}
mutate(pulse, pounds = weight * 2.20462, stones = weight * 0.157473 )

```


## Extra exercises

In the survey dataset: 

1. Add a new column `diffHandSpan` : the absolute difference between span1 (writing hand) and span2 (non-writing hand). Hint: use `abs` function (?abs).  

```{r}
mutate(survey, diffWritingHandSpan=abs(span1-span2))
```

2. Change `height` unit from cm to inch (1 cm = 0.393701 inch).

```{r}
mutate(survey, height=height*0.393701)
```

## Extra extra exercises

Here we are going to apply multiple functions from the morning slot in composition.

1. Produce the tibble containing the personal information of only those having height (in foot unit) 
between and including 6.0 and 6.5 feet.     

```{r}
filter( 
  mutate( 
    select(survey, Name=name, Age=age, Gender=gender, Height=height), feet=Height/30.48 
  ), feet >= 6.0 & feet <= 6.5 
) 

```


2. How many observations are in the following call, please try to reason before running the statement. 

```{r, eval=FALSE}
# 0 rows. Ignoring the mutate step, first filtering on height>190 results in a set without heights <=190.
filter(mutate(filter(survey, height> 190),feet=height/30.48),height<=190 )
```

