---
title: 'Filter'
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
```

> Use `filter` to select observations (rows) from a tibble. 

In the previous section we learned  how to select variables (columns) with all its values using the `select` function. In this section we are intersted in a collection of observations (rows). For this we have the function `filter` which takes as its first argument the tibble followed by a **logical condition** describing the values to be filtered on. 

## Logical conditions

A *logical condition* is a way to express whether some statement is true or false. For example, let's say we want to know *which individuals in our pulse dataset are taller than 190 cm?* You can achieve this with `filter` :

```{r}
filter(pulse, height>190) # filter all observations with height greater than 190 cm 
```

or another example would be, *who's weight is exactly 55 kg?*

```{r}
filter(pulse, weight==55) # filter all observations with weight equal to 55 kg 
```

`filter` uses logical vectors to achieve this results. You've already seen an example of a [logical vector](basic_vectors0.html#logical_vector) in the section on [Vectors](basic_vectors0.html). The result of a comparison with relational operators {`==`, `!=`, `<`, `<=`, `>`, `>=`} is a logical.

`r info_block("'==' stands for 'equal to' and '!=' is 'not equal to'")` 

In the examples above the comparisons `height>190` and `weight==55` are in fact logical vectors with the same length as variables `height` and `weight` in the pulse dataset. 

> The function `filter` takes the logical vector and returns only the rows in positions for which the value is TRUE.


## Multiple conditions : and ('&'), or ('|')

In the previous examples with `filter` we only had a single condition. We can expand the conditions to ask more complex questions. For example *which males weigh less than 70?*  Here we have two conditions: 

  - `gender=="male"` : male observations 
  - `weight<70`      : weigh less than 70  

> And : '&'

Both these conditions must be met, with other words for each observation `gender=="male"` **and**  `weight<70` must be true and this is expressed with the symbol **&** : 

```{r}
filter(pulse, gender=="male" & weight<70)
```


Another example: *select females with high frequency of exercise*:

```{r}
filter(pulse, gender=="female" & exercise=="high")
```


> Or : '|'

The *or* conditional operator is true when at least one of the conditions is true. For example, *list all rows with indiviudals who drink or smoke or both*: 

```{r}
filter(pulse, alcohol=="yes" | smokes=="yes")
```

There are 70 observations who **at least** do one of the two: drink or smoke.

> Which group is missing from the result above?

```{r, echo=FALSE, results='asis'}
msg <- "Those that neither smoke nor drink."
qa(msg)
``` 
 
Another example,  *filter rows from years 1995 and 1997*

```{r}
filter(pulse, year==1995 | year==1997)
```

Here we have the same variable `year` in our `|` construct.

> What is wrong with this command:  filter(pulse, year==1995 & year==1997) ?

```{r, echo=FALSE, results='asis'}
msg <- "An observation in a variable can not take two values at the same time. Here we have conditions on both sides of '&' on the same varible, in this case <b>year</b>, and since year can not be 1995 and 1997 at the same time for an observation the result of this query is empty."
qa(msg)
``` 
