---
title: "Join (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

1. For this exercise we will first split  the `survey` dataset into two separate tables 
in order to join them again! Call these `df1` and `df2`, these will have disjoint set of 
variables except `name` and `age`, the variables `name` and `age` combined are unique in 
all observations and will be used later for joining. Take for example all variables related 
to arm or hand in df1 and the rest in df2:

```
df1 : "name"   "span1" "span2" "hand"  "fold"   "clap"  "age"
df2 : "name"   "gender"  "pulse"  "exercise"   "smokes"  "height" "m.i" "age"
```

```{r}
df1 <- survey %>%  select(name,age,span1,span2,hand,fold,clap)
df1
df2 <- survey %>%  select(name,age,gender,pulse,exercise,smokes,height,m.i)
df2
```

a. Join df1 and df2 by `name` and `age` such that you obtain the original survey table. 

```{r}
# In the join the order of the variables are different than the original survey tibble 
# but the content is identical.
inner_join(df1,df2,by=c("name","age"))
```

b. In exercise (a) does it make any difference to choose either of `inner_join`, `left_join` or `full_join`? Hint: compare two tables with function `all_equal`.       

```{r}
# Answer is no, this because both df1 and df2 come from the same source 
# `survey` with equal number of matching observations with matching keys. 
#

all_equal(left_join(df1,df2,by=c("name","age")), survey)   # left_join
all_equal(inner_join(df1,df2,by=c("name","age")), survey)  # inner_join
all_equal(full_join(df1,df2,by=c("name","age")), survey)   # full_join
```

c. Are the pairs `name` and `age`  also good candidates as the key, i.e. is the combination of `name` and `age` uniquely identify each observation in the survey data? What about the combination of `name` with `span1` or `span2`? 

```{r}
# Yes, there are no duplicates in the pairs (name,age) in survey data.
sum(duplicated(survey %>% select(name,age)))==0
# The pair name and span1 will also suffice as a key.
sum(duplicated(survey %>% select(name,span1)))==0
# Multiple observations in the survey data have the same 
# name span2 combinations, therefore not a good candidate 
# as a key.
sum(duplicated(survey %>% select(name,span2)))==0
```


<!--  removed exercise 3 -->


```{r, echo=FALSE, eval=FALSE}
# 3. Create the following tibbles (copy/paste):

df1 <- tibble(name=c("Bobby","Gerald","Amelia","Lottie","Dennis"),
        height=c(179.1,NA,167.64,170,200), 
        age=c(18.917,16.917,17.333,17.25,18.5),
        span1=c(20,18,15.9,19,22))

df2 <- tibble(name=c("Bobby","Gerald","Amelia","Lottie","Dennis"),
        age=c(18.917,16.917,17.333,17.25,18.5),
        span2=c(19.5,13.3,16.5,18.5,21.5))

# - Join df1 and df2 by `name`. 
left_join(df1,df2, by="name")

# - Do the same but now with `{name,age}` as the key. Explain the difference.
# When joining only by 'name' we force the join function to resolve the name 
# conflict on 'age', it generated two columns 'age.x' and 'age.y' for df1$age 
# and df2$age respectively. But when using 'age' as part of the key (name,age) 
# then 'age' is treated as a common column and will occur only once.

left_join(df1,df2, by=c("name","age"))
# There is no name conflict here, both 'name' and 'age' are as keys for joining and all other variables have unique names. 
left_join(df1,df2, by=c("name", "age"))

```
  
## Extra exercises

1. The `name` and `span2` were not good combinations as a key, show the rows with identical 
 `name` and `span2` values.
 
```{r}
# step 1 : first find the {name,span2} of duplicated rows as a tibble
dups <- survey %>%  filter(duplicated(survey %>% select(name,span2)))  %>% select(name,span2) 
dups
# step 2 : join all observations from survey where there is a match with {name,span2} in dups
inner_join(survey, dups , c("name","span2")) 
```
 


2. Explain the result if you  use only the `name` variable as the key in the primary exercise (1) .

```{r}
# The variable 'name' is not unique, so for all those non-unique names the 
# joining of df1 and df2 will result in a product. For example the name 
# 'Benjamin' appears twice in the survey data, therefore it appears twice
# in each for the tibbles df1 and df2. Each row with the name 'Benjamin' 
# in df1 matches twice to the name 'Benjamin' in df2, so in total there 
# will be 4 rows in the final result.
# 
inner_join(df1,df2,by="name") %>% filter(name=="Benjamin")

# There are 78 more rows extra as the result of the ambiguities in joining
# with 'name' only. 
#
name_age_key_row_count <- inner_join(df1,df2,by=c("name","age")) %>% nrow() # 'name' and 'age' as the key
name_age_key_row_count
name_key_row_count <- inner_join(df1,df2,by="name") %>% nrow() # only 'name' as the key
name_key_row_count
name_key_row_count - name_age_key_row_count # number of additional rows due to non-unique key 'name'
```

```{r, eval=FALSE, echo=FALSE}
# Maybe another time 
#
# 2. Calculate the number of additional rows without doing the join using the column `name` only.   
# We can calculate the number of additional rows by using the frequency of 
# the non-unique names in survey 
freq_non_uniq_names <- survey %>% count(name) %>% filter(n>1) %>% pull(n)
freq_non_uniq_names
# The total additional rows should be the sum of the product of all the these frequencies minus the correct 
#  matches (if 'age' was used.):
sum(freq_non_uniq_names*freq_non_uniq_names - freq_non_uniq_names) 
```

