---
title: "Variables (practice)"
---

## Primary exercises

1. *Create a new variable*.  
Make a variable `heightInMeters` that contains your height expressed in meters.  
Print the variable.  
Find the variable and its value in the Environment pane.

```{r}
heightInMeters <- 1.75
heightInMeters
```

2. *Use a variable in an expression*.  
Use the variable `heightInMeters` to calculate your height in cm.  
Calculate the height in cm once again and store the result in a variable `heightInCentimeters`.  
Show/print `heightInCentimeters`.

```{r}
heightInMeters * 100
heightInCentimeters <- heightInMeters * 100
heightInCentimeters
```

3. *Change a value of a variable*.  
Increase the height stored in `heightInMeters` by 0.1m.  
Check whether the value of the variable indeed changed.  
Has also the value of `heightInCentimeters` changed? Why?

```{r error=TRUE}
heightInMeters <- heightInMeters + 0.1
heightInMeters
heightInCentimeters      # not changed, for change assignment is needed
```

4. *Typo in variable name*.  
Observe the error shown when you try to print the variable but you make a typo.

```{r error=TRUE}
height.in.meters      # Error: object 'height.in.meters' not found
```

## Extra exercises

1. *The pi*.  
`pi` is a variable built in R. Use it to calculate the area of a circle of radius 2.0.

```{r}
radius <- 2.0
pi * radius^2
```

2. *Various assignment operators*.  
In addition to `<-` R has other assignment operators `=` (equivalent to `<-`) and `->` (the variable name on the right side).  
Use each of these operators in an assignment. When do you think `->` might be useful?

```{r}
x <- 1
x
y = 2
y
3 -> z
z
1 + 1/(
  1 + 1/(
    1 + 1/(
      1 + 1/(
        1
      )
    )
  )
) -> u # right arrow is handy while typing and when the cursor is at the end of a long expression,
u      # but makes code less readable (do not use in production code)
```

3. *List defined variables; remove a variable*.  
Use `ls()` to find what variables are currently known. Compare the output with the content of the `Environment` pane.  
Define a new variable `xyz = 5`. Find how to use `rm` to remove the variable from the current environment.

```{r error=TRUE}
xyz <- 5
rm( xyz )
xyz
```

4. *Restart R session*.  
(Warning: you are going to loose all what you have typed).  
First define several variables (so you can see some in the `Envinroment` pane).  
Try `Restart R` option from the `Session` menu.  
What happened? What happened to the variables?

```{r}
# Restarting brings R to the initial state.
# Note, that in menu "Tools/Global options" it is possible to enable:
#   "Restore .RData into workspace at startup".
# When this option is enabled some variables might be present when R is restarted.
# This might lead to unreproducible results so we suggest to disable that option.
```

