---
title: "Vectors"
---

A **vector** is a container of (multiple) elements:

- All elements are of the same type (e.g. all are numbers).
- The elements are kept at numbered positions (there is the first element, second, ...).
- The elements might be given names (first may be named "Jay", second "Gloria", ...).

Some examples of a vector (and its **type**):

- Heights of students (a vector of numbers, `numerical`).
- Names of students (a vector of texts, `character`).
- Whether students have siblings (a vector of FALSE/TRUE values, `logical`).
- Eye color of students (a vector of values from a limited choice list, `factor`).

Vector is the primary data structure of the R language.

## Numerical vector

A vector of numbers you can create with the **combine** function `c`.  
Type the following to create a vector of numbers and store it in variable `x`:

```{r}
x <- c(3.5, 6, 7, 1+1)
x
```

Calculations or functions often work on vectors elementwise.  
This is helpful to do multiple calculations simultaneously:

```{r}
x - 18
x^2
sqrt(x)
```

Some functions summarize a vector to a single number:

```{r}
sum(x)
mean(x)
```

## Sequence of numbers

A simple regular sequence you can create with `:` (**colon operator**). Try:

```{r}
y <- 1:10
y
7:9
```

You can also use the function `seq`:

```{r}
x <- seq(5,15,3)
x
```

## Character vector

A **character** vector, so a vector of any texts, you may also create with the combine function `c`:

```{r}
x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
x
```

Both single quotes (`'word'`) and double quotes (`"word"`) are allowed (must be identical at the beginning and the end of a text).

## Logical vector

In R the following words denote **logical** values: `FALSE` and `TRUE`.

Typically, logical vectors are results of logical conditions. Try yourself:

```{r}
x <- c( 0, 1.5, 5 )
x > 2           # The result of greater-sign-operator is a logical vector
```

The function `sum` might be used on a logical vector to count how many elements are `TRUE`. Enter:
```{r}
sum( x > 2 )
```

We can also explicitely store a calculated logical vector to a variable:
```{r}
v <- ( x > 2 )
v
```

Of course, a vector of logical values can also be created manually with the combine function `c`:

```{r}
x <- c( TRUE, TRUE, FALSE, TRUE, FALSE )
x
```

## Missing values

Use the word `NA` (**not available**) to represent a single **missing value**.  
Missing values can appear in vectors of any type (see examples in the next section).

When a missing value appears in a calculation, the result is usually also a missing value:

```{r}
x <- c( 1, 2, NA, 4, 5 )
x + 3
```

The function `is.na` produces a logical vector which is `TRUE` at the positions at which the missing values are. Try:
```{r}
x
is.na( x )
```

When combined with the function `sum`, it allows to **count the number of missing values** in a vector. Try:
```{r}
sum( is.na( x ) )
```

Or, when **negated** (with `!` symbol) it allows to **count the number of not missing values** in a vector. Try:
```{r}
sum( !is.na( x ) )
```

## Length and elements of a vector

Use `length` to find out the **length** (number of elements) in a vector (numerical, character, logical, ...):

```{r}
x <- c( 11, 12, NA, 14 )
length(x)
y <- c( "Jay", "Gloria", "Claire", NA, 'Mitchell', 'Cameron' )
length(y)
z <- c( FALSE, TRUE, NA, FALSE )
length(z)
```

Use square brackets to **select** an element from a vector.  
The first element is at position 1.
Try the following to select an element from the second position:

```{r}
x[2]
y[2]
z[2]
```

Now try to combine the brackets with `length` to select the last element. Type:

```{r}
x[ length(x) ]
y[ length( y ) ]
z[length(z)]
```

Try the following to **select multiple elements** at once:

```{r}
x[ c(1,3) ]
w <- 2:4
x[w]
z[c(1,1,1,1,2)]
```

Note, when a vector is **printed**, the number reported on the left in square brackets shows the position of the next vector element:

```{r eval=FALSE}
x <- 101:155
x
```
```
 [1] 101 102 103 104 105 106 107 108 109 110 111 112
[13] 113 114 115 116 117 118 119 120 121 122 123 124
[25] 125 126 127 128 129 130 131 132 133 134 135 136
[37] 137 138 139 140 141 142 143 144 145 146 147 148
[49] 149 150 151 152 153 154 155
```

## Type (class) of a vector

Sometimes you may need to find what **type** of a vector is stored in a variable.  
This can be done with the function `class`.  

Type the following to create vectors `x`, `y` and `z`:

```{r}
x <- c( 11, 12, NA, 14 )
y <- c( "Jay", "Gloria", "Claire", NA, 'Mitchell', 'Cameron' )
z <- c( FALSE, TRUE, NA, FALSE )
```

To find the types of these vectors enter:

```{r}
class( x )
class(y)
class( z )
```
