---
title: "Formulas (practice)"
---

```{r , include=FALSE, eval=TRUE}
pulse <- read_pulse()
survey <- read_survey()
```

## Primary exercises

1. *A formula in a function; simple linear regression*.  
Load the `survey.csv` data to variable `survey`.  
You may make a simple scatter plot with `plot( survey$span2, survey$span1 )`.  
Perform a linear regression of `span1` being dependent on `span2`; store the result in `fit` variable.  
`fit` behaves also like a `list` so show `names` of elements stored in `fit`.  
Find and print the element corresponding to the coefficients of the regression -- do you see there a coefficient for `span2`?  
An intended method to access the coefficients is through `coef` function applied to `fit`. Try it. Do you get the same result as observed above?  

```{r message=FALSE,eval=FALSE}
survey <- read_csv( "survey.csv" )
```

```{r message=FALSE}
plot( survey$span2, survey$span1 )
fit <- lm( span1 ~ span2, data = survey )
fit
names( fit )
fit$coefficients
coef( fit )
```

## Extra exercises

1. *Simple regression diagnostic plots (demo)*.  
Apply `plot` function to the `fit` object calculated in the above exercise.  
These are diagnostic plots for the model produced by `lm`.

```{r}
plot( fit )
```

2. *Another test for association/correlation*.  
In one of the extra list exercises a test for correlation between two random vectors was performed.  
Follow the same steps to test correlation between `span1` and `span2` of `survey`.  
Is correlation between these vectors significant?

```{r}
h <- cor.test( survey$span2, survey$span1 ) # vector notation
h <- cor.test( formula = ~ span2 + span1, data = survey )  # formula notation
h
names( h )
h[[ 'estimate' ]]   # correlation coefficient
h[[ 'p.value' ]]    # significance, is < 0.05, strongly significant association
```

