---
title: "Vectors (practice)"
---

## Primary exercises

1. *Dietary intakes*.  
Four patients had daily dietary intakes of 2314, 2178, 1922, 2004 kcal.  
Make a vector `intakesKCal` of these four values.  
What is the class of this vector?  
Convert the values into in kJ using 1 kcal = 4.184 kJ.  

```{r}
intakesKCal <- c( 2314, 2178, 1922, 2004 )
intakesKCal
class( intakesKCal )
intakesKCal * 4.184
```

2. *Combining (appending) vectors*.  
Additional set of intakes is provided: 2122, 2616, NA, 1771 kcal.  
Make another vector `intakesKCal2` with the provided values.  
Use `c()` to combine (merge) `intakesKCal` and `intakesKCal2` into one vector `allIntakesKCal`.  
Print the combined vector and print its calculated `length`.  

```{r}
intakesKCal2 <- c( 2122, 2616, NA, 1771 )
allIntakesKCal <- c( intakesKCal, intakesKCal2 )
allIntakesKCal
length( allIntakesKCal )
```

3. *Mean and sum*.  
Calculate `mean` intake for patients in vector `allIntakesKCal`.  
Can you explain the result?  
Check help for `?mean`, in particular the `na.rm` argument.  
Calculate `mean` of `allIntakesKCal` ignoring not available elements.  
Now, calculate the total `sum` of intakes ignoring the `NA` element.
```{r}
mean( allIntakesKCal )
# since one element is missing, the mean is unknown
# ?mean, adding argument na.rm=TRUE will omit NA elements
mean( allIntakesKCal, na.rm = TRUE )
sum( allIntakesKCal, na.rm = TRUE )
```

4. *Number of (non)missing elements*.  
Understand the result of `is.na( allIntakesKCal )`.  
Now, negate the above result with `!` operator.  
Use above vectors as argument to `sum` to calculate the number of missing and non-missing elements in `allIntakesKCal`.  
Understand `allIntakesKCal[ !is.na( allIntakesKCal ) ]`.
```{r}
is.na( allIntakesKCal )         # TRUE marks positions with missing data
!is.na( allIntakesKCal )        # TRUE marks positions with available data
sum( is.na( allIntakesKCal ) )                # number of missing elements
sum( !is.na( allIntakesKCal ) )               # number of non-missing elements
allIntakesKCal[ !is.na( allIntakesKCal ) ]    # keeps elements which are not NA
```

5. *Descriptive statistics of a set of random numbers*.  
The code `v <- rnorm( 10 )` would store 10 random numbers in the variable `v`.  
Print `v`. 
Then repeat `v <- rnorm( 10 )` and print `v` again. Has `v` changed?
Print `v` and find by eye the smallest and the largest of these numbers.  
Try to use the functions `min` and `max` on `v` -- have you found the same numbers?  
Calculate the `mean`, `median` and the standard deviation (`sd`) of `v`.  
Type `v < 0` and understand the result.  
How to interpret the number produced by `sum( v < 0 )`?
```{r}
v <- rnorm( 10 ) # a vector of random numbers
v
v <- rnorm( 10 ) # another vector of random numbers
v
min( v )
max( v )
mean( v )
median( v )
sd( v )
v < 0            # TRUE corresponds to elements of vector v smaller than 0
sum( v < 0 )     # calculates the number of negative elements in vector v
```

6. *Head and tail*.  
Often there is a need to quickly look at the beginning (`head`) or at the end (`tail`) of a vector.  
Try these functions to show the first 5 and the last 7 elements of a randomly generated vector `v <- rnorm( 20 )`.
```{r}
v <- rnorm( 20 )
v
head( v, 5 )
tail( v, 7 )
```

## Extra exercises

1. *More complex sequences of numbers*.  
Read help (see `Help` pane) about `seq` function.  
Use it to generate sequence: 10, 7, 4, 1, -2, -5.  
Understand the error message of `seq( 10, -5, 3 )`.

```{r}
seq( 10, -5, -3 )
seq( from = 10, to = -5, by = -3 )
```

2. *Repetitions*.  
Read help (see `Help` pane) about `rep` function.  
Use it to generate sequence: 0, 1, 0, 1, 0, 1.  

```{r}
rep( c( 0, 1 ), 3 )
```

3. *Type conversion to a character vector*.  
Sometimes it is necessary to convert a numerical vector to a character vector.  
Understand what the function `as.character` does for argument `1:5`.

```{r}
1:5
as.character( 1:5 )
class( 1:5 )
class( as.character( 1:5 ) )
```

4. *Type conversion to a numerical vector*.  
Sometimes it is necessary to convert a character vector to a numerical vector.  
Understand what the function `as.numeric` does for argument `c( "1", "-1", "x" )`.  
Note the warning message. Why is there `NA`?

```{r}
as.numeric( c( "1", "-1", "x" ) )
```

