---
title: 'Mutate'
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

> Use `mutate` to add new variables or modify the existing ones. 

For example, the pulse dataset has two pulse measurements, let's say we are interested in *average pulse* and we want this information to be available as a separate variable, e.g. `averagePulse`, in the pulse tibble. Then we can do this with:    

```{r}
mutate(pulse, averagePulse = (pulse1+pulse2)/2)
```

By default the new column is added at the last position in the tibble.  

> Does the pulse tibble now contain the variable `averagePulse` ?

```{r, echo=FALSE, results='asis'}
msg <- "No, if you want to keep the new variable `averagePulse` you'll need to use assignment '<-' to replace the original pulse tibble with the newly modified version: <br><br>
pulse <- mutate(pulse, averagePulse = (pulse1+pulse2)/2)
<br>
"
qa(msg)
``` 


Take as another example the variable BMI: $$BMI=\frac{weight_{kg}}{{height_m}^2}$$

Note that BMI definition states that `weight` and `height` must be in *kilograms* and *metres* respectively. In the pulse dataset `weight` is given in kilograms but `height` is in centimetres. We can now first create a new variable `height_metre` containing the height in metres and then calculate BMI: 


```{r}
pulse_bmi <- mutate(pulse, height_metre=height/100) # convert centimetres to metre
pulse_bmi
```

`pulse_bmi` tibble has now the height in metre units, now we can calculate BMI:

```{r}
pulse_bmi <- mutate(pulse_bmi, BMI=weight/(height_metre^2)) 
pulse_bmi
```

Alternatively, you may skip the creation of `height_metre` and calculate BMI directly from the pulse tibble:

```{r}
pulse_bmi <- mutate(pulse, BMI=weight/((height/100)^2)) 
pulse_bmi
```


In the examples above we added a new variable to our dataset, but you can also update an existing variable. For example, let's say we want to have the age expressed (roughly) in days instead of years: 

```{r}
mutate(pulse, age=age*365)
```

here we keep the variable `age` but change its unit from year to days.
