---
title: "Tibble (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

### Create tibble  
  
1. Create a tibble `exercise_group` for a group of individuals with names {Sonja, Steven, Ines, Robert, Tim} with their heights {164, 188, 164, 180, 170}, weights {56.0, 87.0, 54.0, 80.0, 58.5} and frequency of exercise {high, high, low, moderate, low}.

```{r} 
exercise_group <- tibble(name=c("Sonja" , "Steven", "Ines", "Robert", "Tim" ),
            height=c(164, 188, 164, 180, 170),  
            weight=c(56.0, 87.0, 54.0, 80.0, 58.5),
            exercise=c("high", "high", "low", "moderate", "low")
     )  
exercise_group
```

### tibble subset

1. Take the tibble `exercise_group` from the previous exercise and create a new tibble without the `height` and `weight` variables.

```{r}
exercise_group_sub <- exercise_group[c("name","exercise")]
exercise_group_sub
```

### Extract variables as vectors

1. Given the tibble `favourite_colour`, how old were the subjects by the end of 2019?

```{r, echo=FALSE}
favourite_colour  <- tibble(name=c("Lucas","Lotte","Noa","Wim"), 
                           year=c(1995,1995,1995,1994), 
                           colour=c("Blue","Green","Yellow","Purple"))
```

```{r}
2019 - favourite_colour[["year"]] # alternatively:  2019 - favourite_color$year
```

2. What is the mean height in `exercise_group`? Use *mean* function (see ?mean).

```{r}
mean(exercise_group[["height"]]) 
```

### Read tibbles from file

1. Read `pulse.csv` data set into R and inspect its dimensions.

```{r, message=FALSE,eval=FALSE}
pulse <- read_csv(file = "pulse.csv")
```

```{r}
# two alternatives i) nrow and ncol function, ii) dim function.
nrow(pulse) # number of rows
ncol(pulse) # number of columns
dim(pulse)  # dimensions (rows, columns)
```

2. Read `survey.csv` data set into R. 

```{r, message=FALSE,eval=FALSE}
survey <- read_csv(file = "survey.csv")
```

  - Inspect the dimensions.

```{r}
dim(survey)
```

  - Show the first 9 and the last 7 rows.

```{r}
head(survey,9)
tail(survey,7)
```

  - Calculate the mean `age`.
    
```{r}
mean(survey$age, na.rm = TRUE) 
```

  - (*) Calculate the mean height in survey data. See also ?mean for more help on the `mean` function.

```{r}
# Here we use a second argument 'na.rm = TRUE' because there are missing values (NA) in 
# the variable height. By default the mean function returns NA if it first argument, in this 
# case variable 'height', contains any NA. The second argument 'na.rm = TRUE' changes this 
# behaviour by disregarding the observations with missing height and calculates the mean 
# of observations for which the height is available. 
# 
mean(survey$height, na.rm = TRUE) 
```

## Extra exercises

In survey data:

1. What is the mean height of the last 30 observations?

```{r}
survey_last_30 <- tail(survey,30)
mean(survey_last_30$height, na.rm = TRUE)
```

2. What is the mean height of the joint first 15 and last 30 observations? (see `?bind_rows`)

```{r}
# The function 'bind_rows' is used here to concatenate the two tibbles from the results of 
# 'head' and 'tail' functions. The resulting tibble 'survey_first15_last30' has 45 observations.
#
survey_first15_last30 <- bind_rows(head(survey,15), tail(survey,30)) # combined observations
mean(survey_first15_last30$height, na.rm = TRUE)  # mean of the combined observations
```
