---
title: 'Summarise'
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

> Descriptive statistics

## `summarise` variable(s)

The main function for calculating summaries on variables is `summarise`. Examples of descriptive functions are mean, median, sum etc. The functions consume a vector and produce a single value. `summarise` takes a tibble along with specification of descriptives and produces a single row. 

For example, let's say we want to know the mean height and weight of all individuals in the pulse dataset:  

```{r}
pulse %>% summarise(meanHeight=mean(height), meanWeight=mean(weight))
```

The result is a single row with two variables `meanHeight` and `meanWeight` with the corresponding mean values of all observations.

We can also summarise on variable's range, e.g. `age` : 

```{r}
pulse %>% summarise(minAge = min(age), maxAge=max(age)) # <=> range(pulse$age)
```

**n()**: convenient function to calculate total number of rows in the `summarise` context:

```{r}
pulse %>% summarise( count = n(), meanHeight = mean( height ) )
```

## `count` : frequency tables  

With the `count` function we can count the frequency of values in a categorical variables:

```{r}
pulse %>% count(gender)   # frequency of male/female
pulse %>% count(smokes)   # frequency of smoking habit 
pulse %>% count(exercise) # frequency of exercise habit 
```

The result enumerates the distinct values of the variable in the first column and their frequency in a new column `n`.

Multiple variables are allowed, it is the count of each possible combination of values, also known as `contingency table` or `cross table`: 

```{r}
pulse %>% count(gender, exercise)
pulse %>% count(year, gender)
```

## `distinct` values in variables

To identify distinct values in a variable or a group of variables we use the function `distinct`:

```{r}
pulse %>% distinct(year)
pulse %>% distinct(exercise)
pulse %>% distinct(ran) 
```

Again, multiple variables are allowd. To identify distinct combinations of `gender` and `exercise`:

```{r}
pulse %>% distinct(gender, exercise)
```

`r info_block("'distinct' produces the same variables combinations as the 'count' function except without the frequncy column 'n'.")`

You may use distinct also to check whether certain variables have unique values for each observation. Let's for example check whether all individuals in the `pulse` dataset have different names, more precisely, each observation is uniquely identifiable by the variable `name`:

```{r}
pulse %>% nrow()                    # total number of rows 
pulse %>% distinct(name) %>% nrow() # count the number of distinct names
```

There are `r pulse %>% distinct(name) %>% nrow()` distinct names and there in total `r nrow(pulse)` observations in the `pulse` dataset. This could only mean that there are multiple individuals in the pulse dataset with shared names: 

```{r}
nrow(pulse) == nrow( pulse %>% distinct(name)) # is 'name' unique for all observations?
```


## `glimpse`

With `glimpse` function you can quickly inspect the top part of all variables in one page with some additional meta information such as number of rows and columns and variable types: 

```{r}
pulse %>% glimpse()
```

## `arrange`

You may sort rows according to one or more variables with the function `arrange`.  

Try sorting the pulse dataset by `name`:

```{r}
pulse %>%  arrange(name) # sorts the rows by name in dictionary order 
```

or by `height`

```{r}
pulse %>%  arrange(height) # numerical order
```

By default the data is sorted in ascending order, to sort in descending order use `desc` function:

```{r}
pulse %>%  arrange(desc(name))
```

You may also arrange by multiple variables:

```{r}
pulse %>%  arrange(height,weight)
```

Here the data is first ordered by `height` and then by `weight`. 
