---
title: "Using R for data analysis"
subtitle: "Boerhaave Nascholing"
date : "June 11th, 2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA, eval = FALSE)
# knitr::opts_chunk$set(echo = TRUE, comment=NA, eval = TRUE)
library(tidyverse)
```

# Important note

The primary goal of this self-study assignment is to write an R Markdown document containing **the code** which calculates the answers to the questions below. Using the course and online materials is permitted. 

**The steps you need to take:**  

1. Create a new R Markdown file.
2. Develop the code with your answers in the R Markdown file. Put each question in a separate section, with the question number in the section title, just the question number would suffice.
3. Use `Knit` button regularly to check that your code does not produce errors. A knittable R Markdown file is the final result.

**R Markdown header:**  
Replace the header of your R Markdown file with what is shown below and put your name as `author`:

    ---
    title: "Using R for data analysis"
    subtitle: "Boerhaave Nascholing"
    date : "June 11th, 2021"
    author: "name"
    output: html_document
    ---

# Diamonds dataset

For this self-study you will be using the `diamonds` dataset. It contains various attributes of sold diamonds (see also `?diamonds`). The dataset comes with the `tidyverse` package. Once you load the `tidyverse` library you'll have access to the dataset in the `diamonds` variable. Make sure you put `library(tidyverse)` in the R chunk at the top of your R Markdown file.

```{r message=FALSE, eval=TRUE, echo=TRUE}
library( tidyverse )
diamonds
```

## Diamonds dataset tibble

Each row of the `diamonds` tibble describes one sold diamond. There are the following variables (columns):

- `price`: Price (in US dollars).
- `carat`: Weight (in carat units: 1 carat = 0.2g).
- `cut`: Quality of the cut (`Fair`, `Good`, `Very Good`, `Premium`, `Ideal`).
- `color`: Color, from `J` (worst) to `D` (best).
- `clarity`: How clear the diamond is (`I1` (worst), `SI2`, `SI1`, `VS2`, `VS1`, `VVS2`, `VVS1`, `IF` (best)).
- `x`, `y`, `z`: Length, width, depth (each in mm).
- `depth`: Total depth percentage = z / mean(x, y) = 2 * z / (x + y).
- `table`: Width of top of diamond relative to widest point.

# Questions

Q1. [0.5p] Show the type/class of the `diamonds` table. [0.5p] Show the type of the column `clarity`.

```{r}
class(diamonds)
class(diamonds$clarity)
```

Q2. [1p] Show the structure of the `diamonds` table.

```{r}
str(diamonds) # alternatively: glimpse(diamonds)
```

Q3. [1p] Print the rows 12-15 (hint: combine `head` and `tail`).

```{r}
diamonds %>% head( 15 ) %>% tail( 4 )
```

Q4. [1p] Calculate the mean of the `price` column.

```{r}
mean(diamonds$price)
```

Q5. [1p] Give the **number** of levels of the factor in the `clarity` column.

```{r}
nlevels(diamonds$clarity) # alternatively: length(levels(diamonds$clarity))
```

Q6. [3p] Make a `list` with two elements calculated as follows from the `diamonds` table. Name the first list element `medianDepth` and set it to the median diamond depth. Name the second list element `clarities` and set it to the **vector of** levels of the column `clarity`.

```{r}
list(
  medianDepth = median( diamonds$depth ),
  clarities = levels( diamonds$clarity )
)
```

Q7. Frequencies and cross table.

a) [1p] Count all the combinations of the value pairs in columns `cut` and `clarity`.

```{r}
diamonds %>% count(cut,clarity)
```

b) [2p] Print a **cross table** of `cut` and `clarity`, with `cut` categories given in columns.

```{r}
diamonds %>%  
  count(cut,clarity) %>%  
  pivot_wider(names_from = cut, values_from = n)
```

Q8. [3p] Group the `diamonds` table by `color`. In each group calculate min, max, median and mean `price`.

```{r}
diamonds %>%
  group_by(color) %>%
  summarise(
    minPrice=min(price),
    maxPrice=max(price),
    meanPrice=mean(price),
    medianPrice=median(price),
    .groups='drop' 
  )
```

Q9. Diamond volume in a scatter plot.

a) [1p] Add a new column `volume` representing diamond's volume in cubic millimetres given the dimensions `x`, `y` and `z`. Store the tibble with the added column in the variable `diamonds_volume`.

```{r}
diamonds_volume <- diamonds %>% mutate(volume=x*y*z)
```

b) [2p] Use `diamonds_volume` tibble to plot the `volume` (vertical axis) against the `price` (horizontal axis) in a scatterplot. Colour points by `clarity`. Make points 0.5 transparent.

```{r, out.width='75%', fig.align='center'}
ggplot(diamonds_volume) +
  aes(x = price, y = volume, color = clarity) +
  geom_point( alpha = 0.5 )
```

c) [1p] Replot the scatterplot in Q9.b but now with rows where $volume>0$ and $volume\leq600$.  

```{r, out.width='75%', fig.align='center'}
ggplot(diamonds_volume %>% filter(volume > 0, volume <= 600) ) +
  aes(x = price, y = volume, color = clarity) +
  geom_point( alpha = 0.5 )
```


Q10. Read/write CSV files.

a) [1p] Write the table `diamonds_volume` to a *comma-separated values* (CSV) file. Give the following name to the file: `diamonds_volume.csv`

```{r}
write_csv(diamonds_volume, file = "diamonds_volume.csv")
```

b) [1p] Read the file `diamonds_volume.csv` back into variable `d` and show it.

```{r, message=FALSE}
d <- read_csv(file = "diamonds_volume.csv")
d
```
