---
title: 'User-defined functions'
params:
  basename: "advanced_user_functions0"
---

```{r setup, include=FALSE,eval=TRUE}
pulse <- read_pulse()
```

Often we want to repeat certain series of steps in the analysis but each time with a subtle changes to the variables involved. This leads to repetition of the code in the script and consequently difficult to maintain or share with others. To that end, R provides the `function` construct to develop your own functions with your own defined arguments. The following is a scenario to motivate the usage of user-defined functions.


> Motivation 

Here is a scenario based on the pulse dataset and we will develop our analysis step by step. The initial problem definition is to 
*calculate the mean height for individuals ranging over 50 to 60 kg*.

```{r}
pulse %>% filter(weight >= 50 & weight < 60) %>% 
    summarise(meanHeight=mean(height))
```

Now for the same individuals, *how many of them did the pulse increase?*

```{r, collapse=TRUE}
pulse %>% filter(weight >= 50 & weight < 60) %>% 
    summarise(meanHeight=mean(height))
pulse %>% filter(weight >= 50 & weight < 60) %>% 
    summarise(increasedPulse=sum(pulse1<pulse2))
```

Let's remove repetitions: 

```{r}
pulse %>% filter(weight >= 50 & weight < 60) %>% 
    summarise(meanHeight=mean(height), 
              increasedPulse=sum(pulse1<pulse2))

```

We may want to produce other weight ranges, e.g. 60-70 and 70-80:

```{r, collapse=TRUE}
pulse %>% filter(weight >= 50 & weight < 60) %>% 
  summarise(meanHeight=mean(height), 
            increasedPulse=sum(pulse1<pulse2))

pulse %>% filter(weight >= 60 & weight < 70) %>% 
  summarise(meanHeight=mean(height), 
            increasedPulse=sum(pulse1<pulse2))

pulse %>% filter(weight >= 70 & weight < 80) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2))
```

Now include the number of individuals in different ranges:

```{r, collapse=TRUE}
pulse %>% filter(weight >= 50 & weight < 60) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2), individuals=n())

pulse %>% filter(weight >= 60 & weight < 70) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2), individuals=n())

pulse %>% filter(weight >= 70 & weight < 80) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2), individuals=n())
```

> weightRangeSummary function

Define the function `weightRangeSummary` which takes pulse dataset along with the weight interval and produces the summary: 

```{r, eval=TRUE, echo=FALSE}
weightRangeSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(weight >= lowerBound & weight < upperBound) %>% 
    summarise(meanHeight=mean(height),
              increasedPulse=sum(pulse1<pulse2), individuals=n())
}

```

```{r, collapse=TRUE}
weightRangeSummary(pulse,50,60)
weightRangeSummary(pulse,60,70)
weightRangeSummary(pulse,70,80)
```


## User-defined functions in R

**Functions** are constructs that encapsulate series of **statements** so you do not repeat the same statements all over again when needed. 

```{r, eval=FALSE}
functionName <- function(...) {
  statement
  ...
  <value>  # the result of the function, alternatively: return(<value>)
} 
```

### Example 1

We **define** the following function `add_one`. It takes the value of an **argument** `x` and adds number `1` to it. The sum is the **result** of the function:

```{r}
add_one <- function(x) {
	x + 1
}
```

This function can now be **called** with different values of the argument `x`:

```{r}
add_one(2)
add_one(-1)
```

Type the function name without any `()` to see the **body** (definition) of the function?

```{r}
add_one
```

What is the class of `add_one`? 

```{r}
class( add_one )
```

### Example 2

Now let's define a function to calculate $x^2 + 1$:

```{r}
square_add_one <- function(x) {
	result <- x^2    # choose a variable for temporary result
	result + 1
}

square_add_one(2) # 2^2 + 1 = 5
square_add_one(-1) # (-1)^2 + 1 = 2

```

The following versions of this functions are all equivalent:

```{r, eval=FALSE}
# (v1) One operation per line.
square_add_one <- function(x) {
	result <- x^2
	result <- result + 1
	result    # The last statement is returned as the value.
}
# (v2) No additional 'result' variable needed 
square_add_one <- function(x) {
	x^2 + 1
}
```

> weightRangeSummary function definition 

```{r}
weightRangeSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(weight >= lowerBound & weight < upperBound) %>% 
    summarise(meanHeight=mean(height),
              increasedPulse=sum(pulse1<pulse2), individuals=n())
}
```


Account for the missing value in `increasedPulse` : 

```{r}
weightRangeSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(weight >= lowerBound & weight < upperBound) %>% 
    summarise(meanHeight=mean(height),
              increasedPulse=sum(pulse1<pulse2, na.rm = TRUE), individuals=n())
}
```


## Some observations

  - User-defined functions lead to a more structured and easier to read code.
  - Variables declared inside a function cease to exist once the function terminates.
  - Arguments are given in the same order as they are declared in the function unless the argument names are provided explicitly.
