---
title: 'User-defined functions'
params:
  basename: "advanced_user_functions0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)  # UNCOMMENT IF SLOT HAS TASKS.
pulse <- read_pulse()

```

```{r, child="_navigate.Rmd"} 
``` 


Often we want to repeat certain series of steps in the analysis but each time with a subtle changes to the variables involved. This leads to repetition of the code in the script. This leads to codes that are difficult to maintain or share with others. To that end R provides the `function` construct to develop your own functions with your own defined arguments. The following is a scenario to motivate the usage of user-defined functions.


> Motivation 

Here is a scenario of a data analysis based on the pulse dataset and we will develop our analysis step by step.

The initial problems definition is  to *calculate the mean height for individuals ranging over 50 to 60 kg*.


```{r}
pulse %>% filter(weight >= 50 & weight <= 60) %>% 
    summarise(meanHeight=mean(height))
```

now for the same individuals *how many of them did the pulse increase?*

```{r}
pulse %>% filter(weight >= 50 & weight <= 60) %>% 
    summarise(increasedPulse=sum(pulse1<pulse2))
```

refactor (removing repeatitions) and make sure `increasedPulse` does not produce NA's.

```{r}
pulse %>% filter(weight >= 50 & weight < 60) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2))
```

Next, repeat this for other weight ranges 60-70 and 70-80.


```{r}
pulse %>% filter(weight >= 50 & weight < 60) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2))

pulse %>% filter(weight >= 60 & weight < 70) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2))

pulse %>% filter(weight >= 70 & weight < 80) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2))
```


include number of individuals in all groups.

```{r}

pulse %>% filter(weight >= 50 & weight < 60) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2), individuals=n())

pulse %>% filter(weight >= 60 & weight < 70) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2), individuals=n())

pulse %>% filter(weight >= 70 & weight < 80) %>% 
  summarise(meanHeight=mean(height),
            increasedPulse=sum(pulse1<pulse2), individuals=n())

```


> 

Define the function `weightRangeSummary` which takes pulse dataset along with the weight interval and produces the summary: 

```{r, echo=FALSE}
weightRangeSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(weight >= lowerBound & weight < upperBound) %>% 
    summarise(meanHeight=mean(height),
              increasedPulse=sum(pulse1<pulse2), individuals=n())
}

```

```{r,collapse=TRUE}
weightRangeSummary(pulse,50,60)
```


## User-defined functions in R

**Functions** are constructs that encapsulate series of **statements** so you do not repeat the same statements all over again when needed. 

```{r, eval=FALSE}
functionName <- function(...) {
  statement
  ...
  <value>  # the result of the function, alternatively: return(<value>)
} 
```

### Example 1

We **define** the following function `add_one`. It takes the value of an **argument** `x` and adds number `1` to it. The sum is the **result** of the function:

```{r}
add_one <- function(x) {
	x + 1
}
```

This function can now be **called** with different values of the argument `x`:

```{r}
add_one(2)
add_one(-1)
```

Type the function name without any `()` to see the **body** (definition) of the function?

```{r}
add_one
```

What is the class of `add_one`? 

```{r}
class( add_one )
```

### Example 2

Now let's define a function to calculate $x^2 + 1$:

```{r}
square_add_one <- function(x) {
	result <- x^2    # choose a variable for temporary result
	result + 1
}

square_add_one(2) # 2^2 + 1 = 5
square_add_one(-1) # (-1)^2 + 1 = 2

```

The following versions of this functions are all equivalent:

```{r, eval=FALSE}
# (v1) One operation per line.
square_add_one <- function(x) {
	result <- x^2
	result <- result + 1
	result    # The last statement is returned as the value.
}
# (v2) No additional 'result' variable needed 
square_add_one <- function(x) {
	x^2 + 1
}
```

### Multiple arguments example

Functions may take as many arguments as needed. 
Recall the function `add_one` defined above: now we would like to have a more `generic` function which allows different values to be added (not only `1`):

```{r}
addValue <- function(x,i) {
  # add i to x
  x + i
}

addValue(2,1)             # <=> add_one(2)
addValue(11,2)            # add 2 to 11
```

You may use the argument names explicitly:  

```{r}
addValue( x = 14, i = 7 ) # add 7 to 14
```

### Logical result example

Now let's write a function `odd` which would take as an argument a whole number and return a logical `TRUE` if the number is an odd number and FALSE otherwise.

A whole number is odd when it is not integer divisible by 2. 
We can check this by taking it's remainder of integer division (`%%` operator) by 2 and see whether it is non-zero:

```{r}
odd <- function(x) {
  x %% 2 != 0
}

odd(31) 
odd( x = 6 ) # more readable than '6 %% 2 != 0' 
```

> what if `x` is  i) a vector or ii) a character value?

```{r, echo=FALSE, results='asis'}
msg <- "1) '# A tibble: 4 x 3', which says that this is a tibble with dimensions 4x3 (4 observations 
and 3 variables), \n\n 2) the atomic type of each variable, in this case character and double, \n\n 3) the row numbers"
qa(msg)
``` 

```{r}
weightRangeSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(weight >= lowerBound & weight < upperBound) %>% 
    summarise(meanHeight=mean(height),
              increasedPulse=sum(pulse1<pulse2), individuals=n())
}
```


## Some observations

  - User-defined functions lead to a more structured and easier to read code.
  - Variables declared inside a function cease to exist once the function terminates.
  - Arguments are given in the same order as they are declared in the function unless the argument names are provided explicitly.


```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
