---
title: "User-defind functions (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
```

## Primary exercises

1. For weight and height measurements we have the `metric` (cm,kg) and 
`imperial` (inch,lbs) systems. The relation between the measurements in different systems 
is (1 kg = 2.2 lbs) and  (1 inch = 2.54 cm). Write the following unit conversion 
functions:
    - cm2meter : convert cm to meter
    - lbs2kg   : convert weight unit pound to kg
    - inch2cm  : convert inch to cm
    


```{r, eval=FALSE, echo=TRUE}
# Some example calls to the functions:
> inch2cm(70)
[1] 177.8
> lbs2kg(10)
[1] 4.545455
> cm2m(190)
[1] 1.9
```

```{r}
cm2m    <- function(x) x*0.01  # cm to meter
lbs2kg  <- function(x) x/2.2   # pound to kg
inch2cm <- function(x) x*2.54  # inch to cm
```

2. Write a function for the survey dataset to report a summary on average pulse and the number of individuals in the given age range. Call it `ageRangeHeightSummary` and  it should take 3 arguments, the survey data, minimum age and maximum age, for example: 


```{r, echo=FALSE, eval=TRUE}
ageRangeHeightSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(age >= lowerBound & age < upperBound) %>% 
    summarise(meanpulse=mean(pulse, na.rm = TRUE),individuals=n())
}
```

```{r}
ageRangeHeightSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(age >= lowerBound & age < upperBound) %>% 
    summarise(meanpulse=mean(pulse, na.rm = TRUE),individuals=n())
}
```

```{r, eval=TRUE, echo=TRUE}
ageRangeHeightSummary(survey, 16,17)
```



3. Take the function `weightRangeSummary` in the lecture notes, make the necessary update such that it includes the weight intervals.

```{r}
weightRangeSummary <- function(x, lowerBound,upperBound) {
  x %>% filter(weight >= lowerBound & weight < upperBound) %>% 
    summarise(meanHeight=mean(height),
              increasedPulse=sum(pulse1<pulse2, na.rm = TRUE),
              weightLowerBound=lowerBound,
              weightUpperBound=upperBound,
              individuals=n())
}
```

## Extra exercises

1. Write a function `odd` which takes a number as argument and returns a logical `TRUE` if the number is an odd number and FALSE otherwise. A whole number is odd when it is not integer divisible by 2. We can check whether a number is divisible by 2 by using the integer division remainder operator `%%`, known as modulo. The operator returns the remainder of integer (whole number) division:      

```{r eval=TRUE, echo=TRUE}
9 %% 2  # remainder after dividing 9 by 2 
8 %% 2  # remainder after dividing 8 by 2
1:10 %% 2 # remainder after dividing 1..10 by 2
```

As you can see the outcome is 0 if the number at the left is integer divisible by 2 and 1 if not.  We can use this fact to test whether a number is odd, i.e. not divisible by 2.   

```{r}
odd <- function(x) {
  x %% 2 != 0
}

odd(31) 
odd( x = 6 ) # more readable than '6 %% 2 != 0' 
```

2. Write the function `even` which tests whether a number is even.

```{r}
even <- function(x) {
  x %% 2 == 0
}
# Alternatively: define the even function in terms of the 'negated' odd function.

even <- function(x) {
  ! odd(x)
}

```

3. Write a function `divisible` which given values `x` and `m` tests whether `x` is integer divisible by `m`.

```{r}
divisible <- function(x,m) {
  x %% m == 0
}
```

