---
title: "User-defind functions (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
```

## Primary exercises

- Write the function `even` which tests whether a number is even.

```{r}
even <- function(x) {
  x %% 2 == 0
}
```

- Write a function `divisible` which given values `x` and `m` tests whether `x` is integer divisible by `m`.

```{r}
divisible <- function(x,m) {
  x %% m == 0
}
```

- There are two systems for weight and height measurements: metric (cm,kg) and imperial (inch,lbs). The relation between the measurements in different systems is  (1 kg = 2.2 lbs) and  (1 inch = 2.54 cm). Write the following unit conversion functions:
    - cm2meter : convert cm to meter
    - lbs2kg   : convert weight unit pound to kg
    - inch2cm  : convert inch to cm

```{r, eval=FALSE, echo=TRUE}
# examples:
> inch2cm(70)
[1] 177.8
> lbs2kg(10)
[1] 4.545455
> cm2m(190)
[1] 1.9
```

```{r}
cm2m    <- function(x) x*0.01  # cm to meter
lbs2kg  <- function(x) x/2.2   # pound to kg
inch2cm <- function(x) x*2.54  # inch to cm
```
