---
title: "Matrices (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

pulse <- read_pulse()
survey <- read_survey()
```


```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

## Extra exercises

1. *Correlated vectors.*  
Let's define a parameter `size <- 12`. Later this will be the number of rows of the matrix.  
From normal distribution let's create a random vector `x <- rnorm( size )`.  
Now let's create another vector `x1` by adding (on average 10 times smaller) noise to `x`: `x1 <- x + rnorm( size )/10`.  
Correlation coefficient of `x` and `x1` should be close to 1.0: check this with function `cor`.  
Finally, create similarly vectors `x2` and `x3` by adding (other) noise to `x`.

```{r}
size <- 12
x <- rnorm( size )
x1 <- x + rnorm( size )/10
cor( x, x1 )
x2 <- x + rnorm( size )/10
x3 <- x + rnorm( size )/10
```

2. *Matrix from vectors; matrix heatmap.*  
Let's merge `x1`, `x2` and `x3` column-wise into a matrix using  `m <- cbind( x1, x2, x3 )`.  
Check class of `m`.  
Show the first few rows of `m`.  
Try a simple heatmap visualisation of the matrix: `heatmap( m, Colv = NA, Rowv = NA, scale = "none" )`.  
Which colors correspond to lowest/highest matrix values?  
Do the vectos appear correlated?
```{r}
m <- cbind( x1, x2, x3 )
class( m )
head( m )
heatmap( m, Colv = NA, Rowv = NA, scale = "none" ) # high is dark red, low is yellow
                # x1, x2, x3 follow similar color pattern, they should be correlated
```

3. *Matrix of correlated and uncorrelated vectors.*  
Repeat the first exercise and create several additional correlated vectors `y1`...`y4` (but not correlated with `x`), of the same length `size`.  
Now build again `m` from columns `x1`...`x3`,`y1`...`y4` in some random order.  
Show again the heatmap; you should see similarity between some columns.
```{r}
y <- rnorm( size )
y1 <- y + rnorm( size )/10
y2 <- y + rnorm( size )/10
y3 <- y + rnorm( size )/10
y4 <- y + rnorm( size )/10
m <- cbind( y4, y3, x2, y1, x1, x3, y2 )
heatmap( m, Colv = NA, Rowv = NA, scale = "none" ) # high is dark red, low is yellow
```

4. *Matrix of correlations.*  
Use `cc <- cor( m )` to build the matrix of correlation coefficients between columns of `m`.  
Use `round( cc, 3 )` to show this matrix with 3 digits precision.  
Can you interpret the values?  
```{r}
cc <- cor( m )
round( cc, 3 ) # 
heatmap( cc, symm = TRUE, scale = "none" )
# E.g. value for (row: x1, col: y1) is the corerlation of vectors x1, y1.
# Values of 1.0 are on the diagonal: e.g. x1 is perfectly correlated with x1.
# Correlations between x, x vectors are close to 1.0.
# Correlations between y, y vectors are close to 1.0.
# Correlations between x, y vectors are close to 0.0.
```



