---
title: "Variables (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```

## Primary exercises

1. *Create a new variable*.  
Make a variable `heightInMeters` that contains your height expressed in meters.

```{r}
heightInMeters <- 1.75
```

2. *Use a variable in an expression*.  
Use the variable `heightInMeters` to calculate your height in cm.

```{r}
heightInMeters * 100
```

3. *Save result to a variable; print a variable*.  
Calculate your height once again and store the result in a variable.  
Show (print) the value of the new variable.

```{r}
heightInCentimeters <- heightInMeters * 100
heightInCentimeters
```
