---
title: "Formulas (practice)"
---

```{r , include=FALSE, eval=TRUE}
pulse <- read_pulse()
survey <- read_survey()
```

## Primary exercises

1. *A formula in a function; simple linear regression*.  
Load the `survey.csv` data to variable `survey`.  
You may make a simple scatter plot with `plot( survey$span2, survey$span1 )`.  
Perform a linear regression of `span1` being dependent on `span2`; store the result in `fit` variable.  
`fit` behaves also like a `list` so show `names` of elements stored in `fit`.  
Find and print the element corresponding to the coefficients of the regression -- do you see there a coefficient for `span2`?  
An intended method to access the coefficients is through `coef` function applied to `fit`. Try it. Do you get the same result as observed above?  

```{r message=FALSE,eval=FALSE}
survey <- read_csv( "survey.csv" )
```

```{r message=FALSE}
plot( survey$span2, survey$span1 )
fit <- lm( span1 ~ span2, data = survey )
fit
names( fit )
fit$coefficients
coef( fit )
```

## Extra exercises

1. *Simple regression diagnostic plots (demo)*.  
Apply `plot` function to the `fit` object calculated in the above exercise.  
These are diagnostic plots for the model produced by `lm`.

```{r}
plot( fit )
```

2. *Another test for association/correlation*.  
In one of the extra list exercises a test for correlation between two random vectors was performed.  
Follow the same steps to test correlation between `span1` and `span2` of `survey`.  
Is correlation between these vectors significant?

```{r}
h <- cor.test( survey$span2, survey$span1 ) # vector notation
h <- cor.test( formula = ~ span2 + span1, data = survey )  # formula notation
h
names( h )
h[[ 'estimate' ]]   # correlation coefficient
h[[ 'p.value' ]]    # significance, is < 0.05, strongly significant association
```

## Multitopic exercises

### (ADV) Fitting multiple simple linear regressions to parts of a table.

Load the `pulse.csv` dataset into the `pulse` variable.  
Let's define the following goal: **separately for each `gender`** calculate
a linear fit of `weight` as a function of `height`.

First `split` the `pulse` table by `gender` into the `pulseByGender` variable.
```{r}
pulseByGender <- pulse %>% split( .$gender )
```

Now, `pulseByGender` is a list and can be accessed with `[[...]]` operator.  
Write the code to perform a linear fit of `weight` as a function of `height` 
**for females only** (obtained from `pulseByGender` list).
```{r}
lm( weight ~ height, data = pulseByGender[[ "female" ]] )
```

The next goal is to perform the above fit multiple times, separately for each gender.  
Define a variable `genders` with all genders present in the split object.
```{r}
genders <- names( pulseByGender )
genders
```

Now a loop is needed.  
A variable `gender` should iterate over all `genders`.  
For each value of `gender` the respective linear fit should be performed.  
Understand and try the following code:
```{r echo=TRUE}
for( gender in genders ) {
  lm( weight ~ height, data = pulseByGender[[ gender ]] )
}
```

The above code calculates all needed fits but does show/store the results anywhere.  
The results need to be stored, for example in a list.  
Understand/modify the code as follows:
```{r echo=TRUE}
fitByGender <- list()
for( gender in genders ) {
  fitByGender[[gender]] <- lm( weight ~ height, data = pulseByGender[[ gender ]] )
}
fitByGender
fitByGender[[ "female" ]]
```

A special function `lapply` allows to write the above code differently.  
In R this is a preferred notation (despite being less intuitive at the first glance).  
Rewrite the `for` loop into `lapply` as follows:
```{r echo=TRUE}
fitByGender <- lapply( genders, function( gender ) {
  lm( weight ~ height, data = pulseByGender[[ gender ]] )
} )
fitByGender
names( fitByGender )
```

As you can see, `fitByGender` is a list but the elements are not named.  
This is because `lapply` names the elements as they were named in the iterated input.  
Check `names` of `genders` (you will see `NULL` names).
```{r}
genders
names( genders )
```

Consequently, set the names of `genders` elements to be equal to the element values.
```{r}
names( genders ) <- genders
genders
```

Now, let's repeat the `lapply` loop and check the result.  
Check (by eye) that `fitByGender[[ 'female' ]]` shows the same result as at the top of this example.
```{r}
fitByGender <- lapply( genders, function( gender ) {
  lm( weight ~ height, data = pulseByGender[[ gender ]] )
} )
fitByGender
fitByGender[[ 'female' ]] # compare to lm( weight ~ height, data = pulseByGender[[ "female" ]] )
```
