---
title: "Vectors (practice)"
---

## Primary exercises

1. *Dietary intakes. (Create a vector, use it in calculation.)*  
Four patients had daily dietary intakes of 2314, 2178, 1922, 2004 kcal.  
Make a vector `intakesKCal` of these four values.  
What is the class of this vector?  
Convert the values into in kJ using 1 kcal = 4.184 kJ.  

```{r}
intakesKCal <- c( 2314, 2178, 1922, 2004 )
intakesKCal
class( intakesKCal )
intakesKCal * 4.184
```

2. *More dietary intakes. (Combining/appending/merging vectors.)*  
Additional set of intakes is provided: 2122, 2616, NA, 1771 kcal.  
Use `c()` to append the new intakes after values in `intakesKCal` and store the result in `allIntakesKCal`.  
Print the combined vector and print its calculated `length`.  

```{r}
intakesKCal2 <- c( 2122, 2616, NA, 1771 )
allIntakesKCal <- c( intakesKCal, intakesKCal2 )
allIntakesKCal
length( allIntakesKCal )
```

3. *The average and total intakes. (Calculating means and sums, skipping missing values.)*  
Calculate `mean` intake for patients in vector `intakesKCal`.  
Next, calculate `mean` intake for patients in vector `allIntakesKCal`.  
Can you explain the result?  
Check help for `?mean`, in particular the `na.rm` argument.  
Use the extra argument `na.rm=TRUE` to calculate the `mean` of non-`NA` elements of `allIntakesKCal`.  
Check help for `?sum` how to omit `NA` elements in sum calculation.  
Now, calculate the total `sum` of `allIntakesKCal` intakes ignoring the `NA` element.
```{r}
mean( intakesKCal )
mean( allIntakesKCal )
# since one element is missing, the mean is unknown
# ?mean, adding argument na.rm=TRUE will omit NA elements
mean( allIntakesKCal, na.rm = TRUE )
# ?sum also allows na.rm=TRUE argument to skip NA elements
sum( allIntakesKCal, na.rm = TRUE )
```

4. *Selecting valid intakes. (Selecting non-missing elements; logical vectors.)*  
Understand the result of `is.na( allIntakesKCal )`.  
Now, negate the above result with `!` operator.  
Use above vectors as argument to `sum` to calculate the number of missing and non-missing elements in `allIntakesKCal`.  
Understand `allIntakesKCal[ !is.na( allIntakesKCal ) ]`.
```{r}
is.na( allIntakesKCal )         # TRUE marks positions with missing data
!is.na( allIntakesKCal )        # TRUE marks positions with available data
sum( is.na( allIntakesKCal ) )                # number of missing elements
sum( !is.na( allIntakesKCal ) )               # number of non-missing elements
allIntakesKCal[ !is.na( allIntakesKCal ) ]    # keeps elements which are not NA
sum( allIntakesKCal[ !is.na( allIntakesKCal ) ] )    # same as sum( allIntakesKCal, na.rm = TRUE )
```

5. *Generating random kcal intakes. (Generating normally distributed random numbers; descriptive statistics.)*  
The code `v <- rnorm( 10 )` would sample 10 numbers from the normal distribution and store them as a vector in `v`.  
Print `v`. 
Then repeat `v <- rnorm( 10 )` and print `v` again. Has `v` changed?  
Next, read the manual of `rnorm` and find how to generate random numbers with given `mean` and standard deviation (`sd`).  
Now, in `v` simulate kcal intake by generating 15 random numbers with `mean=2000` and `sd=300`.  
Print `v` and find by eye the smallest and the largest of these numbers.  
Try to use the functions `min` and `max` on `v` -- have you found the same numbers by eye?  
Calculate the `mean`, `median` and the standard deviation (`sd`) of `v`.
```{r}
v <- rnorm( 10 ) # a vector of random numbers
v
v <- rnorm( 10 ) # another vector of random numbers
v
v <- rnorm( n = 15, mean = 2000, sd = 300 )
v
min( v )
max( v )
mean( v )    # is it close to 2000? try several random v vectors and see the effect of growing n
median( v )
sd( v )      # is it close to 300? try several random v vectors and see the effect of growing n
```

6. *Selecting and counting some kcal intakes. (Selecting elements by a condition; logical vectors.)*  
In `v` simulate kcal intake by generating 15 random numbers with `mean=2000` and `sd=300`.  
Type `v < 2000` and understand the result.  
How to interpret the number produced by `sum( v < 2000 )`?  
How to interpret the number produced by `sum( !( v < 2000 ) )`?  
```{r}
v <- rnorm( n = 15, mean = 2000, sd = 300 )
v
v < 2000             # TRUE corresponds to elements of vector v SMALLER THAN 2000
v[ v < 2000 ]        # selected elements of v smaller than 2000
sum( v < 2000 )      # number of elements in vector v smaller than 2000
sum( !( v < 2000 ) ) # number of elements in vector v GREATER OR EQUAL than 2000
sum( v >= 2000 )     # same as above
```

7. *Head and tail.*  
Often there is a need to quickly look at the beginning (`head`) or at the end (`tail`) of a vector.  
Try these functions to show the first 5 and the last 7 elements of a randomly generated vector `v <- rnorm( 20 )`.
```{r}
v <- rnorm( 20 )
v
head( v, 5 )
tail( v, 7 )
```

8. *Elements of a vector.*  
Let's assume that eight persons had caloric intakes of 2122, 2616, NA, 1771, 2314, 2178, 1922, 2004 kcal.  
Make a vector `intakesKCal` of these eight values (in the given order).  
Use the square brackets to get the 4th element of `intakesKCal`.  
Use the square brackets and the colon operator (`:`) to get the elements from the second to the fifth (inclusive).  
Define another vector `poses` with values 1, 3, 5, 7. Use it get the 1st, 3rd, 5th and 7th element of `intakesKCal`.  
Finally, get the 1st, 3rd, 5th and 7th element of `intakesKCal` typing numbers directly inside `[...]` (without using an extra `poses` variable).
```{r}
intakesKCal <- c( 2122, 2616, NA, 1771, 2314, 2178, 1922, 2004 )
intakesKCal
intakesKCal[ 4 ]
intakesKCal[ 2:5 ]
poses <- c(1,3,5,7)
intakesKCal[ poses ]
intakesKCal[ c(1,3,5,7) ]
```

## Extra exercises

1. *Sequences of numbers*.  
Read help (see `Help` pane) about `seq` function.  
Use it to generate sequence: 10, 7, 4, 1, -2, -5.  
Understand the error message of `seq( 10, -5, 3 )`.

```{r}
seq( 10, -5, -3 )
seq( from = 10, to = -5, by = -3 )
```

2. *Repetitions*.  
Read help (see `Help` pane) about `rep` function.  
Use it to generate sequence: 0, 0, 1, 0, 0, 1, 0, 0, 1.  

```{r}
rep( c( 0, 0, 1 ), 3 )
```
1380 2589 1586 2622 2849 2226
3. *Type conversion to a character vector*.  
Sometimes it is necessary to convert a numerical vector to a character vector.  
Understand what the function `as.character` does for argument `1:5`.

```{r}
1:5
as.character( 1:5 )
class( 1:5 )
class( as.character( 1:5 ) )
```

4. *Type conversion to a numerical vector*.  
Sometimes it is necessary to convert a character vector to a numerical vector.  
Understand what the function `as.numeric` does for argument `c( "1", "-1", "x" )`.  
Note the warning message. Why is there `NA`?

```{r}
as.numeric( c( "1", "-1", "x" ) )
```

5. *Naming vector elements*.  
It is possible to give names to vector elements.  
Define `ages <- c( Amy = 10, 'Dan' = 6, "Eve" = 11, "Eve 2" = 3, Grandma = NA )`.  
Print `ages` and understand `names( ages )`.  
Use square brackets to access age of `Dan`. Try also for `Eve 2`.
```{r}
ages <- c( Amy = 10, 'Dan' = 6, "Eve" = 11, "Eve 2" = 3, Grandma = NA )
ages
names( ages )
ages[ 'Dan' ]
ages[ 'Eve 2' ]

# Another way to create a vector with named elements
ages2 <- c( 10, 6, 11, 3, NA )
names( ages2 ) <- c( "Amy", "Dan", "Eve", "Eve 2", "Grandma" )
ages2
```

6. *(ADV) Write a text vector to a file and read it back*.  
This exercise demonstrates writing a single-column vector (later multicolumn tables will be discussed).  
First choose a name for the file (e.g. `test.txt`) and store it in the variable `fileName`.  
Next, create a **character/text** vector `v` with several text elements.  
Check manual for `writeLines` and try `writeLines(v)` to see in the console what will be written to a file.  
Now, set the argument `con = fileName` and write to the file.  
Use `readLines( con = fileName )` to read the file and put it back to variable `w`.  
Understand `identical( v, w )`. 
```{r}
fileName <- "test.txt"
v <- c( "First line", "Second", "Third", "4th", "5th", "6th" )
v
writeLines( v )                     # writes to the console
writeLines( v, con = fileName )     # writes to a file
w <- readLines( con = fileName )
identical( v, w )   # checks whether v and w are exactly equal
unlink( fileName )  # removes the file
```

7. *(ADV) Write/read a numerical vector; problems*.  
In the previous exercise change `v` to be a vector of some numbers.  
Use `as.character` to make `writeLines` work (do not change `v`).  
Why `identical( v, w )` fails? Check `class(v)` and `class(w)`.  
What conversions of `w` would be needed to make `identical` work?
```{r}
fileName <- "test.txt"
v <- sample( 1:100, 10 )
v
writeLines( as.character( v ) )                 # conversion to character needed
writeLines( as.character( v ), con = fileName )
w <- readLines( con = fileName )
identical( v, w )     # numbers are not the same as their text representation
w <- as.numeric( w )
identical( v, w )     # still not identical; class(v) is different than class(w)
w <- as.integer( w )
identical( v, w )     # now identical
unlink( fileName )    # removes the file
```

8. *(ADV) Merging data from corresponding vectors.*  
Let's assume that we have data on incomes and spendings of several persons.  
The data are provided in three vectors: `nms`, `incomes` and `spendings` (as shown below).  
One person is described by corresponding elements of the three vectors.  
Find a way to calculate:  
    - `balances`: (income minus spending) for each person;
    - name of the person with the largest balance;
    - sort `balances` in descending order and print the names of persons corresponding to this order.
    
    Hints: `which.max`, `names`, `sort`, `decreasing`.

```{r echo=TRUE}
nms <- c( "Amy", "Bob", "Carl", "Dany", "Ela", "Fred" )
incomes <- c( 1380, 2589, 1586, 2622, 2849, 2226 )
spendings <- c( 1198, 2111, 1224, 780, 3266, 2200 )
```

```{r}
balance <- incomes - spendings
balance
max(balance)
which.max(balance)
nms[which.max(balance)]
names(balance) <- nms
balance
sort(balance)
sort(balance, decreasing = TRUE)
names(sort(balance, decreasing = TRUE))
```

