---
title: 'Formulas'
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
```

## Formula concept

A **formula** is a way to tell R that one variable depends on another.  
A formula has a form of an expression. For example, to specify a (statistical) model in which `y` depends on `x`, say:
```{r eval=FALSE}
y ~ x
```

The above code does not directly use values of `y` and `x`. It only declares the form of relation between these two variables.  
The relation might be passed as an argument to functions.  
Such a function, in order to perform a calculation needs a `tibble` (`data.frame`) containing `y` and `x` columns. Typically, the `tibble` will be provided to the function by the argument called `data`.

A formula can be stored in a variable; try:
```{r}
form <- y ~ x
form
class(form)
```

## Example: Student's t-test

Let's use the `pulse` data to test with **Student's t-test** the hypothesis whether between the two `gender` groups there is a significant difference of the mean pulses before the exercise (`pulse1`).

The `t.test` function might be called with two *vectors* of numbers.  
Let's put female pulses to the first vector and the male pulses to the second vector. Try:
```{r}
femalePulse1 <- pulse %>% filter( gender == "female" ) %>% pull( pulse1 )
malePulse1 <- pulse %>% filter( gender == "male" ) %>% pull( pulse1 )
```

Then, we may call `t.test` on these two vectors:
```{r}
t.test( femalePulse1, malePulse1 )
```

But the manual extraction of the `malePulse1` and `femalePulse1` vectors is not necessary for the format of data available in the `pulse` tibble.  
Try the code below to see the `t.test` function used with the formula notation.  
You will see the same result as above:
```{r}
t.test( pulse1 ~ gender, data = pulse )
```

Sometimes, it might be convenient to have the formula stored in a separate variable.  
The following code is also possible:
```{r}
form <- pulse1 ~ gender
t.test( form, data = pulse )
```

Note, that the result of the `t.test` function has a form of a list. Try:
```{r}
h <- t.test( pulse1 ~ gender, data = pulse )
names( h )
h$p.value
```

## Example: Simple linear regression

Let's use the `pulse` data and try to model individual `weight` as a **linear function** of `height`.

The command for linear regression is `lm` (for **linear model**).  
The arguments needs to specify the a tibble containing the `data` and a formula specifying dependence of the variables to be modelled. Try:
```{r}
fit <- lm(weight ~ height, data = pulse)
fit
```

Again, the returned object resembles a list. Check its `names`:
```{r}
names( fit )
```

The formulas might have a more complex form. The exact meaning of such formulas is beyond the scope of this introduction. Here just an example:
```{r}
lm(pulse2 ~ pulse1 + exercise + pulse1:exercise, data=pulse)
```

