---
title: "Matrices"
# date: "`r format(Sys.time(), '%d %B %Y')`"
params:
  basename: "basic_matrices0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
```

```{r, child="_navigate.Rmd"} 
``` 

A **matrix** is a two-dimensional, rectangular structure of elements of _the same type_.  
The elements can be accessed by their **two-dimensional position**  (**row, column**).  
The rows and columns of the matrix might additionally be named and consequently can be also accessed by a pair (**rowname, colname**).

## Create a matrix

A `matrix` can be **manually created** from a vector.  
Depending on the arguments, elements are put to the matrix in a different order:

```{r}
m <- matrix( 1:6, nrow = 2 );
m
m <- matrix( 1:6, nrow = 2, byrow = TRUE );
m
m <- matrix( 1:6, ncol = 2 );
m
```

Test the class of `m`:
```{r}
class( m )
```

Try the function `str` to get a compact summary of matrix content:
```{r}
str( m )
```

## Dimensions

There are several functions to get dimensions of a matrix.

**Number of columns** is provided by:
```{r}
ncol( m )
```

**Number of rows** is provided by:
```{r}
nrow( m )
```

**Dimensions** (two elements: number of rows and number of columns):
```{r}
dim( m )
```

## Names of the columns and rows

`colnames( m )` and `rownames( m )` are used to set and get the vectors with columns and rows names.

Try to **set the names**:
```{r}
m
colnames( m ) <- c( "A", "B" )
m
rownames( m ) <- c( "X", "Y", "Z" )
m
```

To **get the names** use:
```{r}
rownames( m )
colnames( m )
```

## Accessing matrix elements

Use single brackets notation with two arguments `[ row(s), col(s) ]` to **get/set elements**.  
Note, when getting only a single row or a single column the result is returned as a vector, not matrix.

```{r}
m[ 3, 1 ]             # a single element as a vector
m[ 3, ]               # a single row as a vector
m[ , 1 ]              # a single column as a vector
m[ c( 2, 3 ), 1 ]     # two rows/one column, still a vector
m[ c( 2, 3 ), c( "B", "A" ) ]     # a matrix
m[ c( F, T, T ), c( "B", "A" ) ]  # a matrix
```

Notice the difference in the output class when only a single element is requested:
```{r}
class( m[ 3, 1 ] )
class( m[ c( 2, 3 ), c( "B", "A" ) ] )
```

With the `drop` argument you may enforce that always a matrix is returned (never a vector):
```{r}
m[ 3, 1 ]                 # a vector
m[ 3, 1, drop = FALSE ]   # a matrix
class( m[ 3, 1, drop = FALSE ] )
```

Empty row/column argument field means **all rows** and/or **all columns**:
```{r}
m[ , c( "B", "A" ) ]       # all rows, selected columns
m[ c( "Z", "X", "Y" ), ]   # selected rows, all columns
m[ , ]                     # all rows and columns, same as just "m"
```

## Useful matrix functions

Short summary of matrix operations: http://www.statmethods.net/advstats/matrix.html

### Transposition

```{r}
m <- matrix( 1:6, nrow = 3 );
colnames( m ) <- c( "A", "B" )
rownames( m ) <- c( "X", "Y", "Z" )
m
t(m)
```

### Matrix multiplication

```{r}
m
t( m )
m %*% t( m )
```

### Element-wise multiplication

```{r}
m
m+1
m * (m+1)
```

### Row/columns means/sums

```{r}
m
rowMeans( m )
rowSums( m )
colMeans( m )
colSums( m )
```

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
