---
title: "Data sets"
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = NA)
```

Throughout the course we will use the two data sets described below. 

## Pulse 
  
Students in an introductory statistics class (MS212 taught by Professor John Eccleston and Dr Richard Wilson at The University of Queensland) participated in a simple experiment. The students measured their own pulse rate. They were then asked to flip a coin. If the coin came up heads, they were to run in place for one minute. Otherwise they sat without movement for one minute. Then everyone measured their pulse again. The pulse rates and other physiological and lifestyle data are given in the data table.

|Variable | Explanation |
|---|-----------------------------------------------------|
| name | Name of a participant |
| height	  |	Height (cm) |
| weight	  |	Weight (kg) |
| age	    |	Age (years) |
| gender	  |	Sex (male/female) |
| smokes	  |	Regular smoker? (yes/no) |
| alcohol	|	Regular drinker? (yes/no) |
| exercise	|	Frequency of exercise (high/moderate/low) |
| ran	    |	Whether the student ran or sat between the first and second pulse measurements (ran/sat) |
| pulse1  	|	First pulse measurement (rate per minute) |
| pulse2	  |	Second pulse measurement (rate per minute) |
| year		  | Year of the class (1993 - 1998) |

The pulse data set is available in the data folder as comma-delimited text: `pulse.csv`. It is also available in SPSS-format as `pulse.sav`.


```{r, message=FALSE, echo=TRUE}
pulse <- read_csv("data/pulse.csv")
pulse 
```

 
## Survey

This data frame contains the responses of 233 Statistics I students at the University of Adelaide to a number of questions. It is a slightly modified version of the `survey` data from the `MASS` pacakge.

|Variable | Explanation |
|---|-----------------------------------------------------|
| name | Name of a participant  |
| gender	  |	Sex (male/female) |
| span1   | Span (distance from tip of thumb to tip of little finger of spread hand) of writing hand (cm) | 
| span2   | Span of non-writing hand (cm) | 
| hand  | Writing hand of student (left/right) |
| fold | Fold your arms! which is on top? (right/left/neither) |
| pulse  	|	Pulse measurement (rate per minute)|
| clap | Clap your hands! which is on top? (right/left/neither) |
| exercise	|	Frequency of exercise (freq/some/none)|
| smokes	  |	How much the student smokes (heavy/regul/occas/never)|
| height	  |	Height (cm) |
| m.i	  |	whether the student expressed height in imperial (feet/inches) or metric (centimetres/metres) units. (metric/imperial) |
| age	    |	Age of the student (years)|

The pulse data set is available in the data folder as comma-delimited text: `survey.csv`.


```{r, message=FALSE, echo=TRUE}
survey <- read_csv("data/survey.csv")
survey 
```


## SPSS files

For SPSS files (.sav) we need the function `read_spss` from the R package `haven`. Try `library(haven)`, if the package is not installed then you'll get an error `threre is no package called 'haven'`, in that case you'll need to install it first using:

```{r, eval=FALSE}
install.packages("haven")
```

Once the `haven` package is installed then we need to load it into the current R session first using `library(haven)` upon which we can read `SPSS` files, e.g. `my_spss_file.sav`) :  

```{r, message=FALSE, eval=FALSE} 
## CAREMA
library(haven)
msf <- read_spss(file = "my_spss_file.sav") 
str(msf)
```


## Download data 

pulse.csv<a href="data/pulse.csv" class="fas fa-download" download target="_blank"></a>
survey.csv<a href="data/survey.csv" class="fas fa-download" download target="_blank"></a>
