---
title: "Tibble (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
# library(rooc)
# source("../R/Doc.R")
# source("../R/Globals.R")
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

1. Create tibble  
  
a. Create a tibble `exercise_group` for a group of individuals with names {Sonja, Steven, Ines, Robert, Tim} with their heights {164, 188, 164, 180, 170}, weights {56.0, 87.0, 54.0, 80.0, 58.5} and frequency of exercise {high, high, low, moderate, low}.

```{r} 
exercise_group <- tibble(name=c("Sonja" , "Steven", "Ines", "Robert", "Tim" ),
            height=c(164, 188, 164, 180, 170),  
            weight=c(56.0, 87.0, 54.0, 80.0, 58.5),
            exercise=c("high", "high", "low", "moderate", "low")
     )  
exercise_group
```


b. Update the tibble `exercise_group` with `Ella` and `Oscar`, leave their respective `height`, `weight` and `exercise` values as missing (`NA`). Avoid copy/paste from (a) with inclusion of new names, instead try to reuse the columns inside `exercise_group`.


```{r}
exercise_group <- tibble(name=c(exercise_group$name, "Ella", "Oscar"),
                         height=c(exercise_group$height,NA,NA),
                         weight=c(exercise_group$weight,NA,NA),
                         exercise=c(exercise_group$exercise,NA,NA)
                  )
exercise_group       
```

c. Add the `sex` variable to `exercise_group` with values `male` and `female`.

```{r}
exercise_group <- tibble(name=exercise_group$name,
                         height=exercise_group$height,
                         weight=exercise_group$weight,
                         exercise=exercise_group$exercise,
                         sex=c('female','male','female','male','male','female','male')
                  )
exercise_group
```

2. Create a tibble which keeps track of the smoking habits over the years of `Julio` age 21 started smoking at 17 and stopped in 2020, `Camille` age 20 started smoking in 2021 and `Travis` 19 started at age 16.     

```{r echo=FALSE}
this_year <- as.numeric(format(Sys.time(),"%Y"))
```

```{r}
# List the information first as below, here NA (missing value) is interpreted as not 
# stopped, i.e. still smoking to present date.
# 
# name     age  start          stop
# Julio    21   2022-(21-17)   2020
# Camille  20   2021           NA
# Travis   19   2022-(19-16)   NA

tibble(name=c("Julio", "Camille","Travis"), 
       age=c(21,20,19), 
       start=c(2018,2021,2019), 
       stop=c(2020,NA,NA))
```

### tibble subset

1. Take the tibble `exercise_group` from the previous exercise and create a new tibble `exercise_group_sub` without the `height` and `weight` variables by selection `[`.

```{r}
exercise_group_sub <- exercise_group[c("name","exercise")]
exercise_group_sub
```

2. Create a tibble called `exercise_group_sub` with the 1st and 3rd column.

```{r}
exercise_group_sub <- exercise_group[c(1,3)]
exercise_group_sub
```


### Extract variables as vectors

1. Given the tibble `favourite_colour`, how old were the subjects by the end of 2021?

```{r, echo=FALSE}
favourite_colour  <- tibble(name=c("Lucas","Lotte","Noa","Wim","Marc","Lucy","Pedro"), 
                           year=c(1995,1995,1995,1994,1990,1993,1992), 
                           colour=c("Blue","Green","Yellow","Purple","Green","red","Blue"))
```

```{r}
2021 - favourite_colour[["year"]] # alternatively:  2021 - favourite_color$year
```

2. What is the mean height in `exercise_group`? Use *mean* function (see ?mean).

```{r}
mean(exercise_group[["height"]]) 
```

### Read tibbles from file

1. Read `pulse.csv` data set into R and inspect its dimensions.

```{r, message=FALSE,eval=FALSE}
pulse <- read_csv(file = "pulse.csv")
```

```{r}
# two alternatives i) nrow and ncol function, ii) dim function.
nrow(pulse) # number of rows
ncol(pulse) # number of columns
dim(pulse)  # dimensions (rows, columns)
```

2. Read `survey.csv` data set into R. 

```{r, message=FALSE,eval=FALSE}
survey <- read_csv(file = "survey.csv")
```

  - Inspect the dimensions.

```{r}
dim(survey)
```

  - Show the first 9 and the last 7 rows.

```{r}
head(survey,9)
tail(survey,7)
```

  - Calculate the mean `age`.
    
```{r}
mean(survey$age) 
```

  - Calculate the mean height in survey data.

```{r}
# Here we use a second argument 'na.rm = TRUE' because there are missing values (NA) in 
# the variable height. By default the mean function returns NA if it first argument, in this 
# case variable 'height', contains any NA. The second argument 'na.rm = TRUE' changes this 
# behaviour by disregarding the observations with missing height and calculates the mean 
# of observations for which the height is available. 
# 
mean(survey$height, na.rm = TRUE) 
```


## Extra exercises

1. In survey data:

  a) What is the mean height of the last 30 observations?

```{r}
survey_last_30 <- tail(survey,30)
mean(survey_last_30$height, na.rm = TRUE) # <=> mean(survey_last_30[["height"]], na.rm = TRUE)
```


  b) The variable `age` is the last column in the survey data. Make a tibble where the variable `age` comes directly after `name`.
  
```{r}
# Some solutions
#
# 1) list names : exhaustive 
survey[c("name", "age", "gender", "span1" ,"span2","hand","fold", 
   "pulse", "clap", "exercise", "smokes", "height", "m.i")] 
# 2) indices  
survey[c(1,ncol(survey),2:(ncol(survey)-1))]

# 3) The select(...) function is a more concise solution which will be discussed 
#    in the next section. 
```
  

2. Create the favourite_colour tibble from the lecture but now with `colour` variable as a `factor`. Print the counts for each level.

```{r}
favourite_colour  <- tibble(name=c("Lucas","Lotte","Noa","Wim","Marc","Lucy","Pedro"), 
                           year=c(1995,1995,1995,1994,1990,1993,1992), 
                           colour=factor(c("Blue","Green","Yellow","Purple","Green","red","Blue")))
fct_count(favourite_colour[['colour']])
```


####  `[<row>, <column>]` : row and column selection based on range of indices.

Using single square bracket `[` one could select range of rows, columns or a combination. For example take the `exercise_group` tibble from the primary exercises above, then:


```{r, eval=TRUE, echo=FALSE}
exercise_group <- tibble(name=c("Sonja" , "Steven", "Ines", "Robert", "Tim" ),
            height=c(164, 188, 164, 180, 170),  
            weight=c(56.0, 87.0, 54.0, 80.0, 58.5),
            exercise=c("high", "high", "low", "moderate", "low")
     )

exercise_group <- tibble(name=c(exercise_group$name, "Ella", "Oscar"),
                         height=c(exercise_group$height,NA,NA),
                         weight=c(exercise_group$weight,NA,NA),
                         exercise=c(exercise_group$exercise,NA,NA)
                  )

exercise_group <- tibble(name=exercise_group$name,
                         height=exercise_group$height,
                         weight=exercise_group$weight,
                         exercise=exercise_group$exercise,
                         sex=c('female','male','female','male','male','female','male')
                  )
```


```{r, echo=TRUE, eval=TRUE}
exercise_group[c(2,3),]  # returns rows in the range 2 to 3
exercise_group[,c(1,3)]  # returns columns in 1 and 3
exercise_group[c(2,3),c(1,3)]  # combination the above
```

3. Reproduce the following tibbles from `exercise_group`:

a.

```{r, eval=TRUE}
exercise_group[c(1,4),]
```

b. 

```{r, eval=TRUE}
exercise_group[2:5, c(1,3,4)]
```

c.

```{r, eval=TRUE}
exercise_group[6:7,2:4]
```

