---
title: "Summarise (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

In the survey data:  

1. Take a look at the table with `glimpse` and inspect its dimensions and types.

```{r}
survey %>% glimpse()
```

2. Summarise on mean hand spans `{span1,span2}` and  median `pulse`.

```{r}
# 'pulse' variable has missing values (NA), therefore the use 
# of 'na.rm=TRUE' in the function 'median'.
#
survey %>% summarise(meanWritingHand=mean(span1), 
                     meanNonWritingHand=mean(span2), 
                     medianPulse=median(pulse, na.rm=TRUE))
```

3. Summarise on mean `age` and median `height`.

```{r}
survey %>% summarise(meanAge=mean(age), medianHeight=median(height, na.rm=TRUE))
```

4. Count the number of males and females.

```{r}
survey %>% count(gender)
```

5. Produce the frequency table on `gender` and `fold`. Do the same including `clap`. 

```{r}
survey %>% count(gender, fold)       # (1)
survey %>% count(gender, fold, clap) # (2)
```

## Extra exercises

1. Produce the frequency table on `gender` and `smokes`,  only females. 

```{r}
survey %>% filter(gender=="female") %>% count(gender,smokes)
```

2. Produce the frequency table on `gender`, `exercise` and `smokes` but only for male smokers.

```{r}
# smoker is someone that smokes 'occas', 'regul' or 'heavy', i.e. smokes !='never'
#
survey %>% filter(gender=="male" & smokes!="never") %>% count(gender,exercise, smokes)
```

3. Do individuals all have different names? Put differently, is each observation uniquely identifiable with the variable `name`?

```{r}
# The answer is no, because the number of unique names is less than 
# total number of observations:
#
nrow( survey %>% distinct(name)) == nrow(survey)
```

4. Find pairs of variables which can uniquely identify each observation. Use `name` and one other variable for this purpose.

```{r}
# We see here that name and age combination is unique in the entire table.
nrow( survey %>% distinct(name, age)) == nrow(survey)
nrow( survey %>% distinct(name, height)) == nrow(survey) # The same is true for name and height
```


