---
title: "Select (practice)"
# author: "R. Monajemi"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
```

```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

Apply the following to survey data: 

1) Select *personal information* {name, age, gender, height} into a new tibble `survey_personal_info`.

```{r}
survey_personal_info <- select(survey, name, age, gender, height)
```

2) Select *personal information* as previous exercise into a new tibble `survey_personal_info` but with uppercase initials, e.g. `Name`, `Age` etc. 

```{r}
survey_personal_info <- select(survey, Name=name, Age=age, Gender=gender, Height=height)
```

3) Reorder the variables as such that name,age and gender appear as first, second and the third column followed by the remaining
variables. 

```{r}
select(survey, name,age,gender,everything())
```

4) Deselect variables that relate to hand and/or arm (e.g. span1, span2, hand, etc.). See also description of the [survey data](data.html#survey)

```{r}
select(survey, -span1,-span2,-hand,-fold,-clap)
```

5) Select the top 20 names along with gender.

```{r}
survey_sub <- select(survey, name,gender)
head( survey_sub , 20)
# shorter solution without intermediate variable 'survey_sub' :
# head(select(survey,name),20)
```


6) Reproduce the following tibbles (note that variables are renamed and reshuffled):

  6.1) First 5 observations.

```{r, eval=TRUE}
head( select(survey, SPAN1=span1, SPAN2=span2, everything()), 5)
```

  6.2) Last 3 observations.

```{r, eval=TRUE}
tail( select(survey, hand,fold,clap, everything()), 3)
```



## Extra exercises

1) Rename the `m.i` variable to `system`.
  
```{r}
# Option 1: very tedious, you need to type all the variable names and 
# only rename the 'm.i' variable to 'system'
#
select(survey , name, gender, span1, span2, hand, fold, pulse, clap, 
       exercise, smokes, height, system=m.i, age)
# Option 2: shorter, but side-effect is that m.i (now system) comes at 
# the front.
select(survey, system=m.i, everything()) 
# Option 3: use rename function (see ?rename). 
rename(survey,system=m.i)
```


2) Select `name` along with all categorical variables into a new tibble `survey_cats`.

```{r}
# Categrical data: variables which take on categories as values, e.g. 
# 
# gender   : {male, female}
# hand     : {left, right}
# fold     : {left, right, neither} 
# clap     : {left, right, neither} 
# exercise : {freq, some, none}
# smokes   : {heavy, regul, occas, never}
# m.i      : {metric, imperial}
#
# 
survey_cats <- select(survey, name, gender, hand, fold, clap, exercise, smokes, m.i)
survey_cats
```


3) Create a new tibble `survey_nums` with `name` and all numerical variables.

```{r}
survey_nums <- select(survey, name, span1, span2, pulse, height)
survey_nums

```

