---
title: "Filter (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

In the survey dataset: 


1. Select teenagers, assume age range between and including 10 and 19.
  
```{r}
filter(survey, 10<=age & age<20)
```

2. Select all females with pulse equal to 60

```{r}
filter(survey, pulse==60 & gender=="female")
```

3. Select all male teenagers with pulse above 60.

```{r}
filter(survey, pulse>60 & gender=="male" & (10<=age & age<20) )
```

4. How many males do smoke and never exercise?

```{r}
# The conditions are 'do smoke' and 'never exercise'. With 'do smoke' we have the categories {regul,occas,heavy} therefore 'or' (|) logical connective is in place. For 'never exercise' we have the category 'none'. Since both conditions must be true we will have the 'and' (&) logical connective.

nrow( filter(survey, (smokes=="regul" | smokes=="occas" | smokes=="heavy") & exercise=="none" & gender=="male" ) )
# Alternatively: a shorter condition for 'smokes' is smokes!="never". It means accept all values for 'smokes' as long as it is not equal to "never" and those are exactly {regul,occas,heavy}.

nrow( filter(survey, smokes!="never" & exercise=="none" & gender=="male" ) )
```

5. How many females never smoke and frequently exercise?

```{r}
nrow( filter(survey, smokes=="never" & exercise=="freq" & gender=="female") )
```

6. Produce the following tibbles:

    6.1 Personal information {Name,Age,Gender,Height} of all teenagers.

    ```{r}
    filter(select(survey, Name=name, Age=age, Gender=gender, Height=height ), 10<=Age & Age<20)
    ```

    6.2 Personal information of males with Height between and inclusive 170 to 180.
    
    ```{r}
    filter(select(survey, Name=name, Age=age, Gender=gender, Height=height),
            170<=Height & Height<=180 & Gender=="male")
    ```

7. Has survey data missing `pulse` values? If so how many? What about `age`? 

```{r}
nrow(filter(survey, is.na(pulse))) # number of missing 'pulse' values.
filter(survey, is.na(age)) # empty result set, no missing values in 'age'.
```


## Extra exercises

1. What is the percentage of males who never smoke and frequently exercise? Do the same for female.

```{r}
# men
none_smoker_sportsmen <- nrow( filter(survey, smokes=="never"  & 
                                              exercise=="freq" & 
                                              gender=="male") )
total_men <- nrow( filter(survey, gender=="male") ) 
none_smoker_sportsmen / total_men

# women
none_smoker_sportswomen <- nrow( filter(survey, smokes=="never"  & 
                                                exercise=="freq" & 
                                                gender=="female") )
total_women <- nrow( filter(survey, gender=="female") ) 
none_smoker_sportswomen / total_women
```

2. What is the age range in teenagers? You may use the range function (?range).

```{r}
teenagers <- filter(survey, 10<=age & age<20) 
range(teenagers[["age"]])
```

3. How many males do smoke and never exercise? Use '%in%' operator see ?match for more details.

```{r}
nrow( filter(survey, (smokes %in% c("regul","occas", "heavy")) & exercise=="none" & gender=="male" ) )
```

4. Recall [where](https://dplyr.tidyverse.org/reference/select.html) helper function in `select` extra exercises. Hint: see `anyNA` function (see ?anyNA for the manual).

    4.1 Select all variables from survey data having missing values. 
    
    ```{r}
    select(survey, where(anyNA))
    ```

    4.1 Count the number of missing values for variable you found in 4.1. 
    
    ```{r}
    nrow(filter(survey, is.na(pulse)))     # pulse 
    nrow(filter(survey, is.na(height)))    # height  
    nrow(filter(survey, is.na(m.i)))       # m.i
    ```
    
    4.3 In (4.2) you found that `height` and `m.i` had the same number of missing values, n=27. 
    How can you confirm or refute that missing values of `height` and `m.i` occur in the same 
    observations? 

    ```{r}
    # The number of observations for which both 'height' and 'm.i' are missing at the same 
    # row must be equal to 27. If true then we can confirm that the missing values of `height` 
    # and `m.i` occur in the same observation, refute otherwise.
    nrow( filter(survey, is.na(height) & is.na(m.i)) ) == 27
    ```
    
