---
title: "Select (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

Apply the following to survey data: 

1. Select *personal information* {name, age, gender, height} into a new tibble `survey_personal_info`.

```{r}
survey_personal_info <- select(survey, name, age, gender, height)
```

2. Select *personal information* as previous exercise into a new tibble `survey_personal_info` but with variable names initials in uppercase, e.g. `Name`, `Age` etc. 

```{r}
survey_personal_info <- select(survey, Name=name, Age=age, Gender=gender, Height=height)
```

3. Reorder the variables in survey dataset as such that name,age and gender appear as first, second and the third column followed by the remaining
variables. 

```{r}
select(survey, name,age,gender,everything())
```

4. Deselect variables that relate to hand and/or arm (e.g. span1, span2, hand, etc.). See also description [survey](S01L05l_course_data0.html) data.

```{r}
select(survey, -span1,-span2,-hand,-fold,-clap)
```

5. Select the top 20 names along with gender.

```{r}
# 1)
survey_sub <- select(survey, name,gender)
head( survey_sub , 20)
```


```{r}
# 2) shorter solution without intermediate variable 'survey_sub' :
head(select(survey,name),20)
```


6. Reproduce the following tibbles (note that variables are renamed and reshuffled):
    
    
    6.1 First 5 observations.
    

    ```{r, eval=TRUE}
    # Remark: by enclosing select(...) as the first argument of 'head' function you 
    # can avoid creating intermediate variables.
    
    head( select(survey, SPAN1=span1, SPAN2=span2, everything()), 5)
    ```

    6.1 Last 3 observations.

    ```{r, eval=TRUE}
    tail( select(survey, Hand=hand,Fold=fold,Clap=clap, everything()), 3)
    ```


## Extra exercises

1. Rename the `m.i` variable to `system`.
  
```{r}
# 1) Very tedious, you need to type all the variable names and 
# only rename the 'm.i' variable to 'system'
#
select(survey , name, gender, span1, span2, hand, fold, pulse, clap, 
       exercise, smokes, height, system=m.i, age)
# 2) Shorter but side-effect is that m.i (now system) comes at 
# the front.
select(survey, system=m.i, everything()) 
# 3) Use rename function (see ?dplyr::rename). 
rename(survey,system=m.i)
```


2. Select `name` along with all categorical variables into a new tibble `survey_cats`.

```{r}
# Categrical data: variables which take on categories as values, e.g. 
# 
# gender   : {male, female}
# hand     : {left, right}
# fold     : {left, right, neither} 
# clap     : {left, right, neither} 
# exercise : {freq, some, none}
# smokes   : {heavy, regul, occas, never}
# m.i      : {metric, imperial}
#
# 
survey_cats <- select(survey, name, gender, hand, fold, clap, exercise, smokes, m.i)
survey_cats
```


3. Create a new tibble `survey_nums` with `name` and all numerical variables.

```{r}
survey_nums <- select(survey, name, span1, span2, pulse, height, age)
survey_nums
```


4. For this exercise you'll need an additional helper function `where` explained  
  [here](https://dplyr.tidyverse.org/reference/select.html). 

    4.1 Reproduce the result from the previous exercise (3) without dictating all numerical 
    variable names. Hint: you'll also need `is.numeric` function (see ?is.numeric for help).
    
    ```{r}
    bind_cols(survey['name'], select(survey, where(is.numeric)))
    ```
  
    4.2 Select all non-numerical variables. 
    
    ```{r}
    # 1) 
    select(survey,! where(is.numeric))
    ```
    
    ```{r}
    # 2) Since there are no other non-numerical types the following is also a correct solution.
    select(survey, where(is.character))
    ```

#### Selection by pattern matching

In data sets with large number of variables, finding variables will become tedious. Several helper functions are available to speed up the variable name search.

> starts_with(), ends_with() and contains()

The functions help to find _fixed patterns_ in variable names:

```{r}
# select variables starting with character 'a'
select(pulse, starts_with("a"))
# select variables ending with 'e'
select(pulse, ends_with("e"))
# select variables containing character 'i' 
select(pulse, contains("i"))
```

The helper functions can be used with _logical operators_ {!,|,&} which will be explained later. You have already encountered one in the lecture on Useful R functions, `!`, the _negation_ operator. In short it complements the results. For example, above we could select variables which started with character 'a' with `select(pulse, starts_with("a"))` which resulted into a tibble with the two variables `age` and `alcohol`. Using `!` in front of the helper function in the expression will produce the complement of the previous result, namely all variables that do not start with `a`:

```{r}
select(pulse, ! starts_with("a"))
```

Note that `age` and `alcohol` do not occur in the result.

There are several other helper functions which fall beyond the scope of this lecture, visit [here](https://dplyr.tidyverse.org/reference/select.html) for more details.


5. Select variables, from survey data, by pattern matching. 

    5.1 Select variables that end with 'e'.
    
    ```{r}
    select(survey, ends_with('e'))
    ```

    5.2 Select variables that start with 's'.
    
    ```{r}
    select(survey, starts_with('s'))
    ```

    5.3 Select hand span variables using a helper function.
    
    ```{r}
    # 1)
    select(survey, contains('span'))
    # 2) 
    # select(survey, starts_with('span'))
    ```
  
