---
title: 'Select'
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

> Use the `select` function to select variables (columns) from a tibble. 

Given a tibble `select` can be used to : 

  - take a collection of variables and simultaneously **reorder** and/or **rename** the variables
  - **reshuffle** the variable order

Let's take the `pulse` dataset:   

```{r}
pulse 
```

`select` takes as it first argument a tibble, followed by a comma separated list of variables of your choice and returns a tibble with those chosen variables:

```{r}
select(pulse, name, age)
```

> After this selection, does `pulse` tibble still contain the variables 'name' and 'age'?

```{r, echo=FALSE, results='asis'}
msg <- "Yes, 'select' returns the selection as a tibble and does not modify the underlying tibble. You can check this by entering 'pulse' in the R console." 
qa(msg)
``` 

If you want to keep your selection as a separate tibble you'll need to assign the result into a new environment variable, e.g. `pulse_name_age`:

```{r}
pulse_name_age <- select(pulse, name, age)
pulse_name_age
```

##  Variable order

The order of the selected variables is reflected in the resulting tibble:  

```{r}
select(pulse, age, name )
```

## Deselect variables 

You may also deselect variables, with other words the complement of your selection. This is done by the `-` sign:

```{r}
select(pulse, -smokes, -alcohol)
```

## Select and rename 

With selection it is possible to change the variable names simultaneously:  

```{r}
select(pulse, FirstName = name, Age = age)
```

> What is the variable name in the pulse dataset, 'Age' or 'age'?

```{r, echo=FALSE, results='asis'}
msg <- "age, this because we only run select and do not store its result with assignment ('<-') back into pulse tibble."
qa(msg)
``` 

## Reshuffle variables 

With `select` we can reshuffle the variables in their positions in the tibble. When a data set contains  large number of variables, you may want to bring the more 'important' variables in front for inspection. You can do this with `select` in combination with a helper function `evertything()`:   

```{r}
select(pulse, name, age, everything()) 
```

`everything` function lists all other variable other than `name` and `age` and `select` function places them after `name` and `age`.







