---
title: "Mutate (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises


1.  In the survey dataset add a new column `feet` with heights reported in feet unit (1 foot = 30.48 cm).

```{r}
mutate(survey, feet=height/30.48)
```

2. In the survey dataset add a new column `diffWritingHandSpan` : the difference of span1 (writing hand) and span2 (non-writing hand). 

```{r}
mutate(survey, diffWritingHandSpan=span1-span2)
```

3. In the pulse dataset add new weight variables `pound` and `stone` (1 kg  = 2.20462 pound = 0.157473 stone).

```{r}
mutate(pulse, pound = weight * 2.20462, stone = weight * 0.157473 )

```

4. In the survey dataset convert the variables `smokes` from character to factor with levels {"never","occas","regul", "heavy"}, in that order.

```{r}
mutate(survey, smokes = fct_relevel(factor(smokes), "never","occas","regul", "heavy")) 
```

## Extra exercises

In the survey dataset: 

1. Add a new column `diffHandSpan` : the absolute difference between span1 (writing hand) and span2 (non-writing hand). Hint: use `abs` function (?abs).  

```{r}
mutate(survey, diffWritingHandSpan=abs(span1-span2))
```

2. Change `height` unit from cm to inch (1 cm = 0.393701 inch).

```{r}
mutate(survey, height=height*0.393701)
```


3. Produce the tibble containing the personal information of only those having height (in foot unit) 
between and including 6.0 and 6.5 feet.     

```{r}
filter( 
  mutate( 
    select(survey, Name=name, Age=age, Gender=gender, Height=height), feet=Height/30.48 
  ), feet >= 6.0 & feet <= 6.5 
) 

```


4. How many observations are in the following call, please try to reason before running the statement. 

```{r, eval=FALSE, echo=TRUE}
filter(mutate(filter(survey, height> 190),feet=height/30.48),height<=190 )
```


5. The following table is one of many BMI classifications:

<center>
```{r, eval=TRUE, echo=FALSE}
library("htmlTable")
htmlTable( data.frame(classification=c("underweight","normal","overweight","obese"),
           bmi=c("<18.5","18.5-24.9","25-29.9","\\>=30"), stringsAsFactors = F),  col.rgroup =
c("none", "#F7F7F7") )
```
</center>

Add the variable `BMI_class` to the pulse data set. Note that you will need to `round` (see ?round an additional argument `digits`) the BMI calculation to one decimal digit to fit into classification values in the table given above. Use the helper function [case_when](https://dplyr.tidyverse.org/reference/case_when.html). 

```{r}
pulse_bmi <- mutate(pulse, BMI=round( weight/((height/100)^2), digits = 1)  ) 
mutate(pulse_bmi, BMI_class = case_when(
                BMI < 18.5 ~ "underweight",
                BMI >= 18.5 & BMI <= 24.9 ~ "normal",
                BMI >= 25   & BMI <= 29.9 ~ "overweight",
                BMI >= 30 ~ "obese"
              )                     
      )
```


6. Age classification:

<center>
```{r, eval=TRUE, echo=FALSE}
library("htmlTable")
htmlTable( data.frame(classification=c("adult","adolescent","child","infant"),
           age=c("\\>19","\\>=10 and <=19","\\>=1  and <=9","<1"),
           notation=c("(19,&infin;)","[10,19]","[1,9]","(-&infin;,1)"), stringsAsFactors = F),  col.rgroup = c("none", "#F7F7F7") )
```
</center>

Add the variable `age_group` to the survey dataset. 

```{r}
mutate(survey, age_group = case_when(
                age > 19 ~ "adult",
                age >= 10 & age <= 19 ~ "adolescent",
                age > 1   & age <= 9 ~ "child",
                age <= 1 ~ "infant"
              )                     
      )
```

