---
title: 'tibble vs data.frame*'
---

```{r, echo=FALSE, message=FALSE}
library(rooc)
source("../R/Doc.R")
source("../R/Globals.R")
```

A **tibble** is the upgraded version of base R **data.frame** data type with some convenient additional features, moreover, it is the standard data structure in [tidyverse](https://www.tidyverse.org/). In the following I'll layout the main features and differences in comparison with data.frame. Let's create a data.frame and a tibble manually with the same content.

## Create

Take for example the variables `name`, `year` and `favourite` to represent a person's name, birth year and favourite colour    

__data.frame__
```{r, echo=TRUE }
t1 <- data.frame(name=c("Lucas","Lotte","Noa"), 
                 year=1995, 
                 favourite=c("Blue","Green","Yellow"))
t1
```
        
__tibble__
```{r, echo=TRUE}
# tibble
t2  <- tibble(name=c("Lucas","Lotte","Noa"), 
             year=1995, 
             favourite=c("Blue","Green","Yellow"))
t2
```

> What is the difference between tibble and data.frame output?

```{r, echo=FALSE, results='asis'}
msg <- "The tibble output has two additional pieces of information i) the dimension of 
the data frame and ii) the column types between angle brackets (<...>)."
qa(msg)
``` 

Let's now inspect the types of the tibble and data.frame objects:   

```{r}
class(t1) # data.frame
class(t2) # tibble
```

Clearly they are both data frames, however tibble has additional types associated with it, i.e. `tbl` and `tbl_df`. This is how you can recognize tibble objects. 


## strings as factors

> What are the base types of the data.frame `t1` columns?


```{r, echo=FALSE, results='asis'}
msg <- "'name' and 'favorite' are factors and year is of type double."
qa(msg)
``` 


`data.frame` function by default converts all character variables into [factor](forcats_factor0.html):

```{r}
class(t1$name)
class(t1$favorite)
```

In general, not all character variables are categorical, for example it makes less sense to have the variable 
`name` as a factor, whereas `favourite` is a good candidate. `tibble` does not make any assumptions on the type of the variable and treats all character variables as base type character `chr`.

## column selection 

Columns selection with tibble always results into a tibble, this is not the case with data.frame. 

```{r}
# tibble
t2[3]  
t2[,3]
# data.frame
t1[3]  
t1[,3]
```

Selecting a single column with data.frame using comma results in a vector, and not a data frame as is the case when selecting from a tibble.

## row names 

data.frame type allow row names and they are often used as a kind of variable, however they are dropped when using tibbles and their usage is discouraged.

> Can you think of an argument against using rownames?

```{r, echo=FALSE, results='asis'}
msg <- "The argument against row names is that they are not formally part of the data and therefore 
can not be maintained consistently throughout analysis pipelines."
qa(msg)
``` 


Let's look at the current row names of t1 and t2:

```{r}
rownames(t1)
rownames(t2)
```

The function `rownames` returns a character vector of 1 to the number of rows in the data frame. It can also be used to assign row names:

 
```{r}
t <- t1 # t is a copy of data.frame t1
rownames(t) <- letters[1:nrow(t1)]
t
rownames(t)
```

Let's try to assign row names to a tibble:

```{r}
rownames(t2) <- letters[1:nrow(t2)]
```

It results into a warning and the tibble is unchanged:

```{r}
t2
```


> t is a data.frame with row names a,b and c. What is the result of `rownames(as_tibble(t1))`? 

```{r, echo=FALSE, results='asis'}
msg <- "[1] \"1\" \"2\" \"3\""
qa(msg)
``` 

## partial column name matching

data.frame allows partial column name matching. for example in t1 data.frame all sequences f, fa, fav, ... will match to the column favourite:

```{r}
t1$f    # t1 is a data.frame
t1$fa
t1$fav
```

With tibble this will result into a NULL and a warning:

```{r}
t2$fa   # t2 is a tibble
```

> Can you think of an argument why this choice of partial matching is dropped in tibble?

```{r, echo=FALSE, results='asis'}
msg <- "Having a partial call such as t1\\$fa assumes that t1 will never be updated with a new 
column called fa. Because when that happens then your R script with call t1\\$fa will no 
longer result into t1\\$favourite but to a more exact match t1\\$fa."
qa(msg)
``` 


## tibble $\leftrightarrow$ data.frame

Often you may need to switch between `data.frame` and `tibble`. You can convert an existing `tibble` into 
`data.frame` type using the function `data.frame()`. Take for example `favourite_colour` from [tibble](dplyr_tibble0.html) lecture notes:  

```{r}
favourite_colour  <- tibble(name=c("Lucas","Lotte","Noa","Wim"), 
                           year=c(1995,1995,1995,1994), 
                           colour=c("Blue","Green","Yellow","Purple"))
favourite_colour
class(favourite_colour)
```

`favourite_colour` is a tibble and is converted into as data.frame with:

```{r}
favourite_colour_df <- data.frame(favourite_colour)
favourite_colour_df
```

You can see that the output layout has changed, it does not report the dimensions and variable types, and you can check the type which is a base `data.frame` without `tbl_df` and `tbl`: 

```{r}
class(favourite_colour_df)
```
 
To do the reverse, from `data.frame` to `tibble`, use the function `as_tibble()`:

```{r}
favourite_colour_tibble <- as_tibble(favourite_colour_df)
favourite_colour_tibble
class(favourite_colour_tibble) 
```


```{r, echo=FALSE}
# ## Other features
# - column name repair: column names are not changed, for example in case of duplicate names
# - layout of data frames with large number of columns 
```


```{r, echo=FALSE}
## CAREMA data set (?)
```
