---
title: 'Mutate'
---

```{r setup,include=FALSE,eval=TRUE}
# library(rooc)
# source("../R/Doc.R")
# source("../R/Globals.R")
pulse <- read_pulse() 
survey <- read_survey()
```

> Use `mutate` to add new variables or modify the existing ones. 

## Add new variables

For example, the pulse dataset has two pulse measurements, let's say we are interested in *average pulse* and we want this information to be available as a separate variable, e.g. `averagePulse`, in the pulse tibble. Then we can do this with:    

```{r}
mutate(pulse, averagePulse = (pulse1+pulse2)/2)
```

By default the new column is added at the last position in the tibble.  

> Does the pulse tibble now contain the variable `averagePulse` ?

```{r, echo=FALSE, results='asis'}
msg <- "No, if you want to keep the new variable `averagePulse` you'll need to use assignment '<-' to replace the original pulse tibble with the newly modified version: <br><br>
pulse <- mutate(pulse, averagePulse = (pulse1+pulse2)/2)
<br>
"
qa(msg)
``` 


Take as another example the variable BMI: $$BMI=\frac{weight_{kg}}{{height_m}^2}$$

Note that BMI definition states that `weight` and `height` must be in *kilograms* and *metres* respectively. In the pulse dataset `weight` is given in kilograms but `height` is in centimetres. We can now first create a new variable `height_metre` containing the height in metres and then calculate BMI: 


```{r}
pulse_bmi <- mutate(pulse, height_metre=height/100) # convert centimetres to metre
pulse_bmi
```

`pulse_bmi` tibble has now the height in metre units, now we can calculate BMI:

```{r}
pulse_bmi <- mutate(pulse_bmi, BMI=weight/(height_metre^2)) 
pulse_bmi
```

Alternatively, you may skip the creation of `height_metre` and calculate BMI directly from the pulse tibble:

```{r}
pulse_bmi <- mutate(pulse, BMI=weight/((height/100)^2)) 
pulse_bmi
```

## Update variables

In the examples above we added a new variable to our dataset, but you can also update an existing variable. For example, let's say we want to have the age expressed (roughly) in days instead of years: 

```{r}
mutate(pulse, age=age*365)
```

here we keep the variable `age` but change its unit from year to days. 

Another example would be to convert the `height` and `weight` from metric to imperial units with (1 kg = 2.2 lbs) and (1 inch = 2.54 cm) : 

```{r}
mutate(pulse, height=height/2.54, weight=weight/2.2)
```

## if_else(condition, true, false, ...)

In the previous examples we were updating or adding variables with simple arithmetic using mutate and all values were considered under the same calculation. However, there are situation where we would like to treat values conditionally, this is possible with the helper function `if_else`. 

**Examples:**

Add a new variable `max_pulse` reporting the higher pulse rate of the two measurements `pulse1` and `pulse2` for each observation: 

```{r}
mutate(pulse, max_pulse=if_else(pulse1 < pulse2, pulse2, pulse1))
```

Add a logical variable `adult` which is true if age>=18 and false otherwise:

```{r}
mutate(pulse, adult=if_else(age >= 18 , TRUE, FALSE))
```


Convert `gender` values,  `female` to `f` and `male` to `m`:

```{r}
mutate(pulse, adult=if_else(gender=='female' , 'f', 'm'))

```










