---
title: 'Pipe operator %>%'
---

```{r setup,include=FALSE,eval=TRUE}
# library(rooc)
# source("../R/Doc.R")
# source("../R/Globals.R")
pulse <- read_pulse() 
survey <- read_survey()
```

\newcommand{\pipe}{{\small \;\;{\%}{>}{\%}\;\;} }

> Use $\pipe$ to chain calculations. 

The $\pipe$ operator allows you to run your function calls in sequence from left to right instead of *nested function calls* or using intermediate variables.


<!-- $${\Large f(g(x)) \;\;\;\; \Leftrightarrow  \;\;\;\; x \pipe g \pipe f}$$ 
Here $f$ and $g$ are functions and $x$ is a data object, possibly a vector but in majority of cases a tibble. The left hand side of the equivalence $f(g(x))$ is the standard in R, it calls first the $g(x)$ and then the result of $g(x)$ is taken as input to the function $f$. The right hand side, $x \pipe g \pipe f$,  will lead to the same result as $f(g(x))$.   -->

For example recall exercise 4  in `filter` practice section on `survey` data:  

- *How many males do smoke and never exercise?*  

**Solution:**

```{r}
nrow( filter(survey, smokes!="never" & exercise=="none" & gender=="male") )
```

Here, **first** function `filter` is run on the tibble `survey` with some conditions resulting into a tibble with rows fulfilling those conditions, **second** the function `nrow` is called on the resulting tibble, hence our answer `r nrow( filter(survey, smokes!="never" & exercise=="none" & gender=="male") )`. 

`r info_block("Use ctrl-M to get $\\pipe$ symbol.")`

**Solution with %>% : **

```{r}
survey %>% filter(smokes!="never" & exercise=="none" & gender=="male") %>% nrow()
```

The solution now reads from left to right $survey \pipe filter(...) \pipe nrow$ instead of $nrow(filter(survey,...))$.  

**Observation** In the $\pipe$ example above, note that `filter` does not have the `survey` as its first argument, the same is true for `nrow`. This is basically what the $\pipe$ operator does, take the result of the left hand side and place it as the first argument of the function at its right hand side.  

<!-- 
`r info_block("The main advantage of $\\pipe$ syntax is that it enables you to concentrate on the steps in the order they happen rather than the mathematical notation, e.g. $f(g(h(i(...))))$")`  -->

**Examples:**  

From the pulse dataset produce a tibble of personal information (name, age and gender): 

1. of individuals taller than 190 cm.

```{r}
pulse %>% filter(height>190) %>% 
  select(name,age,gender)
```

2. of individuals with weights between 40 and 50 kg.

```{r}
pulse %>% filter(weight>40 & weight<50) %>% 
  select(name,age,gender)
```

3. of individuals that neither smoke nor drink.

```{r}
pulse %>% filter(smokes=="no" & alcohol=="no") %>% 
  select(name,age,gender)

```


**Complex example:** We are asked to list only the names of the females in the pulse dataset with average pulse $>110$. We can break down the problem into:

  1. take only female observations (filter)
  2. add a new variable `averagePulse` (mutate)
  3. take only those average pulses above 110 (filter)
  4. extract names 

i.e.: 

```{r}
pulse %>%  filter(gender=="female")  %>%               # tibble females 
           mutate(averagePulse=(pulse1+pulse2)/2) %>%  # tibble females with averagePulse
           filter(averagePulse>110) %>%                # tibble females with averagePulse>100
           pull(name)                                  # vector of names
```

`r show_warning(" When in R Markdown file, each line in a sequence of commands with pipe must always end with the $\\pipe$ symbol except the last line, otherwise it is an error. See above example. Note also that each line may contain multiple $\\pipe$ symbols, e.g. first line in the example above, as long as it ends with a $\\pipe$ before the command on the next line. This way R know that there will be more commands on the next line.")`


