---
title: "Group (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

On the survey dataset: 

1. Report mean hand spans `{span1,span2}` per gender.

```{r}
survey %>% group_by(gender) %>% 
           summarise(meanWritingHand=mean(span1),meanNonWritingHand=mean(span2), .groups='drop')
```

2. Report mean and median `age` per gender, including the size of each gender group.

```{r}
survey %>% group_by(gender) %>% 
           summarise(size=n(), meanAge=mean(age), medianAge=median(age), .groups='drop')

```

3. Report mean `pulse` per `gender` and `exercise` group, including size of each group. Do the same for `gender` and `smokes` group.

```{r}
# gender,exercise group
survey %>% group_by(gender, exercise) %>% 
           summarise(size=n(), meanPulse=mean(pulse, na.rm=TRUE), .groups='drop')

# gender,smokes group
survey %>% group_by(gender, smokes) %>% 
           summarise(size=n(), meanPulse=mean(pulse, na.rm=TRUE), .groups='drop')

```

4. Report the mean `age` of those who exercise frequently per `gender` including the group size 

```{r}
# Solution 1:
survey %>% filter(exercise=='freq') %>% 
           group_by(gender) %>% 
           summarise(size=n(), meanAgeFreqExercise=mean(age), .groups='drop')

# Solution 2: first group and summarise and then filter out exercise group 'freq'. This is a 
# less optimal solution since it calculates the summary on all exercise groups, whereas the 
# solution 1 calculates the summary on a smaller table with only 'freq' exercise group.
# 
# survey %>% group_by(gender,exercise) %>%
#            summarise(size=n(), meanAgeFreqExercise=mean(age), .groups='drop') %>%
#            filter(exercise=='freq') %>%
#            select(-exercise)

```

5. Recall primary exercise (7) section `Summarise`, do the same using `group_by`.

```{r}
# Note: that additional argument '.groups = "drop"' is added to suppress tidyverse message.
 survey %>%  
  group_by(gender,smokes) %>% 
  summarise(meidan_age=median(age), .groups = "drop")  %>% 
  filter(smokes=='heavy')
```



## Extra exercises

1. Produce the following results:

1.a frequency of `hand` categories per gender.

```{r eval=TRUE}
survey %>%  group_by(gender,hand) %>%  summarise(n=n())
```

1.b frequency of `fold` categories of females.

```{r eval=TRUE}
survey %>%  filter(gender=="female") %>% group_by(gender,fold) %>% summarise(n=n(), .groups = "drop")

# Alternative solution to b) however less efficient, since the grouping is now done on a 
# larger set including male observations. 
# 
# survey %>%   group_by(gender,fold)  %>% summarise(n=n(), .groups = "drop") %>% filter(gender=="female")
```

1.c most popular name initials in descending order of popularity. See `?stringr::str_sub` for extracting initials from a character vector.   

```{r eval=TRUE}
survey %>% mutate(initial=str_sub(name, 1,1)) %>% group_by(initial) %>% summarise(n=n()) %>% arrange(desc(n))
```

