---
title: 'Tibble'
---

```{r setup,include=FALSE,eval=TRUE}
# library(rooc)
# source("../R/Doc.R")
# source("../R/Globals.R")
pulse <- read_pulse()
survey <- read_survey()

```

> A **tibble** is a table -- a two dimensional data structure with rows (**observations**) and columns (**variables**). 

I'll use the terms *observations* and *rows* interchangeably depending on the context. The same goes for the terms *variables* and *columns*. As you may recall the datasets `pulse` and `survey` were in fact of type tibble. Each variable in a tibble has a fixed [type](S02L01l_basic_vectors0.html) such as `character`, `numeric` etc. Let's start by creating a tibble manually.

## Create a tibble

To create a tibble you need to make sure that the package [tidyverse](https://www.tidyverse.org/) is installed and loaded. See [installation](S01L01l_index.html) for more details.  

Enter the following to load *tidyverse* package:

```{r}
library(tidyverse) 
```

Creating a __tibble__ is done using the keyword _tibble_ taking a sequence of `name=value` pairs where:

- `name` is the variable (column) name,
- `value` is the values of the observation.  


Take for example the variables `name`, `year` and `colour` to represent a person's name, birth year and favourite colour:   

```{r, echo=FALSE}
# favourite_colour  <- tibble(name=c("Lucas","Lotte","Noa","Wim"), 
#                            year=c(1995,1995,1995,1994), 
#                            colour=c("Blue","Green","Yellow","Purple"))
```


```{r, echo=TRUE}
favourite_colour  <- tibble(name=c("Lucas","Lotte","Noa","Wim","Marc","Lucy","Pedro"), 
                           year=c(1995,1995,1995,1994,1990,1993,1992), 
                           colour=c("Blue","Green","Yellow","Purple","Green","red","Blue"))

```

`r info_block("When creating a tibble the column vectors must be of the same length.")`

The variable `favourite_colour` now holds the data. Enter its name in the R `Console` for inspection:

```{r}
favourite_colour
```

> What additional pieces of information do you see beside the content we provided?

```{r, echo=FALSE, results='asis'}
msg <- "1) '# A tibble: 4 x 3', which says that this is a tibble with dimensions 4x3 (4 observations 
and 3 variables), \n\n 2) the atomic type of each variable, in this case character and double (numeric), \n\n 3) the row numbers"
qa(msg)
``` 

## Inspect your data

Type the following to find out the dimensions of the tibble:

```{r}
ncol(favourite_colour)  # number of variables (columns)
nrow(favourite_colour)  # number of observations (rows)
dim(favourite_colour)   # dimensions : 7 rows and 3 columns 
```

## Head and tail

Show top and bottom rows of the tibble: 

```{r}
head(favourite_colour, 2)  # first 2 observations (rows)
tail(favourite_colour, 3)  # last 3 observations (rows)
```

With the second argument to `head` and `tail` functions you can control the number of rows.

`r info_block("By default head and tail show 6 rows, i.e. when the second argument is omitted : head(favourite_colour) or tail(favourite_colour).")`


## `glimpse`

With `glimpse` function you can quickly inspect the top part of all variables in a tibble with some additional meta information such as number of rows and variables and their types: 

```{r, echo=FALSE}
options_width <- options()$width
options(width=80)
```

```{r}
favourite_colour %>% glimpse()
```

```{r, echo=FALSE}
options(width=options_width)
```

## Select variables: [

Often you may need to select certain variables, this can be done using square brackets `[` :

```{r}
favourite_colour["colour"]
```

or combination of variables:

```{r}
favourite_colour[c("name","year")]
```
 
`r info_block("Subset result of a tibble is always a tibble.")`


Selection of variables can also be achieved with indices as we saw in vectors:

```{r}
favourite_colour[2:3]
favourite_colour[c(1,3)]
```

To deselect use negative indices:

```{r}
favourite_colour[-2]
```


## Extract variables as vectors:  [[ or $

If you want to work with variables as individual vectors then you can do this either by double square brackets or $ sign: 

```{r}
favourite_colour[["year"]]
favourite_colour$year
```

In some contexts (later in the course) it is convenient to use the function `pull` which does the same as `[[` and `$` :  

```{r}
pull(favourite_colour, year) 
```


## tibble to/from file

Tibbles can be written to data files and read back again. Many data formats exist but for brevity we will be using
comma-separated-value (csv) format in this course. The functions involved for this purpose are `write_csv` and `read_csv` (see [data import cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf)).

<!-- For more details on reading/writing format see [readr_io0](readr_io0.html). --> 

let us now save our first tibble into a file in csv format:   

```{r}
write_csv(x = favourite_colour, file = "favourite_colour.csv")
```

`favourite_colour` tibble is written to `favourite_colour.csv` text file. You may inspect the file with any editor and it should look something like: 

```
name,year,colour
Lucas,1995,Blue
Lotte,1995,Green
Noa,1995,Yellow
Wim,1994,Purple
```

This way we can permanently store our results in files for later use. We can now read the csv file back into a R 
environment variable, e.g. `favourite_colour_csv` :

```{r}
favourite_colour_csv <- read_csv(file = "favourite_colour.csv")
favourite_colour_csv
```

The message from `read_csv` gives a summary of the variables and their inferred types, it also suggest to use argument `show_col_types = FALSE` if you'd prefere asilent read:

```{r}
favourite_colour_csv <- read_csv(file = "favourite_colour.csv", show_col_types = FALSE)
favourite_colour_csv
```

## Type of a tibble

Now inspect the type of the tibble we just created:

```{r}
class(favourite_colour)
```


A tibble is in its core a 'data.frame', a base R data structure. 

`r info_block("'The types tbl_df' and 'tbl' enforce additional convinient behaviours specific to tibble.")`



