---
title: "Summarise (practice)"
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
survey <- read_survey()
```

## Primary exercises

In the survey data:  

1. Take a look at the table with `glimpse` and inspect its dimensions and types.

```{r}
survey %>% glimpse()
```

2. Summarise on mean hand spans `{span1,span2}` and  median `pulse`.

```{r}
# 'pulse' variable has missing values (NA), therefore the use 
# of 'na.rm=TRUE' in the function 'median'.
#
survey %>% summarise(meanWritingHand=mean(span1), 
                     meanNonWritingHand=mean(span2), 
                     medianPulse=median(pulse, na.rm=TRUE))
```

3. Summarise on mean `age` and median `height`.

```{r}
survey %>% summarise(meanAge=mean(age), medianHeight=median(height, na.rm=TRUE))
```

4. Count the number of males and females.

```{r}
survey %>% count(gender)
```

5. Produce the frequency tables on {`gender`, `fold`} and  {`gender`, `fold`, `clap`}.

```{r}
survey %>% count(gender, fold)
survey %>% count(gender, fold, clap)

```

6. Produce the frequency table on `gender` and `smokes`,  show only females. 

```{r}
survey %>% count(gender,smokes) %>% filter(gender=="female") 
```

7. Summarise median age of male heavy smokers. Do the same for females.

```{r}
# male
survey %>%  filter(gender=='male' & smokes=='heavy') %>%  summarise(median_age=median(age))
# female
survey %>%  filter(gender=='female' & smokes=='heavy') %>%  summarise(median_age=median(age))
```



## Extra exercises


1. Produce the frequency table on `gender`, `exercise` and `smokes` but only for male smokers.

```{r}
# smoker is someone that smokes 'occas', 'regul' or 'heavy', i.e. smokes !='never'
#
survey %>% filter(gender=="male" & smokes!="never") %>% count(gender,exercise, smokes)
```

2. Do individuals all have different names? Put differently, is each observation uniquely identifiable with the variable `name`?

```{r}
# The answer is no, because the number of unique names is less than 
# total number of observations:
#
nrow( survey %>% distinct(name)) == nrow(survey)
```

3. Find pairs of variables which can uniquely identify each observation. Use `name` and one other variable for this purpose.

```{r}
# We see here that name and age combination is unique in the entire table.
nrow( survey %>% distinct(name, age)) == nrow(survey)
nrow( survey %>% distinct(name, height)) == nrow(survey) # The same is true for name and height
```


4. An alternative solution for the primary exercise (6) is solution (2) below:


```{r, echo=TRUE}
# solution (1)
survey %>% count(gender,smokes) %>% filter(gender=="female") 
# solution (2)
survey %>% filter(gender=="female") %>% count(gender,smokes)
```

where first `filter` is applied and then `count`. You may check the results and see that the outcomes are identical.
Explain the difference between the two solutions.

```{r}
# The alternative, solution (2), is more efficient, because 'count(gender, smokes)' has 
# less rows (only 'females') to work with, whereas  solution (1) counts both males and 
# females upon which the males are filtered out with filter(...).
```


