---
title: 'Group'
---

```{r setup,include=FALSE,eval=TRUE}
# library(rooc)
# source("../R/Doc.R")
# source("../R/Globals.R")
pulse <- read_pulse() 
survey <- read_survey()
```

> Group observations into sets and summarise on those sets.


## `group_by` : group observations

You can group your data given a set of variables. For example we can group per `age` and `gender`: 

```{r}
pulse %>% group_by( age,gender )
```

Note the tag **Groups: age, gender [21]** in the output, meaning that `group_by` found 21 `{age,gender}` groups in
our dataset. 

Once groups are marked with `group_by`, then we can analyse those groups with `summarise` producing a single row output per group. For example we can count the number of observation per group using function `n()`: 

```{r}
pulse %>% group_by( age, gender ) %>% 
          summarise( n = n())
```

`r show_warning("You may get a warning like 'summarise() regrouping ... (override with .groups argument)'. It is a reminder to remove the group from the result set. You may ignore this or set the 'summarise' argument .groups='drop', seel also ?summarise.") `

```{r}
pulse %>% group_by( age, gender ) %>% 
          summarise( n = n(), .groups='drop')
```

> Which function produces the same output as above given {age,gender}?

```{r, echo=FALSE, results='asis'}
msg <- "'count' function: <br><br>
puls %>% count(age,gnder)
<br><br>"
qa(msg)
```

But of course we want to do more than just count the size of the groups. We can for example calculate the mean `height` and `weight` per `{age,gender}` group: 

```{r}
pulse %>% group_by(age,gender) %>% 
          summarise(size=n(),meanHeight=mean(height), meanWeight=mean(weight))
```


<!-- ======================================================= -->

## Queries

We now have all the tools we need to apply more complex queries on our data. For example, group per `gender` on those that `ran` and summarize on mean `age`, `pulse1` and `pulse2`. First we need to filter only those who  *ran* (see explanation on `ran` in [pulse](data.html)) and only then group and summarise:  

```{r}
pulse %>% filter(ran == "ran") %>% 
          group_by( gender ) %>% 
          summarise( size = n(), meanAge= mean(age), meanPluse1 = mean(pulse1), meanPulse2 = mean( pulse2 ) )
```

now for those who `sat`:

```{r}
pulse %>% filter(ran == "sat") %>% group_by( gender ) %>% 
    summarise( size = n(), meanAge= mean(age), meanPluse1 = mean(pulse1), meanPulse2 = mean( pulse2 ) )
```

Note that there are missing values, account for it by using `na.rm=TRUE`:  

```{r}
pulse %>% filter(ran == "sat") %>% 
          group_by( gender ) %>% 
          summarise( count = n(), meanAge= mean(age), 
               meanPluse1 = mean(pulse1,na.rm=TRUE), meanPulse2 = mean( pulse2 , na.rm=TRUE) )
```

`r show_warning(" It is good practice to check your results. For example, the group sizes in the original survey table who ran and sat must match the sum of sizes in the different summaries shown above under 'size' column. For ran==\"ran\" summary we have 22+24=44 and for ran==\"sat\" summary we have 29+35=64. We can check them against totals below and we see that they do:")`

```{r}
pulse %>%  count(ran) 
```

## `ungroup` : remove grouping

To remove the grouping use we have `ungroup` function:

```{r}
pulse %>% group_by(age,gender) %>% ungroup() # results in the original pulse tibble
```

<!-- ## Useful functions
```{r}
pulse %>% group_by(age,weight) %>% group_size()  # group sizes (vector)
pulse %>% group_by(age,weight) %>% n_groups()    # number of groups 
``` 
-->

