---
title: "ggplot2 introduction (practice)"
---

```{r , include=FALSE,eval=TRUE}
pulse <- read_pulse()
survey <- read_survey()
```

## Primary exercises


1. A function with `if`/`else` statements.

Let's take the following from a World Health Organization document:

  - *An `adult` is a person older than 19 years of age unless national law defines a person as being an adult at an earlier age.*
  - *An `adolescent` is a person aged 10 to 19 years inclusive.*
  - *A `child` is a person 19 years or younger unless national law defines a person to be an adult at an earlier age.*
  - *However, in these guidelines when a person falls into the 10 to 19 age category they are referred to as an adolescent (see adolescent definition).*
  - *An `infant` is a child younger than one year of age.*

The categories can be enumerated as follows:

| group | condition for age [y] |
|-----|----------|
| `adult` | `age > 19` |
| `adolescent` | `age >= 10 && age <= 19` |
| `child` | `age >= 1 && age < 10` |
| `infant` | `age < 1` |



Implement a function  `ageGroup` which takes age argument (in years) and returns the age group label.

**Function template** Here is a template of the `ageGroup` function:

```{r, eval=FALSE, echo=TRUE}
ageGroup <- function( age ) { 
  # age: age given in years
  ...
}
```

and here are expected results:

```{r, eval=FALSE, echo=TRUE}
ageGroup(0)   # "infant"
ageGroup(10)  # "adolescent"
ageGroup(9)   # "child"
ageGroup(20)  # "adult"
ageGroup(-1)  # should generate an error
```

To generate the error we request the program to `?stop` with an error message.

```{r}
# 1) Poor implementation
#
ageGroup <- function( age ) { 
  # age: age given in years
  result <- "child"
  if( age > 19 ) {
    result <- "adult"
  }
  if( age >= 10 & age <= 19 ) {
    result <- "adolescent"
  } 
  if( age < 1 ) {
    result <- "infant"
  } 
  if( age < 0 ) {
    stop( "Age can't be negative." )
  } 
  result
}
```


```{r}
# 2) Improved implementation 
# 
# In the previous scenario: when `age` is 30 and the `result` is set to `adult` 
# it makes no sense to check whether `age` is in other ranges,  
# `if`/`else` construction allows for that.
#
ageGroup <- function( age ) { 
  # age: age given in years
  if( age > 19 ) {
    result <- "adult"
  } else if( age >= 10 ) {
    result <- "adolescent"
  } else if( age >= 1 ) {
    result <- "child"
  } else if( age >= 0 ) {
    result <- "infant"
  } else {
    stop( "Age can't be negative." )
  }
  result
}
```

```{r, collapse=TRUE,error=TRUE}
ageGroup(0)   # "infant"
ageGroup(10)  # "adolescent"
ageGroup(9)   # "child"
ageGroup(20)  # "adult"
ageGroup(-1)  # should generate an error
```


