---
title: "ggplot2 scales (practice)"
---

```{r , include=FALSE, eval=TRUE}
pulse <- read_pulse()
```

## Primary exercises

1. *Logarithmic scale and manual colors*.  
For the `pulse` data construct the plot shown below.  
Add `scale_x_log10` and `scale_y_log10` to change both location scales to logarithmic.  
Enforce the same `limits` from 50 to 150 on both scales.
```{r warning=FALSE,eval=TRUE}
ggplot( pulse ) +
  aes( x = pulse1, y = pulse2, color = ran ) +
  geom_point() +
  scale_x_log10( limits = c( 50, 150 ) ) + 
  scale_y_log10( limits = c( 50, 150 ) ) +
  xlab( "pulse1 (before exercise) [1/min]" ) +
  ylab( "pulse2 (after exercise) [1/min]" ) +
  scale_color_manual( values = c( ran = "black", sat = "orange" ) )
```

## Extra exercises

1. *Advanced scales*.  
The [`scales` package](https://scales.r-lib.org/reference/index.html) extends `ggplot2` functionality by providing *"tools to override the default breaks, labels, transformations and palettes"*.  
Have a look at the package reference manual to get familiar with types of scales it may provide for your projects.
