---
title: 'ggplot2 geoms'
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
```

Generally speaking, `ggplot2` geoms specify **plot types**.  
Each **geom** produces a plot **layer** and multiple layers can be combined.  
Here we demonstrate several frequently used geoms.  
Try to regenerate the plots in your R Markdown document.

`r go_extern( "https://rstudio.com/resources/cheatsheets/", "RStudio Cheatsheets/Data Visualization Cheatsheet/Panel Scales", "to see numerous geoms provided by the library" )`

`r go_extern( "https://www.r-graph-gallery.com/", "The R Graph Gallery", "to see how R (often with ggplot2 library) can be used for data visualisation" )`

## Histograms

Let's start with the histogram of the `pulse2` variable from the `pulse` data:
```{r}
ggplot( pulse ) +
  aes( x = pulse2 ) +
  geom_histogram( color = "black", fill = "gray" )
```

Note, that there were two groups of subjects: one did `run`, the other did not.  
Try to add color to split histogram bars to make groups visible:
```{r}
ggplot( pulse ) +
  aes( x = pulse2, fill = ran ) +
  geom_histogram( color = "black" )
```

The above histogram has the **groups stacked**.  
To visualize each group separately, try to add `position` argument as below (remember to add `alpha`; otherwise some bars might get hidden):
```{r}
ggplot( pulse ) +
  aes( x = pulse2, fill = ran ) +
  geom_histogram( color = "black", position = "identity", alpha = 0.6 )
```

An another possible value of the `position` argument:
```{r}
ggplot( pulse ) +
  aes( x = pulse2, fill = ran ) +
  geom_histogram( color = "black", position = "dodge" )
```
