---
title: 'ggplot2 Plot Panels/Size and Plot Types'
---

```{r setup,include=FALSE,eval=TRUE}
pulse <- read_pulse() 
```

`ggplot2` facets allow to split a plot into panels depending on one (or more) categorical variables.  
`ggplot2` themes configure plot graphical settings: colors, fonts, ...

`r go_extern( "https://rstudio.com/resources/cheatsheets/", "RStudio Cheatsheets/Data Visualization Cheatsheet/Panel Scales", "to check commands for facets and themes" )`


## Storing plot in a variable

`ggplot` is a function and it returns an **object (of class `ggplot`)** representing a plot.  
This object can be stored in a variable.  
Notice, that no plot is shown by the following code:
```{r}
p <- ggplot( pulse ) +
  aes( x = weight, y = height, color = exercise, shape = gender ) +
  geom_point( size = 3, alpha = 0.8 )
```

Only once the variable is printed the plot is shown.
```{r}
p
```

## Saving plots

Such a plot object may also be **saved to a file** with the `ggsave` function.  
In `Help` you may find how to specify width, height and dpi resolution of the image.  
Multiple file formats are supported and by default they are detrmined from the filename.  
For example, to save plot `p` in PNG format:
```{r eval=FALSE}
ggsave( "my_plot.png", plot = p )
```

## Plot size in R Markdown

It is possible to control the dimensions of the plots in your R Markdown report.  
Try to add additional options to the first line of a chunk which produces a plot.  
Follow this example (the dimensions `fig.width` and `fig.height` are specified in inches and `dpi` sets resolution in pixels per inch): 

````{r,result="as.is",echo=FALSE}
cat( "```{r fig.width=3,fig.height=2,dpi=75}\np\n```\n" )
````
```{r fig.width=3,fig.height=2,dpi=75,echo=FALSE}
p
```

## Themes

`ggplot2` allows detailed configuration of plots, far beyond the scope of this course.  
We advise to use google search with phrases like ["ggplot2 rotate axis labels"](https://www.google.com/search?q=ggplot+rotate+axis+labels).

Let's again use the plot stored in the variable `p` and combine it with general

**themes**.  
Try each of the following lines and observe the effects:
```{r eval=FALSE}
p + theme_minimal()
p + theme_dark()
p + theme_bw()
```


**Plot Types**

Generally speaking, `ggplot2` geoms specify **plot types**.  
Each **geom** produces a plot **layer** and multiple layers can be combined.  
Here we demonstrate several frequently used geoms.  
Try to regenerate the plots in your R Markdown document.

`r go_extern( "https://rstudio.com/resources/cheatsheets/", "RStudio Cheatsheets/Data Visualization Cheatsheet/Panel Scales", "to see numerous geoms provided by the library" )`

`r go_extern( "https://www.r-graph-gallery.com/", "The R Graph Gallery", "to see how R (often with ggplot2 library) can be used for data visualisation" )`

## Histograms

Let's start with the histogram of the `pulse2` variable from the `pulse` data:
```{r}
ggplot( pulse ) +
  aes( x = pulse2 ) +
  geom_histogram( color = "black", fill = "gray" )
```

Note, that there were two groups of subjects: one did `run`, the other did not.  
Try to add color to split histogram bars to make groups visible:
```{r}
ggplot( pulse ) +
  aes( x = pulse2, fill = ran ) +
  geom_histogram( color = "black" )
```

The above histogram has the **groups stacked**.  
To visualize each group separately, try to add `position` argument as below (remember to add `alpha`; otherwise some bars might get hidden):
```{r}
ggplot( pulse ) +
  aes( x = pulse2, fill = ran ) +
  geom_histogram( color = "black", position = "identity", alpha = 0.6 )
```


An another possible value of the `position` argument:
```{r}
ggplot( pulse ) +
  aes( x = pulse2, fill = ran ) +
  geom_histogram( color = "black", position = "dodge" )
```
