---
title: 'tidyverse'
---

## Package tidyverse

[Tidverse](https://www.tidyverse.org/) is a collection of R packages which we will be using intensively for the remainder of this course. If you haven't installed `tidyverse` then you can do this with:

```{r, eval=FALSE}
install.packages("tidyverse") # needed only once
```

and load it into R: 

```{r, eval=FALSE}
library(tidyverse) # each time you start RStudio
```

Something similar to the following will appear on your screen after loading tidyverse:

```
> library(tidyverse)
── Attaching packages ────────────────────────────────────────────────────────────────────── tidyverse 1.3.1 ──
✓ ggplot2 3.3.3     ✓ purrr   0.3.4
✓ tibble  3.1.2     ✓ dplyr   1.0.6
✓ tidyr   1.1.3     ✓ stringr 1.4.0
✓ readr   1.4.0     ✓ forcats 0.5.1
── Conflicts ───────────────────────────────────────────────────────────────────────── tidyverse_conflicts() ──
x dplyr::filter() masks stats::filter()
x dplyr::lag()    masks stats::lag()
> 
```

It shows the sub-packages and the versions and also some conflicts. 

`r show_warning("You may ignore these Conflicts.")`

You now have access to an extensive set of functions for **data manipulation and visualisation**. 

## Outline

- `dplyr`: Data transformation ([cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf){target="_blank"})
  - tibble : a two-dimensional data structure with rows and columns
  - functions to manipulate tibble:
    - select : extract columns
    - filter : extract rows
    - mutate : add a new column
  - pipe operator `%>%` 
  - summaries and sorting 
  - grouping
  - queries 
  - join : combining tables
- `tidyr`: Change the table layout ([cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/tidyr.pdf){target="_blank"})  
  - pivot_longer : wide to long format
  - pivot_wider : long to wide format
- `forcats`: Manipulation on factors and their levels. ([cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/factors.pdf){target="_blank"})  
  - fct_* functions  


## Links

- [Tidyverse](https://www.tidyverse.org/)
- [R for data science ; Hadley Wickham, Garrett Grolemund (online book)](https://r4ds.had.co.nz/)
- [More cheat sheets (*)](https://rstudio.com/resources/cheatsheets/)

