---
title: "Reshape (practice)"
#author: "R. Monajemi"
---

```{r , include=FALSE,eval=TRUE}
survey <- read_survey()
pulse <- read_pulse()
```

## Primary exercises

In the survey dataset:

1. The variables `span1` and `span2` are of the same type representing the span of the writing hand and the non-writing hand respectively. We would like to analyse these values in groups. To do this we need to convert the variables `span1` and `span2` into variables `span` and `size` with `span` ranging over *span1* and *span2* and `size` over the values of `span1` and `span2` variables.     
  
```{r}
survey %>% pivot_longer(c(span1,span2), names_to = "span", values_to = "size")
```

2. Reproduce the following plot:  

```{r, eval=TRUE}
ggplot(survey %>% pivot_longer(c(span1,span2), names_to = "span", values_to = "size")) + 
  aes(x=gender, y=size, color=span) + geom_boxplot()
```


## Extra exercises

1. Produce the following summary per gender from variables hand, fold and clap.

```{r, eval=TRUE}
survey %>% 
  select(gender,hand,fold,clap) %>% 
  pivot_longer(!gender, names_to = "action", values_to = "side") %>% 
  group_by(gender,side) %>% 
  summarise(count=n(), .groups='drop') %>% 
  pivot_wider(names_from = "side",values_from = "count")
```
  
2. Often reshaping of the variables are necessary to visualize the information with ggplot. From *pulse* dataset reproduce the  *boxplot* shown below. Here `level` is the values of  `pulse1` and `pulse2`.   

```{r, eval=TRUE}
ggplot(pulse %>% 
        drop_na() %>% 
        mutate(exercise=fct_relevel(exercise, c("low","moderate","high"))) %>%
        pivot_longer(c(pulse1,pulse2), names_to = "pulse",values_to = "level") ) + 
  aes(x=exercise, y=level, color=pulse) + geom_boxplot()
```







```{r, include=FALSE}
# 1. Summarise minimum and maximum `span` per `writing` from the primary exercise (1). 
survey %>% pivot_longer(c(span1,span2), names_to = "span", values_to = "size") %>% group_by() %>%  
          summarise(min=min(span,na.rm = T), max=max(span,na.rm = T))
```

<!-- 
1. Two other categorical variables in the pulse dataset which can be gathered are `smokes` and `alcohol`. Do this into new variables `drug_habit` and `drug_use`.
--> 
