---
title: 'tidyverse'
---


```{r setup, include=FALSE,eval=TRUE}
# library(rooc)
# source("../R/Doc.R")
# source("../R/Globals.R")
# pulse <- read_pulse() 

```

## Package tidyverse

[Tidverse](https://www.tidyverse.org/) is a collection of R packages which we will be using intensively for the remainder of this course. If you haven't installed `tidyverse` then you can do this with:

```{r, eval=FALSE}
install.packages("tidyverse") # needed only once
```

and load it into R: 

```{r, eval=FALSE}
library(tidyverse) # each time you start RStudio
```

Something similar to the following will appear on your screen after loading tidyverse:

```
> library(tidyverse)
── Attaching core tidyverse packages ─────────────────────────────────────────────────────── tidyverse 2.0.0 ──
✔ dplyr     1.1.2     ✔ readr     2.1.4
✔ forcats   1.0.0     ✔ stringr   1.5.0
✔ ggplot2   3.4.2     ✔ tibble    3.2.1
✔ lubridate 1.9.2     ✔ tidyr     1.3.0
✔ purrr     1.0.2     
── Conflicts ───────────────────────────────────────────────────────────────────────── tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
ℹ Use the conflicted package to force all conflicts to become errors
>
```

It shows the sub-packages and the versions and also some conflicts. 

`r show_warning("You may ignore these Conflicts.")`

You now have access to an extensive set of functions for **data manipulation and visualisation**. 

## Outline

- `dplyr`: Data transformation
  - tibble : a two-dimensional data structure with rows and columns
  - functions to manipulate tibble:
    - select : extract columns
    - filter : extract rows
    - mutate : add a new column
  - pipe operator `%>%` 
  - summaries and sorting 
  - grouping
  - queries 
  - join : combining tables
- `tidyr`: Change the table layout  
  - pivot_longer : wide to long format
  - pivot_wider : long to wide format

## Links

- [Tidyverse](https://www.tidyverse.org/)
- [R for data science (2e) ; by Hadley Wickham, Mine Çetinkaya-Rundel, Garrett Grolemund (online book)](https://r4ds.hadley.nz/)
- [More cheat sheets (*)](https://rstudio.com/resources/cheatsheets/)

