---
title: "Reshape (practice)"
#author: "R. Monajemi"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
pulse <- read_pulse()
```


```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

In the survey dataset:

1. The variables `span1` and `span2` are of the same type representing the span of the writing hand and the non-writing hand respectively. We would like to analyse these values in groups. To do this we need to convert the variables `span1` and `span2` into variables `span` and `size` with `span` ranging over *span1* and *span2* and `size` over the values of `span1` and `span2` variables.     
  
```{r}
survey %>% gather(key=span,value=size,span1, span2)
```

2. Reproduce the following plot:  

```{r, eval=TRUE}
ggplot(survey %>% gather(key=span,value=size,span1, span2)) + 
  aes(x=gender, y=size, color=span) + geom_boxplot()
```


## Extra exercises

1. Often reshaping of the variables are necessary to visualize the information with ggplot. From *pulse* dataset reproduce the  *boxplot* shown below. Here `level` is the values of  `pulse1` and `pulse2`.   

```{r, eval=TRUE}
ggplot(pulse %>% 
        drop_na() %>% 
        mutate(exercise=fct_relevel(exercise, c("low","moderate","high"))) %>%
        gather(pulse,level, pulse1,pulse2)) + 
  aes(x=exercise, y=level, color=pulse) + geom_boxplot()
```







```{r, include=FALSE}
# 1. Summarise minimum and maximum `span` per `writing` from the primary exercise (1). 
survey %>% gather(span,size, span1,span2)  %>% group_by() %>%  
          summarise(min=min(span,na.rm = T), max=max(span,na.rm = T))
```

<!-- 
1. Two other categorical variables in the pulse dataset which can be gathered are `smokes` and `alcohol`. Do this into new variables `drug_habit` and `drug_use`.
--> 