---
title: "Useful functions"
---

## Sandbox vectors

Let's define some vectors which can be used for demonstrations:
```{r}
manyNumbers <- sample( 1:1000, 20 )
manyNumbers
manyNumbersWithNA <- sample( c( NA, NA, NA, manyNumbers ) )
manyNumbersWithNA
duplicatedNumbers <- sample( 1:5, 10, replace = TRUE )
duplicatedNumbers
letters
LETTERS
mixedLetters <- c( sample( letters, 5 ), sample( LETTERS, 5 ) )
mixedLetters
```

## Are all/any elements TRUE

- *Input:* logical vector
- *Output:* single logical value
- *Task:* try, understand what happens when you use `manyNumbersWithNA` instead of `manyNumbers`.

```{r}
all( manyNumbers <= 1000 )
all( manyNumbers <= 500 )
any( manyNumbers > 1000 )
any( manyNumbers > 500 )
all( !is.na( manyNumbers ) )
any( is.na( manyNumbers ) )
```

## Which elements are TRUE

*Input:* logical vector
*Output:* vector of numbers (positions)

```{r}
which( manyNumbers > 900 )
which( manyNumbersWithNA > 900 )
which( is.na( manyNumbersWithNA ) )
```

## Filtering vector elements

- *Input:* any vector and filtering condition
- *Output:* elements of the input vector
- *Note:* several ways to get the same effect

```{r}
manyNumbers[ manyNumbers > 900 ] # indexing by logical vector
manyNumbers[ which( manyNumbers > 900 ) ] # indexing by positions

somePositions <- which( manyNumbers > 900 )
manyNumbers[ somePositions ]
```

## Are some elements among other elements

- *Input:* two vectors
- *Output:* a logical vector corresponding to the first input vector

```{r}
"A" %in% LETTERS
c( "X", "Y", "Z" ) %in% LETTERS
all( c( "X", "Y", "Z" ) %in% LETTERS )
all( mixedLetters %in% LETTERS )
any( mixedLetters %in% LETTERS )
mixedLetters[ mixedLetters %in% LETTERS ]
mixedLetters[ !( mixedLetters %in% LETTERS ) ]
```

```{r}
manyNumbers %in% 300:600
which( manyNumbers %in% 300:600 )
sum( manyNumbers %in% 300:600 )
```

## Pick one of two (three) depending on condition

- *Input:* a logical vector and two vectors additional vectors (for TRUE, for FALSE)
- *Output:* elements of the additional vectors
- *Note:* it can take care of `NA`s

```{r}
if_else( manyNumbersWithNA >= 500, "large", "small" )
if_else( manyNumbersWithNA >= 500, "large", "small", "UNKNOWN" )

# here integer 0L is needed instead of real 0.0 
# manyNumbersWithNA contains integer numbers and the method complains
if_else( manyNumbersWithNA >= 500, manyNumbersWithNA, 0L ) 
```

## Duplicates and unique elements

- *Input:* a vector

```{r}
unique( duplicatedNumbers )
unique( c( NA, duplicatedNumbers, NA ) )
duplicated( duplicatedNumbers )
```

## Positions of max/min elements

```{r}
which.max( manyNumbersWithNA )
manyNumbersWithNA[ which.max( manyNumbersWithNA ) ]
which.min( manyNumbersWithNA )
manyNumbersWithNA[ which.min( manyNumbersWithNA ) ]
range( manyNumbersWithNA, na.rm = TRUE )
```

## Sorting/ordering of vectors

```{r}
manyNumbersWithNA
sort( manyNumbersWithNA )
sort( manyNumbersWithNA, na.last = TRUE )
sort( manyNumbersWithNA, na.last = TRUE, decreasing = TRUE )
```

```{r}
manyNumbersWithNA[1:5]
order( manyNumbersWithNA[1:5] )
rank( manyNumbersWithNA[1:5] )
```

```{r}
sort( mixedLetters )
```

## Ranking of vectors

```{r}
manyDuplicates <- sample( 10:15, 10, replace = TRUE )
rank( manyDuplicates )
rank( manyDuplicates, ties.method = "min" )
rank( manyDuplicates, ties.method = "random" )
```

## Rounding numbers

```{r}
v <- c( -1, -0.5, 0, 0.5, 1, rnorm( 10 ) )
v
round( v, 0 )
round( v, 1 )
round( v, 2 )
floor( v )
ceiling( v )
```

## Naming vector elements

```{r}
heights <- c( Amy = 166, Eve = 170, Bob = 177 )
heights
names( heights )
names( heights ) <- c( "AMY", "EVE", "BOB" )
heights
heights[[ "EVE" ]]
```

## Generating grids

```{r}
expand_grid( x = c( 1:3, NA ), y = c( "a", "b" ) )
```

## Generating combinations

```{r}
combn( c( "a", "b", "c", "d", "e" ), m = 2, simplify = TRUE )
combn( c( "a", "b", "c", "d", "e" ), m = 3, simplify = TRUE )
```
