---
title: "Calculator (practice)"
---

## Primary exercises

1. *Simple expression.*  
Calculate 25 plus the square of 12.

```{r}
25 + 12^2
```

2. *Calling a function.*  
What is the square root of the number you have just calculated (here type the number directly)?

```{r}
sqrt( 25 + 12^2 )
```

3. *Prompt with `+`.*  
In the console type an incomplete expression `(1+2` and press Enter. Practice how to return to the proper console prompt `>`. Finally, clear the console.

```{r}
# To stop in the console press: Esc or Ctrl-C
# To clear in the console press: Ctrl-L
```

4. *Some error messages.*  
In the console type an error expression `(1+2)*(3+4))`. Understand the error message. Also understand errors reported for expressions: `0,25`, `sin()`, `xyz`.

```{r}
# (1+2)*(3+4))     -> forgotten (
# 0,25             -> use dot as decimal point: 0.25
# sin()            -> sin needs an argument
# xyz              -> xyz is not known
```

5. *Getting help.*  
Find the manual page for the `sin` function.

```{r}
# In the console type: ?sin
```

## Extra exercises

1. *Command history.*  
In the console try: `history()`. Locate the `History` pane and observe what happens when you double-click a line there.  
A save button allows to save typed commands but this is not a practical way to save the code.

```{r}
# history() in RStudio simply shows History pane
# double click copies a line from History to the Console
# we will learn to write scripts instead of saving History
```

2. *Logarithms and exponentials.*  
Check the manual to understand the difference between `log10( 100 )`, `log2( 100 )` and `log( 100 )`.  
Find what would be the result of `exp( 1 )`.

```{r}
log10( 100 )   # logarithm base 10 of 100, 10^2 = 100
log2( 100 )    # logarithm base 2 of 100
log( 100 )     # natural logarithm of 100
exp( 1 )       # base of natural logarithm, e^1
```

3. *Trigonometry.*  
Calculate $cos(90^{\circ})$.  
Remember that in R the trigonometric functions require angles specified in [radians](https://en.wikipedia.org/wiki/Radian).

```{r}
cos( 90 / 180 * pi ) # conversion from degrees to radians
                     # the result is very close to zero, but not exact zero
                     # the numeric calculation is approximate
```

