---
title: "Data sets"
---

## Data sets

Throughout the course we will use the two data sets described below.  
The data sets can be downloaded from the Materials page.

### Pulse 
  
Students in an introductory statistics class (MS212 taught by Professor John Eccleston and Dr Richard Wilson at The University of Queensland) participated in a simple experiment. The students measured their own pulse rate. They were then asked to flip a coin. If the coin came up heads, they were to run in place for one minute. Otherwise they sat without movement for one minute. Then everyone measured their pulse again. The pulse rates and other physiological and lifestyle data are given in the data table.

|Variable | Explanation |
|---|-----------------------------------------------------|
| name | Name of a participant |
| height	  |	Height (cm) |
| weight	  |	Weight (kg) |
| age	    |	Age (years) |
| gender	  |	Sex (male/female) |
| smokes	  |	Regular smoker? (yes/no) |
| alcohol	|	Regular drinker? (yes/no) |
| exercise	|	Frequency of exercise (high/moderate/low) |
| ran	    |	Whether the student ran or sat between the first and second pulse measurements (ran/sat) |
| pulse1  	|	First pulse measurement (rate per minute) |
| pulse2	  |	Second pulse measurement (rate per minute) |
| year		  | Year of the class (1993 - 1998) |

The pulse data set is available in the data folder as comma-delimited text (extension `.csv`) file `pulse.csv`.


```{r, message=FALSE,warning=FALSE,echo=TRUE}
# note, we keep 'pulse.csv' file in 'data' directory of the project
pulse <- read_csv("data/pulse.csv")
pulse 
```

 
### Survey

This data frame contains the responses of 233 Statistics I students at the University of Adelaide to a number of questions.  
It is a slightly modified version of the `survey` data from the `MASS` pacakge.

|Variable | Explanation |
|---|-----------------------------------------------------|
| name | Name of a participant  |
| gender	  |	Sex (male/female) |
| span1   | Span (distance from tip of thumb to tip of little finger of spread hand) of writing hand (cm) | 
| span2   | Span of non-writing hand (cm) | 
| hand  | Writing hand of student (left/right) |
| fold | Fold your arms! which is on top? (right/left/neither) |
| pulse  	|	Pulse measurement (rate per minute)|
| clap | Clap your hands! which is on top? (right/left/neither) |
| exercise	|	Frequency of exercise (freq/some/none)|
| smokes	  |	How much the student smokes (heavy/regul/occas/never)|
| height	  |	Height (cm) |
| m.i	  |	whether the student expressed height in imperial (feet/inches) or metric (centimetres/metres) units. (metric/imperial) |
| age	    |	Age of the student (years)|

The survey data set is available in the data folder as comma-delimited (`.csv`) text: `survey.csv`.

```{r, message=FALSE,warning=FALSE,echo=TRUE}
# note, we keep 'survey.csv' file in 'data' directory of the project
survey <- read_csv("data/survey.csv")
survey 
```
