---
title: "TODO notes"
output: html_document
---

# Tasks

 (selection: unique, which.max, match, sort, order, duplicated, rank, round, ceiling, complete.cases, expand.grid, complete, combn/sample, choose, median)
naming vector elements

## Extra exercises

3. *Sorting, reversing and sampling*.  
Check the effect of `sort` function on the vector of names `x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )`.  
Next, print the effect of `sample` function on `x`.  
Repeat the last step to see another permutation of `x`.  
Check `Help` of `sample` and understand the effect of `sample( x, size = 2 )`.  
Finally, understand effect of the `rev` function applied to the vecotr `x`.

```{r}
x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
sort( x )              # x sorted alphabetically
sample( x )            # a random permutation of x
sample( x )            # a random permutation of x
sample( x, size = 2 )  # randomly chosen 2 different elements of x
rev( x )               # x in reversed order
rev( sort( x ) )       # x sorted in reversed alphabetical order
sort( x, decreasing = TRUE )  # a faster version of above
```

4. *Counting elements, finding unique elements*.  
Let's consider the following vector: `x<-c( 1,3,2,3,5,5,8,NA,2,1,6,8,NA,3,3,5 )`.  
Read the `Help` for `unique`. Apply `unique` to `x` and study the result.  
Understand the output of `table(x)`.
```{r}
x <- c( 1,3,2,3,5,5,8,NA,2,1,6,8,NA,3,3,5 )
unique(x)    # all elements present in x with no duplications
table( x )   # number of occurrences of elements from x; no NA
```


2. *Sum of a sequence of numbers*.  
Make a vector of all numbers between 1 and 100.  
Check the type of that vector (note, that a vector of numbers might be different than `numeric`).  
Calculate 1+2+3+…+100.  
Hint: Read help (see `Help` pane) for function `sum`. 

```{r}
v <- 1:100
class( v )
sum( v )
```


3. *Temperature units*.  
Degrees Celsius (`Cel`) can be calculated from degrees Fahrenheit (`Fah`) as `Cel = 5/9 * (Fah - 32)`.  
How much is 60 degrees Fahrenheit in Celsius?  
Use `seq` function to create a vector that contains 0,5,10,...,95,100 degrees Fahrenheit.  
Convert this vector to Celsius degrees.  
From your calculation in the previous exercise, select 10 Fahrenheit in Celcius.

```{r}
Fah <- 60
Cel <- 5 / 9 * (Fah - 32)
Cel
Fah <- seq( 0, 100, 5 )
Cel <- 5 / 9 * (Fah - 32)
Cel
Fah[3]
Cel[3]
```


5. *Combining (appending) vectors*.  
How to merge a vector `1:10` with `c( 2314, 2178, 1922, 2004 )`?
```{r}
c( 1:10, c( 2314, 2178, 1922, 2004 ) )   # direct solution

x <- 1:10
y <- c( 2314, 2178, 1922, 2004 )
c( x, y )                                # indirect solution
```


1. Practice the integer division operator `%/%`. Compare `4 / 3` with `4 %/% 3`. 

```{r}
4 / 3     # floating division
4 %/% 3   # largest integer smaller/equal to the floating division result
```

2. Understand result of `-4 %/% 3`, compared to `4 %/% 3`.

```{r}
4 %/% 3   # largest integer smaller/equal to the floating division result
-4 %/% 3  # largest integer SMALLER/EQUAL to the floating division result
```

## Ramin's

- (Ramin) d %>% filter( is.na( a_column ) ) => DONE!
- (Ramin) pull, distinct, rank => DONE except rank. We decided that rank and order are advanced topics. 
- (Ramin) top_n
- (Ramin) where is na.rm explained for the first time? 
    - (Szymon says) It is not explained yet!

## Szymon's

- (Bart or extra exercise) table, prop.table (for logistic regression)
- (Szymon) saveRDS, readRDS
- (Szymon) exercises ggplot::geom :
  - barplot with errors
  - show `geom_density` over histogram
  - `geom_smooth`: geom to show fit
  - geom to show function
- (Szymon) exercises formulas

## undecided

- set/get names() of vector elements
- `letters` and `LETTERS`; `pi`
- how to load a vector from file
- any, all
- some sections, for example forcats, has a different icon for info than the info_block function
		=> make sure that all sections use the same type of icons

## done

- (Szymon) replace x+3 with x in is.na paragraph in basic vectors. it is distracting.
- (Szymon) add format to all exercises after R Markdown introduced
- (Bart or extra exercise) `table` added, `prop.table` skipped (for logistic regression)
- (Szymon) exercises ggplot::scales
- (Szymon) exercises formulas

