---
title: "Vectors (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```

## Primary exercises

1. *Kilocalories*.  
Four patients had daily dietary intakes of 2314, 2178, 1922, 2004 kcal.  
Make a vector of these four values.  
What is the type of this vector?  
Convert these values into in kJ using 1 kcal = 4.184 kJ.  

```{r}
intakesKCal <- c( 2314, 2178, 1922, 2004 )
intakesKCal
class( intakesKCal )
intakesKCal * 4.184
```

2. *Sum of a sequence of numbers*.  
Make a vector of all numbers between 1 and 100.  
Check the type of that vector (note, that a vector of numbers might be different than `numeric`).  
Calculate 1+2+3+…+100.  
Hint: Read help (see `Help` pane) for function `sum`. 

```{r}
v <- 1:100
class( v )
sum( v )
```

3. *Temperature units*.  
Degrees Celsius (`Cel`) can be calculated from degrees Fahrenheit (`Fah`) as `Cel = 5/9 * (Fah - 32)`.  
How much is 60 degrees Fahrenheit in Celsius?  
Use `seq` function to create a vector that contains 0,5,10,...,95,100 degrees Fahrenheit.  
Convert this vector to Celsius degrees.  
From your calculation in the previous exercise, select 10 Fahrenheit in Celcius.

```{r}
Fah <- 60
Cel <- 5 / 9 * (Fah - 32)
Cel
Fah <- seq( 0, 100, 5 )
Cel <- 5 / 9 * (Fah - 32)
Cel
Fah[3]
Cel[3]
```

4. *Descriptive statistics of a set of random numbers*.  
The code `v <- rnorm( 15 )` would store 15 random numbers in the variable `v`.  
Print `v` and look to find the smallest and the largest of these numbers.  
Try to use the functions `min` and `max` on `v` -- have you found the same numbers?  
Calculate the `mean`, `median` and the standard deviation (`sd`) of `v`.  
Count the number of elements smaller than zero.
```{r}
v <- rnorm( 15 )
min( v )
max( v )
mean( v )
median( v )
sd( v )
sum( v < 0 )
```

5. *Combining (appending) vectors*.  
How to merge a vector `1:10` with `c( 2314, 2178, 1922, 2004 )`?
```{r}
c( 1:10, c( 2314, 2178, 1922, 2004 ) )   # direct solution

x <- 1:10
y <- c( 2314, 2178, 1922, 2004 )
c( x, y )                                # indirect solution
```

## Extra exercises

1. *More complex sequences of numbers*.  
Read help (see `Help` pane) about `seq` function.  
Use it to generate sequence: 10, 7, 4, 1, -2, -5.  
Understand the error message of `seq( 10, -5, 3 )`.

```{r}
seq( 10, -5, -3 )
seq( from = 10, to = -5, by = -3 )
```

2. *Repetitions*.  
Read help (see `Help` pane) about `rep` function.  
Use it to generate sequence: 0, 1, 0, 1, 0, 1.  

```{r}
rep( c( 0, 1 ), 3 )
```

3. *Sorting, reversing and sampling*.  
Check the effect of `sort` function on the vector of names `x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )`.  
Next, print the effect of `sample` function on `x`.  
Repeat the last step to see another permutation of `x`.  
Check `Help` of `sample` and understand the effect of `sample( x, size = 2 )`.  
Finally, understand effect of the `rev` function applied to the vecotr `x`.

```{r}
x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
sort( x )              # x sorted alphabetically
sample( x )            # a random permutation of x
sample( x )            # a random permutation of x
sample( x, size = 2 )  # randomly chosen 2 different elements of x
rev( x )               # x in reversed order
rev( sort( x ) )       # x sorted in reversed alphabetical order
sort( x, decreasing = TRUE )  # a faster version of above
```

4. *Type conversion to a character vector*.  
Sometimes it is necessary to convert a numerical vector to a character vector.  
Understand what the function `as.character` does for argument `1:5`.

```{r}
1:5
as.character( 1:5 )
```

5. *Type conversion to a numerical vector*.  
Sometimes it is necessary to convert a character vector to a numerical vector.  
Understand what the function `as.numeric` does for argument `c( "1", "-1", "x" )`.  
Note the warning message. Why is there `NA`?

```{r}
as.numeric( c( "1", "-1", "x" ) )
```

