---
title: "Variables"
date: "`r format(Sys.time(), '%d %B %Y')`"
params:
  basename: "basic_variables0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
```

```{r, child="_navigate.Rmd"} 
``` 

# Tasks

- In the pane 'Console' type yourself commands described here. Do not use copy/paste!
- Familiarize with error messages caused by typing errors.

# Variable and its value

A variable stores a value, which can be explicitly provided or a result of a calculation.

Choose the names of variables freely, optimally meaningful to represent the problem of interest.

```{r}
gardenWidth <- 5
gardenLength <- 6
gardenArea <- gardenWidth * gardenLength
```

To find out what the value of a variable is, type its name:

```{r}
gardenArea
```

A frequent error is to ask for a non-existing variable:

```{r error=TRUE}
gggardenArea
```

# Assignment operator(s)

The arrow symbol `<-` is called the "assignment" operator:

```{r}
x <- 5
```



The same assignment can also be performed with symbol `=` (equal):

```{r}
x = 5
```

`r suggest("maybe leave out the assignment '=' symbol completely as a way of discouragement.")`

# No assignment, no change

Assignments are needed to store a result. 
When there is no assignment, the result is printed to screen and lost. For example, the sequence of statements below is the wrong way to "increase _x_ by one", because the result is calculated but not stored:

```{r}
x <- 5
x + 1
x
```

Here is the correct way:

```{r}
x <- 5
x <- x + 1
x
```

`r info_block("Remember: no assignment, no change.")`

# How variables are stored

The variables are stored in R memory.  
In one of the RStudio panes you can find a tab called _environment_.  
It contains a list of variables currently in memory.

When you close R, all variables are lost.  
Therefore, we will learn how **to write scripts** to store commands used to calculate variables.  
The scripts provide **reproducibility**. They store not only the results of calculations, but also document how the result was calculated.

# Valid variable names

Variable names in R are case sensitive (all other symbols also).

Note that `_` and `.` are allowed in variable names. 
Numbers (not at the beginning) are allowed too. 

Here are some examples (and each one is a different variable):

```{r eval=FALSE}
gardenLength <- 6
garden.length <- 6
garden_length <- 6
gArDeNlEnGtH <- 6     # this is bad, don't use this

len1 <- 5
len2 <- 6
```

`r suggest(" garden.length is a bad example, variables with dots are discouraged because they may clash with S3 method.")`


Choose meaningful variable names for readable scripts.

Naturally, there are some limitations. Do not use variable names :

- containing a space
- containing any of `@#!%$^()-+=!~?,<>}{][`
- with R reserved names: `for`, `while`, `repeat`, `if`, `else`, `TRUE`, `FALSE`, `Inf`, `NA`, `NaN`

# Autocomplete names

RStudio provides autocomplete feature, useful for long variable names.

Start typing the first letters of a (variable) name and press `Tab`.

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
