---
title: "Factors (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```

## Primary exercises

1. In a study participants were asked whether their sport activity is `none`, `oncePerWeek`, `severalPerWeek` or `daily`. Build a proper factor for the responses below. Store the factor in variable `w`. Print the factor. Write the code to count the numbers of occurrences of each level and print the counts.
```
severalPerWeek, none, none, oncePerWeek, oncePerWeek, 
oncePerWeek, oncePerWeek, ?, none, none
```

```{r}
v <- c( "severalPerWeek", "none", "none", "oncePerWeek", "oncePerWeek", "oncePerWeek", "oncePerWeek", NA, "none", "none" )
w <- factor( v, levels = c( "none", "oncePerWeek", "severalPerWeek", "daily" ) )
w
fct_count( w )
```

2. A factor with a random content.  
Read help about function `sample`. Then study the following lines of code and understand the results. Next, understand why an error is generated and use `replace` argument to generate a vector with 100 samples. Store this vector in a variable `v` and build a factor `w` from it. Finally, count the numbers of occurrences of each level in `w`. Ensure, that levels are in the provided order.

```{r error=TRUE,echo=TRUE}
lvs <- c( "none", "oncePerWeek", "severalPerWeek", "daily" )
sample( lvs, 3 )
sample( lvs, 3 )
sample( lvs, 3 )
sample( lvs, 100 )
```

```{r}
v <- sample( lvs, 200, replace = TRUE )
w <- factor( v, levels = lvs )
w
fct_count( w )
```

