---
title: "ggplot2 scales (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

pulse <- read_pulse() 
survey <- read_survey()
```


```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

1. *Simple regression*.  
Load the `survey.csv` data to variable `survey`.  
Perform a linear regression of `span1` against `span2` and store the result in `fit` variable.  
`fit` behaves also like a `list` so show `names` of elements stored in `fit`.  
Find and print the element corresponding to the coefficients of the regression -- do you see there a coefficient for `span2`?  
An intended method to access the coefficients is through `coef` function applied to `fit`. Do you get the same result as previously?  

```{r}
survey <- read_csv( "survey.csv" )
fit <-lm(span1 ~ span2, data=survey)
fit
names( fit )
fit$coefficients
coef( fit )
```

## Extra exercises

1. *Simple regression diagnostic plots (demo)*.  
Apply `plot` function to the `fit` object calculated in the above exercise.  
These are diagnostic plots for the model produced by `lm`.

```{r}
plot( fit )
```

