---
title: "Vectors (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```

## Primary exercises

1. Four patients had daily dietary intakes of 2314, 2178, 1922, 2004 kcal.  
Make a vector of these four values.  
What is the type of this vector?  
Convert these values into in kJ using 1 kcal = 4.184 kJ.  

```{r}
intakesKCal <- c( 2314, 2178, 1922, 2004 )
intakesKCal
class( intakesKCal )
intakesKCal * 4.184
```

2. Make a vector of all numbers between 1 and 100.  
Check the type of that vector (note, that a vector of numbers might be different than `numeric`).  
Calculate 1+2+3+…+100.  
Hint: Read help (see `Help` pane) for function `sum`. 

```{r}
v <- 1:100
class( v )
sum( v )
```

3. Degrees Celsius (`Cel`) can be calculated from degrees Fahrenheit (`Fah`) as `Cel = 5/9 * (Fah - 32)`.  
How much is 60 degrees Fahrenheit in Celsius?  
Use `seq` function to create a vector that contains 0,5,10,...,95,100 degrees Fahrenheit.  
Convert this vector to Celsius degrees.  
From your calculation in the previous exercise, select 10 Fahrenheit in Celcius.

```{r}
Fah <- 60
Cel <- 5 / 9 * (Fah - 32)
Cel
Fah <- seq( 0, 100, 5 )
Cel <- 5 / 9 * (Fah - 32)
Cel
Fah[3]
Cel[3]
```

## Extra exercises

1. Read help (see `Help` pane) about `seq` function.  
Use it to generate sequence: 10, 7, 4, 1, -2, -5.  
Understand the error message of `seq( 10, -5, 3 )`.

```{r}
seq( 10, -5, -3 )
seq( from = 10, to = -5, by = -3 )
```

2. Read help (see `Help` pane) about `rep` function.  
Use it to generate sequence: 0, 1, 0, 1, 0, 1.  

```{r}
rep( c( 0, 1 ), 3 )
```

3. Check the effect of `sort` function on the vector of names `x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )`.

```{r}
x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
sort( x )
```

4. Sometimes it is necessary to convert a numerical vector to a character vector.  
Understand what the function `as.character` does for argument `1:5`.

```{r}
1:5
as.character( 1:5 )
```

5. Sometimes it is necessary to convert a character vector to a numerical vector.  
Understand what the function `as.numeric` does for argument `c( "1", "-1", "x" )`.  
What is `NA`?

```{r}
as.numeric( c( "1", "-1", "x" ) )
```

