---
title: "Variables"
date: "`r format(Sys.time(), '%d %B %Y')`"
params:
  basename: "basic_variables0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
```

```{r, child="_navigate.Rmd"} 
``` 

## Variable and its value

A **variable** stores a **value**, which can be explicitly provided or a result of a calculation.

Choose the names of variables freely, optimally meaningful to represent the problem of interest.

Type the following lines in the `Console`. Observe changes in the `Environment` pane after each Enter.

```{r}
gardenWidth <- 5
gardenLength <- 6
gardenArea <- gardenWidth * gardenLength
```

To find out what the value of a variable is, type its name:

```{r}
gardenArea
```

A frequent error is to ask for a non-existing variable. Type to see the error:

```{r error=TRUE}
gggardenArea
```

## Assignment operator

The arrow symbol `<-` is called the **assignment operator**:

```{r}
x <- 5
```

## No assignment, no change

Assignment is needed to store a result.  
When there is no assignment, the result is printed to screen and lost.  

For example, the sequence of statements below is the wrong way to "increase _x_ by one".  
The result is calculated but it is not stored.  
Type these commands line by line. Observe when value of `x` changes in the `Environment` pane:

```{r}
x <- 5
x
x + 1
x
```

Here is the correct way to "increase _x_ by one".  
Type these commands line by line. Observe when value of `x` changes in the `Environment` pane:

```{r}
x <- 5
x
x <- x + 1
x
```

`r info_block("Remember: no assignment, no change.")`

## How variables are stored

The variables are stored in R memory and RStudio shows them in the `Environment` pane.  
When you close RStudio, all variables are lost.

We will learn how to write **scripts** to store commands used to calculate variables.  
The scripts provide **reproducibility**.  
They store not only the results of calculations, but also document how the results were calculated.

## Valid variable names

Variable names in R are case sensitive (all other symbols also).  
Note that `_` is allowed in variable names.  
Numbers (not at the beginning) are allowed too. 

Here are some examples (and each one is a different variable):

```{r eval=FALSE}
gardenLength <- 6
garden_length <- 6
gArDeNlEnGtH <- 6     # this is bad, don't use this
garden.length <- 6    # this leads to mistakes, don't use this

len1 <- 5
len2 <- 6
```

Choose meaningful variable names for readable scripts.

Naturally, there are some limitations. Do not use variable names:

- containing a space
- containing any of `@#!%$^()-+=!~?,<>}{][`
- with R reserved names: `for`, `while`, `repeat`, `if`, `else`, `TRUE`, `FALSE`, `Inf`, `NA`, `NaN`

## Autocomplete names

RStudio provides autocomplete feature, useful for long variable names.

Start typing the first letters of a (variable) name and press `Tab`.

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
