---
title: "Factors (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```

## Primary exercises

1. *Manually created factor.*  
In a study participants were asked whether their sport activity is `none`, `oncePerWeek`, `severalPerWeek` or `daily`.  
Build a proper factor for the responses below and store it in a variable `w`.  
Print the factor.  
Write the code to count the numbers of occurrences of each level and print the counts.
```
severalPerWeek, none, none, oncePerWeek, oncePerWeek, oncePerWeek, oncePerWeek, ?, none, none
```

```{r}
v <- c( "severalPerWeek", "none", "none", "oncePerWeek", "oncePerWeek", "oncePerWeek", "oncePerWeek", NA, "none", "none" )
w <- factor( v, levels = c( "none", "oncePerWeek", "severalPerWeek", "daily" ) )
w
fct_count( w )
```

2. *A factor with a random content.*  
Read help about the function `sample`.  
Then study and try the following lines of code to understand the results.  
Next, understand why an error is generated and use `replace` argument to generate a vector with 100 samples.  
Store this vector in a variable `v` and build a factor `w` from it.  
Finally, count the numbers of occurrences of each level in `w`.  
Ensure, that levels are in order provided in the variable `lvl`.

```{r error=TRUE,echo=TRUE}
lvs <- c( "none", "oncePerWeek", "severalPerWeek", "daily" )
sample( lvs, 3 )
sample( lvs, 3 )
sample( lvs, 3 )
sample( lvs, 100 )
```

```{r}
v <- sample( lvs, 200, replace = TRUE )
w <- factor( v, levels = lvs )
w
fct_count( w )
```

3. *Reordering factor levels.*  
When a factor is shown on an axis of a plot, the order is given by its levels.  
The factor `w` from the previous exercise will be then shown in this order: `none`, `oncePerWeek`, `severalPerWeek`, `daily`.  
But for a picture in a manuscript the following order might be needed: `daily`, `severalPerWeek`, `oncePerWeek`, `none`.  
Apply to `w` one of the `fct_` functions from the `tidyverse` library to produce a factor `w2` with the requested order.  
Show the levels of `w2`.  
Again show the number of elements of each level in `w2` and compare it with the table of the previous exercise.

```{r}
w2 <- fct_relevel( w, c( "daily", "severalPerWeek", "oncePerWeek", "none" ) )
levels( w2 )
fct_count( w2 )
```

## Extra exercises

1. *Special ordering of levels.*  
Go to the factors cheat sheet and find how to order the factor by the frequency of occurrences.  
Reuse `w` from the previous exercise and construct a factor `w3` with the same values but levels sorted with descending number of occurrences. Count occurrences to demonstrate correctness.  
Now, find a way to sort the levels in the increasing order.

```{r}
w3 <- fct_infreq( w )
fct_count( w3 )
fct_count( fct_rev( w3 ) )
```

