---
title: "Lists"
# date: "`r format(Sys.time(), '%d %B %Y')`"
params:
  basename: "basic_lists0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
```

```{r, child="_navigate.Rmd"} 
``` 

A **list** can be thought of as zero or more data elements (of _any type_), kept in order.  
The elements can be accessed by their **position**  (or **index**): first element, second element, ...  
Typically the elements are also named and consequently can be also accessed by the **names**.

A list allows to keep together related information.  
For example, functions often return lists when they calculate more than one value at once.

## Create a list

Type the following code to **manualy create** a list and store it in a variable `person`: 
```{r}
person <- list(
  name = "Bob",
  age = 44,
  children = c( "Amy", "Dan", "Eve" )
)
```

Now, check how the `person` list is printed:
```{r}
person
```

Test the class of `person`:
```{r}
class( person )
```

Try the function `str` (**structure**) on `person`.  
It gives a compact display of data stored in a variable and is useful for a `list`:
```{r}
str( person )
```

## Length and elements by position

Use `length` to find the **number of elements** in a list:
```{r}
length( person )
```

Elements of a list can be accessed **by their position** (first, second, ...) with help of `[[...]]` (double bracket) operator.  
Try the following to get the first and the last elements:
```{r}
person[[1]]
person[[ length(person) ]]
```

You may indeed observe that the elements of the `person` list are of different classes:
```{r}
class( person[[1]] )
class( person[[2]] )
```

## Names and elements name

To **get names** of all elements of a list enter:
```{r}
names( person )
```

Each of these names can be used with the `[[...]]` operator to **get/set an element by name**:
```{r}
person[[ 'name' ]]
person[[ 'children' ]]
```

Try to **add a new element** to the list using:
```{r}
person[[ 'smoker' ]] <- TRUE
person
```

and next **change the element**:
```{r}
person[[ 'smoker' ]] <- FALSE
person
```

and then **remove that element** from the list:
```{r}
person[[ 'smoker' ]] <- NULL
person
```

Note what happens when you try to access a **not present element**:
```{r}
person[[ 'isMarried' ]]
```

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
