---
title: "Vectors"
# date: "`r format(Sys.time(), '%d %B %Y')`"
params:
  basename: "basic_vectors0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
```

```{r, child="_navigate.Rmd"} 
``` 

## Concept of a vector

A **vector** is a container of (multiple) elements:

- All elements are of the same type (e.g. all are numbers).
- The elements are kept at numbered positions (there is the first element, second, ...).
- The elements might be given names (first may be named "Jay", second "Gloria", ...).

Some examples of a vector (and its **type**):

- Heights of students (a vector of numbers, `numerical`).
- Names of students (a vector of texts, `character`).
- Eyes color of students (a vector of values from a limited choice list, `factor`).
- Whether students have siblings (a vector of FALSE/TRUE values, `logical`).

Vector is the primary data structure of the R language.

## Numerical vector

You can create a vector using the **combine** function `c`.  
Type the following to create a vector of numbers and store it in variable `x`:

```{r}
x <- c(3, 6, 7, 2)
x
```

Calculations or functions often work on vectors elementwise.  
This is helpful to do multiple calculations simultaneously:

```{r}
x - 18
x^2
sqrt(x)
```

Some functions summarize a vector to a single number:

```{r}
sum(x)
mean(x)
```

## Sequence of numbers

A simple regular sequence you can create with `:` (**colon operator**). Try:

```{r}
y <- 1:10
y
7:9
```

You can also use the function `seq`:

```{r}
x <- seq(5,15,3)
x
```

## Character vector

A character vector you may also create with the combine function `c`:

```{r}
x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
x
```

Both single quotes (`'word'`) and double quotes (`"word"`) are allowed (must be identical at the beginning and the end of a text).

## Length and elements of a vector

Use `length` to find out the **length** (number of elements) in a vector (numerical, character, or any other type):

```{r}
x <- c( 11, 12, 13, 14 )
length(x)
y <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
length(y)
```

Use square brackets to **select** an element from a vector.  
The first element is at position 1.
Try the following to select an element from the second position:

```{r}
x[2]
y[2]
```

Now try to combine the brackets with `length` to select the last element. Type:

```{r}
x[ length(x) ]
```

Try the following to **select multiple elements** at once:

```{r}
x[ c(1,3) ]
w <- 2:4
x[w]
```

Note, when a vector is printed, the number reported on the left in square brackets shows the position of the next vector element:

```{r eval=FALSE}
x <- 101:155
x
```
```
 [1] 101 102 103 104 105 106 107 108 109 110 111 112
[13] 113 114 115 116 117 118 119 120 121 122 123 124
[25] 125 126 127 128 129 130 131 132 133 134 135 136
[37] 137 138 139 140 141 142 143 144 145 146 147 148
[49] 149 150 151 152 153 154 155
```

## Type (class) of a vector

Sometimes you may need to find what **type** of a vector is stored in a variable.  
This can be done with the function `class`.  

Type the following to create two vectors `x` and `y`:

```{r}
x <- c( 11, 12, 13 )
y <- c( "Jay", "Gloria", "Claire" )
```

To find the types of these vectors type:

```{r}
class( x )
class(y)
```


```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
