---
title: "Basic vectors"
date: "`r format(Sys.time(), '%d %B %Y')`"
params:
  basename: "basic_vectors0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
```

```{r, child="_navigate.Rmd"} 
``` 

# Tasks

- In the pane 'Console' type yourself commands described here. Do not use copy/paste!
- Familiarize with error messages caused by typing errors.

# Concept of a vector

A vector is a container of (multiple) elements.  
All elements are of the same type (e.g. all are numbers).  
The elements are kept in order (so, there is the first element, second, ...).  
This is the primary data structure of the R language.  

A single vector can represent each one of:

- heights of students in a class (a vector of numbers, `numerical`)
- names of students in a class (a vector of texts, `character`)
- eyes color of students in a class (a vector of values from a limited choice list, `factor`)
- presence of siblings of students in a class (a vector of FALSE/TRUE values, `logical`)

# Try a numerical vector

You can make a vector using the combine function `c`.  
Here it will be a vector of numbers:

```{r}
x <- c(3, 6, 7, 2)
```

Calculations or functions often work on vectors elementwise.  
This is helpful to do multiple calculations simultaneously:

```{r}
x - 18
x^2
sqrt(x)
```

Some functions summarize a vector to a single number:

```{r}
sum(x)
mean(x)
```

# Sequence of numbers

A simple regular sequence you can make with `:` (colon) operator:

```{r}
y <- 1:10
7:9
```

You can also use the function `seq`:

```{r}
x <- seq(5,15,3)
x
```

# Try a character vector

A character vector you may also create it with the combine function `c`:

```{r}
x <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
x
```

Both single quotes (`'word'`) and double quotes (`"word"`) are allowed (must be identical at the beginning and the end of a text).

# Elements of a vector

Use `length` to find out the number of elements in a vector (numerical, character, or any other type):

```{r}
x <- c( 10, 11, 12, 13 )
length(x)
y <- c( "Jay", "Gloria", "Claire", "Phil", 'Mitchell', 'Cameron' )
length(y)
```

Use square brackets to select a single element from a vector.  
The first element is at position 1.
To select an element from the second position use:

```{r}
x[2]
y[4]
```

Consequently, to select the last element:

```{r}
x[ length(x) ]
```

You can also select multiple elements at once:

```{r}
x[ c(1,3) ]
x[2:4]
```

Note, when a vector is printed, the number reported on the left in square brackets shows the position of the next vector element:

```{r}
x <- seq( 100, 160, 1 )
x
```

# TODOs

## For tasks

- `head` and `tail`
- `rev`
- nested combine: `c( c( ... ), 1:10 )`
- select through a variable
- set/get names() of vector elements
- `letters` and `LETTERS`; `pi`
- `as.numeric` and `as.character`
- `sort`, `order`, `sample`
- `which`, `which.max`
- `max`, `min`, `median`, `sd`

## To link out

- how to load a vector from file

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
