---
title: "Factors"
# date: "`r format(Sys.time(), '%d %B %Y')`"
params:
  basename: "basic_factors0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
```

```{r, child="_navigate.Rmd"} 
``` 

A **factor** is a vector of values from a limited set of possibilities (levels).

For example, a factor can be used to store colors of hair of some people.  
Let's assume that we can classify each person to one of the following hair colors.  
This will be the **levels** of the factor:

```{r eval=FALSE}
c( "black", "brown", "ginger", "blond", "gray", "white" )
```

Now, let's assume that for 10 people we observed the following hair colors.  
`NA` (not available, missing value) stands for a person whose hair color could not be determined.  
This will be the **values** of the factor:

```{r eval=FALSE}
c( "brown", "ginger", "black", "brown", NA, "brown", "black", "brown", "blond", "gray" ) 
```

## Create a factor

Enter the following to create the **factor** and store it in variable `hairColors`:

```{r}
hairColors <- factor(
  c( "brown", "ginger", "black", "brown", NA, "brown", "black", "brown", "blond", "gray" ),
  levels = c( "black", "brown", "ginger", "blond", "gray", "white" )
)
hairColors
```

## Levels and coding

In R the factor is stored as two vectors.  
The first vector codes the levels with subsequent integers: `1=black`, `2=brown`, ...  
Type the following to get this vector:

```{r}
levels( hairColors )
```

and to find the **number of factor levels** enter:
```{r}
nlevels( hairColors )
```

The second vector keeps integers corresponding to the values.  
Try the following and understand the relation of the integers and the levels:

```{r}
as.integer( hairColors )
```

## Count occurrences

## Rename levels

## Change levels order


`r todo( "Link to cheat sheet" )`




##
```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
