---
title: "Group (practice)"
# author: "R. Monajemi"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
```


```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

1. Report mean hand spans `{span1,span2}` per gender.

```{r}
survey %>% group_by(gender) %>% 
           summarise(meanWritingHand=mean(span1),meanNonWritingHand=mean(span2), .groups='drop')
```

2. Report mean and median `age` per gender, including the size of each gender group.

```{r}
survey %>% group_by(gender) %>% 
           summarise(size=n(), meanAge=mean(age), medianAge=median(age), .groups='drop')

```

3. Report mean `pulse` per `gender` and `exercise` group, including size of each group. Do the same for `gender` and `smokes` group.

```{r}
# gender,exercise group
survey %>% group_by(gender, exercise) %>% 
           summarise(size=n(), meanPulse=mean(pulse, na.rm=TRUE), .groups='drop')

# gender,smokes group
survey %>% group_by(gender, exercise) %>% 
           summarise(size=n(), meanPulse=mean(pulse, na.rm=TRUE), .groups='drop')

```

4. Report the mean `age` of those who exercise frequently per `gender` including the group size 

```{r}
# Solution 1:
survey %>% filter(exercise=='freq') %>% 
           group_by(gender) %>% 
           summarise(size=n(), meanAgeFreqExercise=mean(age), .groups='drop')

# Solution 2: first group and summarise and then filter out exercise group 'freq'. This is a 
# less optimal solution since it calculates the summary on all exercise groups, whereas the 
# solution 1 calculates the summary on a smaller table with only 'freq' exercise group.
# 
# survey %>% group_by(gender,exercise) %>%
#            summarise(size=n(), meanAgeFreqExercise=mean(age), .groups='drop') %>%
#            filter(exercise=='freq') %>%
#            select(-exercise)

```


## Extra exercises
