---
title: "Select (practice)"
# author: "R. Monajemi"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
```

```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

Apply the following to survey data: 

1) Select `name` along with all categorical variables into a new tibble `survey_cats`.

```{r}
survey_cats <- select(survey, name, gender, hand, fold, clap, exercise, smokes, m.i)
survey_cats
```

x) From survey data create *personal information* tibble with variables `Name`, `Age` and `Gender`. 

```{r}
survey_personal_info <- select(survey, Name=name, Age=age, Gender=gender)
```


2) Reorder the variables as such that name,age and gender come first. 

```{r}
select(survey, name,age,gender,everything())
```


4) Deselect variables that relate to hand and/or arm (e.g. span1, span2, hand, etc.)

```{r}
select(survey, -span1,-span2,-hand,-fold,-clap)
```


## Extra exercises


1) Rename the `m.i` variable to `system`.
  
```{r}
# Option 1: very tedious, you need to type all the variable names and 
# only rename the 'm.i' variable to 'system'
#
select(survey , name, gender, span1, span2, hand, fold, pulse, clap, 
       exercise, smokes, height, system=m.i, age)
# Option 2: shorter, but side-effect is that m.i (now system) comes at 
# the front.
select(survey, system=m.i, everything()) 
# Option 3: use rename function (see ?rename). 
rename(survey,system=m.i)
```

