---
title: 'tibble'
# author: "R. Monajemi"
params:
  basename: "dplyr_tibble0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)  
```

```{r, child="_navigate.Rmd"} 
``` 
   
A tibble is a two dimensional data structure with rows and columns, also referred to as `observations` 
and `variables`. I'll use these terms interchangeably depending on the context. The variables may take 
any [atomic types]() such as character,integer etc. Let's start by creating one manually.

## Create a tibble    
 
Creating a __tibble__ is done using the keyword _tibble_ taking a list of `key=value` pairs with `key` being the variable name and `value` the values of the observations. Take for example the variables `name`, `year` and `favourite` to represent a person's name, birth year and favourite colour:   

```{r, echo=TRUE}
# tibble
favourite_color  <- tibble(name=c("Lucas","Lotte","Noa","Wim"), 
             year=c(1995,1995,1995,1994), 
             favourite=c("Blue","Green","Yellow","Purple"))

favourite_color  
```
 
> What additional pieces of information do you see beside the content we provided?

```{r, echo=FALSE, results='asis'}
msg <- "1) '# A tibble: 3 x 4', which says that this is a tibble with 4x3 dimensions (4 observations 
and 3 variables), \n\n 2) the types of each variable and \n\n 3) the row numbers."
qa_toggle(msg)
``` 

Now inspect the type of the tibble object we just created:

```{r}
class(favourite_color)
```

As you can see a tibble is in it's core a base R `data.frame` with some additional behaviour 
enforced by {`tbl_df`,`tbl`}. For more details on the differences between tobble and data.frame 
see [tibble vs data.frame*](dplyr_tibble1.html).

 
## tibble $\leftrightarrow$ data.frame

Often you may need to switch between `data.frame` and `tibble`. You can convert an existing `tibble` into 
`data.frame` type using the function `data.frame()`: 

```{r}
favourite_color_df <- data.frame(favourite_color)
favourite_color_df
class(favourite_color_df)
```
 
To do the reverse, from `data.frame` to `tibble`, use the function `as_tibble()`:

```{r, eval=FALSE}
t <- as_tibble(favourite_color_df)
t
class(t) 
```

   

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_related.Rmd"} 
``` 

```{r, child="_navigate.Rmd"} 
``` 