---
title: 'select'
# author: "R. Monajemi"
params:
  basename: "dplyr_select0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)
pulse <- read_pulse()
```

```{r, echo=FALSE, results='asis'}
alert_block(params)
``` 

Use the `select` function to select variables (columns) from tibbles. Let's take the `pulse` dataset:   

```{r}
pulse 
```

`select` takes as it first argument a tibble, followed by a comma separated list of variables of your choice and returns a tibble with those chosen variables:

```{r}
select(pulse, name, age)
```

> Does `pulse` still contain the variables 'name' and 'age'?

```{r, echo=FALSE, results='asis'}
msg <- "Yes, 'select' returns the selection as a tibble and does not modify the underlying tibble. Use assignment '<-' to a variable to keep the selection."
QA(msg)
``` 

In order to keep your selection as a separate object you'll need to assign the result into a new fresh object, e.g. `name_age`:

```{r}
name_age <- select(pulse, name, age)
name_age
```

##  Variable order

The order of the selected variables is reflected in the resulted tibble:  

```{r}
select(pulse, age, name )
```

## Deselect variables 

You may also deselect variables, with other words the complement of your selection. This is done by the `-` sign:

```{r}
select(pulse, -smokes, -alcohol)
```

## Select and rename 

With selection it is possible to change the variable names simultaneously:  

```{r}
select(pulse, FirstName = name, Age = age)
```

> What is the variable name in the pulse dataset, 'Age' or 'age'?

```{r, echo=FALSE, results='asis'}
msg <- "age"
QA(msg)
``` 

## Reshuffle variables 

With `select` we can reshuffle the variables. Sometimes tibbles may have large number of variables, you can bring more 'important' variables with `select` and a convenience function `evertything`:   

```{r}
select(pulse, name, age, everything()) 
```

`everything` function lists all other variable other than `name` and `age` to be placed after `name` and `age` by `select`.


```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_related.Rmd"}  
``` 

```{r, echo=FALSE, results='asis'}
alert_block(params)
``` 
