---
title: 'Join'
# author: "R. Monajemi"
params:
  basename: "dplyr_join0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)  # UNCOMMENT IF SLOT HAS TASKS.
pulse <- read_pulse() 
```

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
```

> Combine tables 

`*_join(x,y,by)`  is a family of functions to combine two tables on common `variable(s)`. For example, take a group of people (n=7) for which we have collected age and height data:

```{r}
# height 
df1 <- tibble(name = c("Isa","Jaylinn","Mila","Milas","Yara"),
           height = c(160, 172, 182, 157, 162)) 
df1
# age
df2 <- tibble(name = c("Fiene","Jaylinn","Mila","Noah","Yara"),
           age = c(20,24,17,23,17)) 
df2
```

The information here is incomplete, for some we have only `height` and for some only `age`. Now we want to combine these tables into a single table with variables `name`, `age` and `height`. We can do this with a join function which is able to combine observations (rows) by **matching on the common variable(s) between the two tables**. Matching is done **by** one or more variables, in this case the variable `name`.

`r info_block("Keep in mind that for a proper joining of data, the matched variable(s), also known as the key, must be unique in both tables.")`

## Join : single variable 

You may control the layout of the combined table by choosing one of the three join function laid out below.

### inner_join

Return all rows in `df1` where there are matching values of `name` in `df2` and all columns in both `df1` and `df2`: 

```{r}
inner_join(df1,df2, by = "name")  #  <=> inner_join(df1,df2, by = "name")

```

### left_join

Return all rows from `df1` and all columns in both `df1` and `df2`, `NA` for missing values in `df2`:

```{r}
left_join(df1,df2, by = "name")
```


### full_join 

All rows and all columns in `df1` and `df2`.

```{r}
full_join(df1,df2, by = "name")
```

## Non-unique keys

If the keys are not unique in either of the tibbles then the result will be ambiguous. Take for example `df1` defined above, we now add another observation with  name `yara`:

```{r}
df1 <- tibble(name = c("Isa","Jaylinn","Mila","Milas","Yara", "Yara"),
           height = c(160, 172, 182, 157, 162, 168))
df1
```

This will make the key, variable `name`, non-unique. Notice the result of a  full join:

```{r}
full_join(df1, df2, "name")
```

As you can see the join found two matches for *Yara* in `df1`, so both are taken and the age for *Yara* in `df2` is replicated.  

`r info_block("Use only unique key in joins to avoid ambiguous results.")`

## Variable name conflicts

Take `df1` as defined above and `df3`: 

```{r}
df3 <- tibble(name =  c("Fiene","Jaylinn","Mila","Noah","Yara"),
           height = c(165, 172, 179, 173, 170), 
           ) 
df3
```

both of these tables have the variable height, after joining by `name` :

```{r}
left_join(df1,df3,by="name")
```

we see that `left_join` somehow needs to distinguish between the `height` from `df1` and the `height` from `df3` with suffixes `.x` and `.y` respectively. 


There are several more join functions not covered here such as right_join (opposite of left_join), semi_join and anti_join (see [dplyr cheat sheet](pdf/data-transformation.pdf)).


<!--

#### right_join

Return all rows from `df2` and all columns in both `df1` and `df2`, `NA` for missing values in `df1`

```{r}
right_join(df1,df2,"name")
```

#### semi_join

Identical to `inner_join` keeping only columns in `df1`  # (exception: no duplicated rows in df1.)

```{r}
semi_join(df1,df2,"name")
```

#### anti_join

Complement of `semi_join`, return rows from `df1` with no match in `df2`. 

```{r}
anti_join(df1,df2,"name")
```

-->

## Join : multiple variables 

To illustrate let's split the pulse dataset into two separate tables with common variables `name` and `height`: 

```{r}
p1 <- pulse  %>% select(name,height,age,gender,year)
p1
p2 <- pulse  %>% select(-age,-gender,-year)
p2
```

`p1` has the common variables `{name,height}` (the key) and additional variables `{age,gender,year}`, and `p2` has the same key variables `{name,height}` with some other measurement variables such as `pulse1` `pulse2`  etc.


> Is the combination `{name,height}` a valid choice as the key to join two tables p1 and p2? 

```{r, echo=FALSE, results='asis'}
msg <- "Yes. This because the combination is unique in the table. You can check this with the command nrow(distinct(pulse,name,height)) == nrow(pulse)."
qa(msg)
```

Now we want to join them back together using `full_join`: 

```{r}
full_join(p1,p2) # common variables are used when no variable is specified. 
```

It is a **better practice** to explicitly specify the variables: 

```{r} 
pulse_joined <- full_join(p1,p2, by = c("name","height"))
pulse_joined
```

You may check: 

```{r}
all_equal(pulse,pulse_joined)
```

`r show_warning(" The function all_equal is an experimental function and may become obsolete! ")`

```{r, child="_exercises_links.Rmd"}
``` 

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
```
