---
title: "Mutate (practice)"
# author: "R. Monajemi"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=TRUE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=TRUE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=TRUE)
}

survey <- read_survey()
pulse <- read_pulse()

```


```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises


1.  In the survey dataset add a new column `feet` with heights reported in feet unit (1 foot = 30.48 cm).

```{r}
mutate(survey, feet=height/30.48)
```

2. In the survey dataset add a new column `diffWritingHandSpan` : the difference of span1 (writing hand) and span2 (non-writing hand). 

```{r}
mutate(survey, diffWritingHandSpan=span1-span2)
```

3. In the pulse dataset add new weight variables `pounds` and `stones` (1 kg  = 2.20462 pound = 0.157473 stone).

```{r}
mutate(pulse, pounds = weight * 2.20462, stones = weight * 0.157473 )

```


## Extra exercises

In the survey dataset: 

1. Add a new column `diffHandSpan` : the absolute difference between span1 (writing hand) and span2 (non-writing hand). Hint: use `abs` function (?abs).  

```{r}
mutate(survey, diffWritingHandSpan=abs(span1-span2))
```

2. Change `height` unit from cm to inch (1 cm = 0.393701 inch).

```{r}
mutate(survey, height=height*0.393701)
```

