---
title: "Filter (practice)"
# author: "R. Monajemi"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
```
 

```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

In the survey dataset: 


1) Select teenagers, assume age range between and including 10 and 19.
  
```{r}
filter(survey, 10<=age & age<20)
```

2) Select all females with pulse equal to 60

```{r}
filter(survey, pulse==60 & gender=="female")
```

3) Select all male teenagers with pulse above 60.

```{r}
filter(survey, pulse<60 & gender=="male" & (9<age & age<20) )
```

4) How many males do smoke and never exercise?

```{r}
nrow( filter(survey, (smokes %in% c("regul","occas")) & exercise=="none" & gender=="male" ) )
```

5) How many females never smoke and frequently exercise?

```{r}
nrow( filter(survey, (smokes %in% c("regul","occas")) & exercise=="none" & gender=="female" ) )
```

## Extra exercise

1) What is the percentage of males who never smoke and frequently exercise? Do the same for female.

```{r}
# men
none_smoker_sportsmen <- nrow( filter(survey, smokes=="never"  & 
                                              exercise=="freq" & 
                                              gender=="male") )
total_men <- nrow( filter(survey, gender=="male") ) 
none_smoker_sportsmen / total_men

# women
none_smoker_sportswomen <- nrow( filter(survey, smokes=="never"  & 
                                                exercise=="freq" & 
                                                gender=="female") )
total_women <- nrow( filter(survey, gender=="female") ) 
none_smoker_sportswomen / total_women
```

2. What is the age range in teenagers? You may use the range function (?range).

```{r}
teenagers <- filter(survey, 10<=age & age<20) 
range(teenagers[["age"]])
```

