---
title: 'Summarise'
# author: "R. Monajemi"
params:
  basename: "dplyr_summarise0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)  # UNCOMMENT IF SLOT HAS TASKS.
pulse <- read_pulse()
```

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
```

Descriptive statistics enable you to describe basic features of your data, e.g. counts, sum, mean, median, etc.


## Counting occurrences of values

To count distinct values in a variable:

```{r}
pulse %>% count(year)
pulse %>% count(gender)
pulse %>% count(age)
```

For multiple variables, it is the count of each possible combination:

```{r}
pulse %>% count(gender, exercise)
pulse %>% count(year, gender)
```

## Summarise variable(s)

`summarise` function produces a single row of variable(s) summary given `summary function`

```{r}
pulse %>% summarize( count = n(), meanHeight = mean( height ) )
pulse %>% summarize( count = n(), meanAge = mean( age ) )
```

## Distinct values in variable(s)

To identify distinct values in a variable use the function `distinct`:

```{r}
pulse %>% distinct(year)
pulse %>% distinct(exercise)
pulse %>% distinct(ran) 
```

For multiple columns it is possible to identify distinct rows:
```{r}
pulse %>% distinct(gender, exercise)
```




```{r, child="_exercises_links.Rmd"}
``` 

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
```

