---
title: 'filter'
# author: "R. Monajemi"
params:
  basename: "dplyr_filter0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params) 
pulse <- read_pulse()
```

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
``` 

In the previous section with could select variables (columns) with all its values using the `select` function. In this section we are intersted in collection of 
observations (rows). For this we have the function `filter` which takes as its first argument the tibble followed by a condition describing the values to be filtered on.   

## Single condition

```{r}
filter(pulse, age == 22 ) # filter out all observations with age equal to 22 
```

## Multiple conditions 

```{r}
filter(pulse, age <= 20 & exercise == "moderate") 
```

## Condition on multiple values

```{r}
filter(pulse, exercise == "high" |  exercise =="moderate") 
```

```{r, child="_exercises_links.Rmd"}
```

```{r, child="_related.Rmd"}  
``` 

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
``` 
