---
title: 'tibble'
# author: "R. Monajemi"
params:
  basename: "dplyr_tibble0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)  
```

 
```{r, echo=FALSE, results='asis'}
alert_block(params)
``` 

A **tibble** is a table -- a two dimensional data structure with rows (**observations**) and columns (**variables**). 
I'll use these terms interchangeably depending on the context. 
Each variable has a fixed [type](basic_vectors0.html) such as `character`, `integer` etc. 
Let's start by creating a tibble manually.

## Create a tibble

To create a tibble you need to make sure that the package [tidyverse](https://www.tidyverse.org/) is installed and loaded. See [installation](installation0.html) for more details.  

Enter the following to load *tidyverse* package:

```{r}
library(tidyverse) 
```

Creating a __tibble__ is done using the keyword _tibble_ taking a list of `key=value` pairs where:

- `key` is the variable (column) name,
- `value` provides the values of the observations. 

Take for example the variables `name`, `year` and `colour` to represent a person's name, birth year and favourite colour:   

```{r, echo=TRUE}
favourite_colour  <- tibble(name=c("Lucas","Lotte","Noa","Wim"), 
                           year=c(1995,1995,1995,1994), 
                           colour=c("Blue","Green","Yellow","Purple"))
```

The variable `favourite_colour` now holds the data. Enter its name in the R `Console` for inspection:

```{r}
favourite_colour
```

> What additional pieces of information do you see beside the content we provided?

```{r, echo=FALSE, results='asis'}
msg <- "1) '# A tibble: 4 x 3', which says that this is a tibble with dimensions 4x3 (4 observations 
and 3 variables), \n\n 2) the atomic type of each variable, in this case character and double, \n\n 3) the row numbers"
qa(msg)
``` 

## Inspect your data

Type the following to find out the dimensions of the tibble:

```{r}
ncol(favourite_colour)  # number of variables (columns)
nrow(favourite_colour)  # number of observations (rows)
dim(favourite_colour)   # dimensions : 4 rows and 3 columns 
```

or top and bottom rows of the tibble: 

```{r}
head(favourite_colour, 2)  # first 2 observations (rows)
tail(favourite_colour, 3)  # last 3 observations (rows)
```

## subsetting:  [

Often you may need to select certain variable, this can be done using square brackets `[` :

```{r}
favourite_colour["colour"]
```

to subset combination of variables:

```{r}
favourite_colour[c("name","year")]
```
 
`r info_block("Subset result of a tibble is always a tibble.")`

## Extract variables as vectors:  [[ or $

If you want to work with variables as individual vectors then you can do this either by double square brackets or $ sign: 

```{r}
favourite_colour[["year"]]
favourite_colour$year
```

## tibble to/from file

Tibbles can be written to data files and read back again. Many data formats exist but for brevity we will be using
comma-separated-value (csv) format in this course. The functions involved for this purpose are `write_csv` and `read_csv`.
For more details on reading/writing format see [readr_io0](readr_io0.html).

let us now save our first tibble into a file in csv format:   

```{r}
write_csv(x = favourite_colour, path = "data/favourite_colour.csv")
```

`r todo( "where to explain explicit vs implicit function argument names?" )


`favourite_colour` tibble is written to `favourite_colour.csv` text file in sub-directory `data`. You may inspect the file with any editor and it should look something like: 

```
name,year,colour
Lucas,1995,Blue
Lotte,1995,Green
Noa,1995,Yellow
Wim,1994,Purple
```

This way we can permanently store our results in files for later use. We can now read the csv file back into a R 
object, e.g. `favourite_colour_csv` :

```{r}
favourite_colour_csv <- read_csv(file = "data/favourite_colour.csv")
```

`read_csv` gives a summary of the variables and their inferred types.

```{r}
favourite_colour_csv
```

## Type of tibble

Now inspect the type of the tibble object we just created:

```{r}
class(favourite_colour)
```

A tibble is in its core a 'data.frame', a base R data structure. For those of you interested in the differences between tibble and data.frame please see [tibble vs data.frame*](dplyr_tibble1.html).

`r info_block("'The types tbl_df' and 'tbl' enforce additional convinient behaviours specific to tibble.")`

`r todo("is help (?) already explained? It is used in the exercises.")`

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_related.Rmd"} 
``` 

```{r, echo=FALSE, results='asis'}
alert_block(params)
``` 
