---
title: 'Summarise'
# author: "R. Monajemi"
params:
  basename: "dplyr_summarise0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)  # UNCOMMENT IF SLOT HAS TASKS.
pulse <- read_pulse()
```

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
```

> Descriptive statistics

## Summarise variable(s)

The main function for calculating summaries on variables is `summarise`. Examples of descriptive functions are mean, median sum, and so on. The function consume a vector and produce a single value. `summarise` takes a tibble along with specification of descriptives and produces a single row. 

For example, let's say we want to know the mean height and weight of all individuals in the pulse dataset:  

```{r}
pulse %>% summarise(meanHeight=mean(height), meanWeight=mean(weight))
```

The result is a single row with two variables `meanHeight` and `meanWeight` with the corresponding mean values of of all observations.

We can also summarise on variable's range, e.g. `age` : 

```{r}
pulse %>% summarise(minAge = min(age), maxAge=max(age)) # <=> range(pulse$age)
```

**n()**: convenient function to calculate total number of rows in summarise context:

```{r}
pulse %>% summarise( count = n(), meanHeight = mean( height ) )
```


## Frequency tables

With the `count` function we can count the frequency of values in a categorical variables:

```{r}
pulse %>% count(gender)
pulse %>% count(smokes)
pulse %>% count(exercise)
```

The result enumerates the distinct values of the variable in the first column and their frequency in the last column `n`.

Multiple variables are allowed, it is the count of each possible combination values, also known as `contigency table` or `cross table`: 

```{r}
pulse %>% count(gender, exercise)
pulse %>% count(year, gender)
```

## Distinct values in variables

To identify distinct values in a variable we use the function `distinct`:

```{r}
pulse %>% distinct(year)
pulse %>% distinct(exercise)
pulse %>% distinct(ran) 
```

For multiple columns it is possible to identify distinct combinations of values:

```{r}
pulse %>% distinct(gender, exercise)
```

`r info_block("'distinct' produces the same combination of variables values as 'count', except without the frequncy column 'n'.")`

## glimpse

## arrange

```{r, child="_exercises_links.Rmd"}
``` 

```{r, echo=FALSE, results='asis'}
navigate_slots(params)
```

