---
title: "tibble (tasks)"
# author: "R. Monajemi"
#params:
#  solutions: "yes"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```
  
## Create tibble  
  
1. Create a tibble for {Sonja, Steven, Ines, Robert, Tim} with their heights {164, 188, 164, 180, 170}, weights {56.0, 87.0, 54.0, 80.0, 58.5} and frequency of exercise {high, high, low, moderate, low}.

```{r} 
exercise_group <- tibble(name=c("Sonja" , "Steven", "Ines", "Robert", "Tim" ),
            height=c(164, 188, 164, 180, 170),  
            weight=c(56.0, 87.0, 54.0, 80.0, 58.5),
            exercise=c("high", "high", "low", "moderate", "low")
     )  
exercise_group
```

2. Create the same tibble in the previous exercise but now with variable exercise as `factor`.

```{r}
exercise_group <- tibble(name=c("Sonja" , "Steven", "Ines", "Robert", "Tim" ),
            height=c(164, 188, 164, 180, 170),  
            weight=c(56.0, 87.0, 54.0, 80.0, 58.5),
            exercise=factor(c("high", "high", "low", "moderate", "low"))
     )  
exercise_group
```

## tibble subset

1. Take the tibble from the previous exercise and create a new tibble without the `height` and `weight` variables.

```{r}
exercise_group_sub <- exercise_group[c("name","exercise")]
exercise_group_sub
```

## Extract variables

1. Given the tibble `favourite_color`, how old were the subjects by the end of 2019?

```{r, echo=FALSE}
favourite_color  <- tibble(name=c("Lucas","Lotte","Noa","Wim"), 
                           year=c(1995,1995,1995,1994), 
                           colour=c("Blue","Green","Yellow","Purple"))
```

```{r}
2019 - favourite_color[["year"]] # alternatively:  2019 - favourite_color$year
```

2. What is the mean height in `exercise_group`?

```{r}
mean(exercise_group[["height"]]) 
```


## tibble to/from file

1. Read csv files `pulse.csv` and `survey.csv` into R as objects `pulse` and `survey` respectively.

```{r}
pulse <- read_csv(file = "_data/pulse.csv")
survey <- read_csv(file = "_data/survey.csv")
```

## Extra exercise*

1. ...


   

 