---
title: 'Introduction'
#author: "R. Monajemi"
params:
  basename: "introduction0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
# build_nocode(params)
```

```{r, child="_navigate.Rmd"} 
``` 

## Why this course? 

Modern science requires modern statistical methods:
 
- Genomics/bioinformatics
- Advanced survival analysis
- Causal modelling
- ...

Statisticians develop new methods and often make them first available as packages in R.  
No need to wait until these methods are programmed into SPSS.

## What is _R_? 

### A short history

_S_ was a programming language for statistics developed by John Chambers (Bell Labs) in the 1970s-80s.  
It had two primary implementations:

- S-plus (1988, commercial);
- _R_ (1993, GNU public license) by Robert Gentleman and Ross Ihaka (Univ. of Auckland).

Now _R_ is known as "a free software environment for statistical computing and graphics."

### R project

Current _R_ developments are streamlined by non-profit organisation _The R Project for Statistical Computing_ at https://www.r-project.org/.

The project supports a large and continuously growing community of users and package developers:

- volunteer work (mostly by academics);
- anyone can see the source code;
- anyone can contribute (write code, report bugs, write documentation).

_CRAN_ (_Comprehensive R Archive Network_) at https://cloud.r-project.org/ is a central repository for R language interpreter and R packages.  
It also contains manuals and mailing lists (well indexed on google).  
Free download. New major version every year, minor versions in between.

### RStudio

_RStudio_ from https://rstudio.com/ is an open source integrated development environment for programming in _R_ language.

It provides useful features to help in development of the R code and in organization of projects.

RStudio is not necessary to work with R programs, but it is highly recommended and is used in this course.


## Course overview

### What we teach

This is not a statistics course!  
The focus is _R_ as a _language_ for data analysis.

The goal of the course is to teach students how _R_ language can be used to build a report with very simple statistical analysis of data provided in a table.  

This course teaches _R_ proficiency:

- The mechanics of _R_.
- How to use other people's _R_ scripts.
- How to write your own _R_ scripts.
- How to use _R_ packages.


### Course structure

The course materials are designed for self-study.  

The necessary R concepts are split in a number of small sections.  
Each section introduces several new R language commands.  
We ask students to type these commands and observe their effects (do not use copy-paste; own typing is important in order to learn how R communicates mistakes).  
Exercises at the end of the sections demonstrate other usage of the commands or introduce _extra_ material with related functionality.


## The exam

To pass the exam the students are requested to build a knittable R Markdown document reporting analysis of a table. The document has the following parts:

- read a table from a provided file;
- show some selected content from the table;
- summarise the table data;
- produce some plots of the data.

The students are allowed to use the course (or other) materials during the exam but it is strictly forbidden to communicate with other persons during the exam.

The material present in the _extra_ exercises is not necessary for the exam.

```{r, child="_navigate.Rmd"} 
``` 
  
