---
title: 'R'
#author: "R. Monajemi"
params:
  basename: "introduction0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
# build_nocode(params)
```

```{r, child="_navigate.Rmd"} 
``` 

## Why this course?

Modern science requires modern statistical methods:
 
- Genomics/bioinformatics
- Advanced survival analysis
- Causal modeling
- ...

Statisticians develop methods and make them available as packages in R.

No need to wait until they are programmed into SPSS.

## What is R? 

### _R_: a short history

_S_: a programming language for statistics by John Chambers (Bell Labs) in the 1970s-80s.

Two implementations:

- S-plus (1988, commercial)
- R (1993, GNU public license) by Robert Gentleman and Ross Ihaka (Univ. of Auckland)

The R Foundation for Statistical Computing. Non-profit. 

Huge community (users, package writers).

### Open source

Free software.

Volunteer work (mostly by academics).

Anyone can see the source code.

Anyone can contribute:

- write code
- report bugs
- write documentation


### Obtaining R

CRAN: Comprehensive R Archive Network 

- Repository for R and packages
- Go to http://www.r-project.org

Free download

- New major version (R 3.5.0) every year 
- New minor versions (R 3.5.2) in between

Also on CRAN

- Manuals (don't read them)
- Mailing lists + archives (well indexed on google)


### R and RStudio

RStudio: open source integrated development environment to R (2011)

Adds useful features to help write code and organize projects

Not necessary to use R, but highly recommended

RStudio organises input and output in useful windows


## Course overview

### What we teach

This is not a statistics course!
	
To learn about statistics, follow _Basic Methods and Reasoning in Biostatistics_ (or more advanced courses).

This course teaches R proficiency:

- The mechanics of R
- How to use other people's R scripts
- How to write your own R scripts
- How to use R packages 

Focus: R as a _language_ for data analysis.


### Course structure

- Interactive lectures 
- Practice time in between 

Hands-on at four levels:

1. *Type with me*
2. Mini-exercises briefly interrupt the lecture
3. Longer exercises to do it on your own
4. Advanced exercises introduce advanced concepts for quick learners (optional)

Eight half-day slots.

 
```{r, child="_navigate.Rmd"} 
``` 
  