---
title: "Reshape (practice)"
#author: "R. Monajemi"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

survey <- read_survey()
pulse <- read_survey()
```


```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

In the survey dataset:

1. The variables `span1` and `span2` are of the same type representing the span of the writing hand and the non-writing hand respectively. We would like to analyse these values in groups. To do this we need to convert  the variables `span1` and `span2` into variables `writing` and `span` with `writing` ranging over *span1* and *span2* and `span` over the values of `span1` and `span2` variables. Use the function `gather(...)`.     
  
```{r}
survey_span <- survey %>% gather(key=writing,value=span,span1, span2)
survey_span
```

## Extra exercises

1. Summarise minimum and maximum `span` per `writing` from the primary exercise (1). 

```{r}
survey_span  %>% group_by(writing) %>%  
          summarise(min=min(span,na.rm = T), max=max(span,na.rm = T))
```

<!-- 
1. Two other categorical variables in the pulse dataset which can be gathered are `smokes` and `alcohol`. Do this into new variables `drug_habit` and `drug_use`.
--> 