---
title: 'Reshape'
# author: "R. Moanjemi"
params:
  basename: "tidyr_reshape0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
build_nocode(params)  # UNCOMMENT IF SLOT HAS TASKS.
pulse <- read_pulse()
```

```{r, child="_navigate.Rmd"} 
``` 


## `gather` : wide to long format

This function allows collapsing 'similar' variables into one variable while guaranteeing the data set's consistency. For example take the variables `pulse1` and `pulse2` in the following subset of the pulse data set:   


```{r}
pulses <- pulse  %>%  
      select(name,pulse1,pulse2)  %>% 
      head(3) 
pulses
```

We can transform the table as such that all pulse values are under a single variable, say `pulse`:


```{r}
dfLong <- pulses %>% gather(key=pulse,value=level, pulse1,pulse2)
dfLong
```

This is called the long version of the original (wide) table and contains the same information. 

> Are there other variables in pulse data set which can be 'gathered'? If so make the conversion to the long format.




## `spread` : long to wide format

```{r}
dfWide <- dfLong %>% spread(key=pulse,value=level)
dfWide
```

```{r, child="_exercises_links.Rmd"}
``` 

```{r, child="_navigate.Rmd"} 
``` 
