---
title: 'tidyverse'
# author: "R. Monajemi"
params:
  basename: "tidyverse0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, paged.print=FALSE)
```

```{r, child="_navigate.Rmd"}  
``` 

## tidyverse package

[Tidverse](https://www.tidyverse.org/) is a colection of R packages which we will be using intensively for the remainder of this course. If you haven't installed `tidyverse` then you can do this with:

```{r, eval=FALSE}
install.packages("tidyverse") # needed only once
```

and load it into R: 

```{r, eval=FALSE}
library(tidyverse) # each time you start RStudio
```

Something similar to the following will appear on your screen after loading tidyverse:

```
> library(tidyverse)
── Attaching packages ─────────────────────────────────────── tidyverse 1.3.0 ──
✔ ggplot2 3.2.1     ✔ purrr   0.3.3
✔ tibble  3.0.3     ✔ dplyr   0.8.4
✔ tidyr   1.0.2     ✔ stringr 1.4.0
✔ readr   1.3.1     ✔ forcats 0.4.0
── Conflicts ────────────────────────────────────────── tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
> 
```

It shows the sub-packages and the versions and also some conflicts. 

`r show_warning("You may ignore these Conflicts.")`

You now have access to a powerful set of functions for **data manipulation and visualisation**. 

## outline

- Data transformation with dplyr ([cheat sheet](pdf/data-transformation.pdf){target="_blank"})
  - tibble : a two-dimensional data structure with rows and columns
  - functions to manipulate tibble:
    - select : extract columns
    - filter : extract rows
    - mutate : add a new column
  - pipe operator `%>%` 
  - summaries and sorting 
  - grouping
  - queries 
  - join : combining tables
- reshape : change the table layout ([cheat sheet](pdf/data-import.pdf){target="_blank"})  

## Links 
- [Tidyverse](https://www.tidyverse.org/)
- [R for data science ; Hadley Wickham, Garrett Grolemund (online book)](https://r4ds.had.co.nz/)

```{r, child="_related.Rmd"}  
``` 

```{r, child="_navigate.Rmd"}  
``` 
