---
title: "Variables (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```

## Primary exercises

1. *Create a new variable*.  
Make a variable `heightInMeters` that contains your height expressed in meters.  
Print the variable.  
Find the variable and its value in the Environment pane.

```{r}
heightInMeters <- 1.75
heightInMeters
```

2. *Use a variable in an expression*.  
Use the variable `heightInMeters` to calculate your height in cm.  
Calculate the height in cm once again and store the result in a variable `heightInCentimeters`.  
Show/print `heightInCentimeters`.

```{r}
heightInMeters * 100
heightInCentimeters <- heightInMeters * 100
heightInCentimeters
```

3. *Change a value of a variable*.  
Increase the height stored in `heightInMeters` by 0.1m.  
Check whether the value of the variable indeed changed.  
Has also the value of `heightInCentimeters` changed? Why?

```{r error=TRUE}
heightInMeters <- heightInMeters + 0.1
heightInMeters
heightInCentimeters      # not changed, for change assignment is needed
```

4. *Typo in variable name*.  
Observe the error shown when you try to print the variable but you make a typo.

```{r error=TRUE}
height.in.meters      # Error: object 'height.in.meters' not found
```
