---
title: "Calculator (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}
```

## Primary exercises

1. *Simple expression.*  
Calculate 25 plus the square of 12.

```{r}
25 + 12^2
```

2. *Calling a function.*  
What is the square root of the number you have just calculated (here type the number directly)?

```{r}
sqrt( 25 + 12^2 )
```

3. *Prompt with `+`.*  
In the console type an incomplete expression `(1+2` and press Enter. Practice how to return to the proper console prompt `>`. Finally, clear the console.

```{r}
# To stop in the console press: Esc or Ctrl-C
# To clear in the console press: Ctrl-L
```

4. *Some error messages.*  
In the console type an error expression `(1+2)*(3+4))`. Understand the error message. Also understand errors reported for expressions: `0,25`, `sin()`, `xyz`.

```{r}
# (1+2)*(3+4))     -> forgotten (
# 0,25             -> use dot as decimal point: 0.25
# sin()            -> sin needs an argument
# xyz              -> xyz is not known
```

5. *Getting help.*  
Find the manual page for the `sin` function.

```{r}
# In the console type: ?sin
```

