---
title: "ggplot2 facets and themes (practice)"
---

```{r , include=FALSE}
if (!is.null(knitr::opts_knit$get("solutions"))) {
  if (knitr::opts_knit$get("solutions")=="no") {
    knitr::opts_chunk$set(echo = FALSE, eval = FALSE, comment = NA, paged.print=FALSE)
  } else  {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
  }
} else {
    knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, paged.print=FALSE)
}

pulse <- read_pulse()
survey <- read_survey()
```


```{r, child="_exercises_format.Rmd",eval=TRUE,echo=FALSE} 
``` 

## Primary exercises

1. *Visualize effect of running on pulse*.  
Reconstruct the following image based on the `pulse` data.  
Add component `coord_fixed()` to enforce that unit sizes are identical for both axes.  
Use `geom_abline( slope = 1 )` to get the line showing where `pulse1` and `pulse2` are equal.  
Disable the warning message.

```{r warning=FALSE,eval=TRUE}
ggplot( pulse ) +
  aes( x = pulse1, y = pulse2 ) +
  geom_point() +
  facet_wrap( ~ ran ) +
  coord_fixed() +
  geom_abline( slope = 1, color = "red", linetype = 2 ) +
  theme_bw() +
  xlab( "pulse1 (before exercise) [1/min]" ) +
  ylab( "pulse2 (after exercise) [1/min]" )
```


## Extra exercises

1. *Rotate axis labels*.  
When using facets the numbers on the bottom axis often overlap.  
Modify the top chart above by adding (after `theme_bw`):  
`  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))`.  
`vjust` and `hjust` are numbers in range [0,1]. Understand their effect.  
Try `angle=45`.

```{r warning=FALSE,eval=TRUE}
ggplot( pulse ) +
  aes( x = pulse1, y = pulse2 ) +
  geom_point() +
  facet_wrap( ~ ran ) +
  coord_fixed() +
  geom_abline( slope = 1, color = "red", linetype = 2 ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab( "pulse1 (before exercise) [1/min]" ) +
  ylab( "pulse2 (after exercise) [1/min]" )
```
