---
title: 'Basics'
author: "<author>"
params:
  basename: "slot0"
---

This session consists of three parts:

  1. **RStudio:** RStudio is currently defacto *Integrated Development Enviroment* (IDE) for R.  It 
  is assumend that you have already installed [R](https://www.r-project.org/) and 
  [RStudio](https://rstudio.com/) in that order. If you haven't then please install [R](https://www.r-project.org/) first (the latest version is : R 3.6.3 "Holding the Windsock") and then [RStudio](https://rstudio.com/). **Note of caution**, installing R and RStudio on different (network) drives drives may cause unpredictable behaviour. So installing both on the same drive, prefereably local drive, is advised.   

> [Basics 1](https://rladiessydney.org/courses/ryouwithme/01-basicbasics-1/)
  
  2. **Installing and loading packages**

> [Basics 2](https://rladiessydney.org/courses/ryouwithme/01-basicbasics-2/)
  
  3. **Reading data into R**

> [Basics 3](https://rladiessydney.org/courses/ryouwithme/01-basicbasics-3/)


```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
library(tidyverse)
build_nocode(params) # build nocode version of tasks ; replace with source("setup.R") when running stand-alone (knit)
```


```{r, child="_exercises_links.Rmd"}
``` 